\title{A grammar of Kalamang}
% \subtitle{The Papuan language of the Karas Islands}
\author{Eline Visser}
\renewcommand{\lsSeries}{cogl}%use series acronym in lower case
\renewcommand{\lsSeriesNumber}{4}

\BackBody{This book is a grammar of Kalamang, a Papuan language of western New Guinea in the east of Indonesia. It is spoken by around 130 people in the villages Mas and Antalisa on the biggest of the Karas Islands, which lie just off the coast of Bomberai Peninsula. This work is the first comprehensive grammar of a Papuan language in the Bomberai area. It is based on eleven months of fieldwork. The primary source of data is a corpus of more than 15 hours of spoken Kalamang recorded and transcribed between 2015 and 2019.
	
This grammar covers a wide range of topics beyond a phonological and morphosyntactic description, including prosody, narrative styles, and information structure. More than 1000 examples illustrate the analyses, and are where possible taken from naturalistic spoken Kalamang. The descriptive approach in this grammar is informed by current linguistic theory, but is not driven by any specific school of thought. Comparison to other West Bomberai or eastern Indonesian languages is taken into account whenever it is deemed helpful.

Kalamang has several typologically interesting features, such as unpredictable stress, minimalistic give-constructions consisting of just two pronouns, aspectual markers that follow the subject, and the NP and predicate – rather than the noun and verb – as important domains of attachment.

This grammar is accompanied by an openly accessible archive of linguistic and cultural material and a dictionary with 2700 lemmas. It serves as a document of one of the world's many endangered languages.}


\BookDOI{10.5281/zenodo.6499927}
\ISBNdigital{978-3-96110-343-0}
\ISBNhardcover{978-3-98554-039-6}

\proofreader{Amir Ghorbanpour,
Amy Amoakuh,
Andi Hölzl,
Heriberto Avelino,
Bev Erasmus,
Bianca Prandi,
Cesar Perez Guarda,
Christopher Straughn,
Craevschi Alexandru,
Fahad Almalki,
Jeroen van de Weijer,
Katja Politt,
Liam McKnight,
Ludger Paschen,
Maša Beslin,
Matthew Windsor,
Neneng Sri,
Jean Nitzke,
Phil Duncan,
Siva Kalyan,
Sune Gregersen,
Thera Crane,
Tihomir Rangelov,
Yvonne Treis
}

\typesetter{Eline Visser, Sebastian Nordhoff}
\renewcommand{\lsID}{344}
