\chapter{Word classes}\is{word class|(}\is{part-of-speech|see{word class}}\is{grammatical category|see{word class}}
\label{ch:wc}
Word classes are categories of words in the lexicon that share morphosyntactic behaviour. They are also referred to as grammatical categories or parts-of-speech and are the building blocks of grammar. In this chapter, the Kalamang word classes are introduced and defined. For most word classes the major definitional criteria are based on the morphosyntactic behaviour of the words, but sometimes semantic patterns determine ambiguous cases or support the syntactic criteria. In the case of demonstratives (§\ref{sec:wcdem}) and question words (§\ref{sec:wcq}) functional properties are an important criterion for grouping words together, as they can \textit{replace} words from (an)other class(es) in the clause.

I start with a description of the major open word classes: verbs (§\ref{sec:wcverb}) and nouns (§\ref{sec:wcnoun}). Pronouns, a closed subclass of nouns, are described in §\ref{sec:wcprons}. The closed minor word classes are described subsequently: quantifiers in §\ref{sec:wcquantnum}, demonstratives in §\ref{sec:wcdem}, postpositions in §\ref{sec:wcpost}, adverbials in §\ref{sec:wcadv}, question words in §\ref{sec:wcq}, conjunctions in §\ref{sec:wconj} and interjections in §\ref{sec:wcyesno}. Each section starts with a summary of the criteria used to define the class, followed by examples.

\section{Verbs}\is{verb|(}
\label{sec:wcverb}
    The open class of verbs consists of words that typically function as predicates. A predicate is an expression that takes a subject to form a clause, and it expresses something about this subject. Furthermore, verbs:
    
    \begin{enumerate}
    	\item may occur in complex predicates\is{predicate!complex};
    	\item may be used as adnominal modifiers\is{attribute} with attributive marker \textit{=ten}.
    \end{enumerate}

Semantically, verbs are words that denote events, states or properties. In the case of homonymous nouns and verbs like \textit{rer} `story' and \textit{rer} `tell a story', the semantic criteria determine that the second \textit{rer} is a verb. Verbs can be active or dynamic (denoting events) or stative (denoting states or properties).  Kalamang has no separate adjective class; instead, properties (e.g. colours and attributes) are expressed with monovalent verbs.\is{adjective*} \is{attribute!predicative} I refer to these modifier words as stative verbs.\is{modifier word}\is{property verb|see{modifier word}}\is{stative verb|see{modifier word}}

Kalamang has two verb classes: regular verbs and irregular verbs. Regular verbs can take mood enclitics, negator \textit{=nin} and predicate linker \textit{=i} directly on the root. On irregular verbs, an /n/ or /t/ is inserted before some enclitics. Also, an optional /n/ may appear on uninflected forms. Lastly, they typically have a vowel-final imperative instead of using imperative \textit{=te} (§\ref{sec:imp}). The most common verbal inflections are illustrated in Table~\ref{tab:ntverbs2}. Regular verbs may have any final phoneme, while all irregular verbs are vowel-final. The irregular verb class has two subclasses: transitive/intransitive verb pairs in \textit{-ma} and \textit{-cie} (§\ref{sec:macie}) and directional verbs in \textit{-ra} or \textit{-a} (§\ref{sec:dir}).

\begin{table}
	\caption{Behaviour of regular and irregular verbs under inflection}
	
		\begin{tabularx}{.8\textwidth}{XXl}
			\lsptoprule
			&	regular & irregular\\  \midrule
			root & \textit{deir} & \textit{jie}\\
			gloss & bring & get; buy\\
			alt. root & \textendash  &  \textit{jie-\textcolor{blue}{n}}\\
			\textit{=i} \textsc{plnk} & \textit{deir=i} & \textit{jie-\textcolor{blue}{n}=i}\\
			\textit{=et} \glet & \textit{deir=et} & \textit{jie-\textcolor{red}{t}=et}\\
			\textit{=kin} \glkin  & \textit{deir=kin} & \textit{jie-\textcolor{red}{t}=kin}\\
			\textit{=nin} \textsc{neg} & \textit{deir=nin} & \textit{jie-\textcolor{red}{t}=nin}\\
			\textit{=in} \textsc{proh} & \textit{deir=in} & \textit{jie-\textcolor{red}{t}=in}\\
			imperative & \textit{deir=te} & \textit{di}\\ 	
			\lspbottomrule
		\end{tabularx}
	\label{tab:ntverbs2}
\end{table}

Each verb licenses zero to three arguments\is{argument}. Verbs follow the subject and the object. There is no agreement marking on the verb.\footnote{Except for verbs in the imperative mood, which distinguish between singular and plural subjects (§\ref{sec:verbdistr}).} (\ref{exe:kalis}) contains a zero-intransitive clause without argument. (\ref{exe:monovstat}) has a monovalent stative verb and (\ref{exe:monov}) a monovalent active verb, each with one argument. (\ref{exe:biv}) shows a bivalent verb with its two arguments. (\ref{exe:triv}) contains a trivalent verb with its three arguments.\is{monovalent clause|see{intransitive clause}}\is{intransitive clause}\is{active verb}\is{bivalent clause|see{transitive clause}}\is{transitive clause}\is{trivalent clause|see{three-participant clause}}\is{three-participant clause}


\begin{exe}
	\ex \gll mu toni kalis=kin\\
			\textsc{3pl} think rain={\glkin}\\
			\glt `They think it will rain.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC0-1}{[narr22\_4:04]}}
		
		\label{exe:kalis}
	\ex
	{\gll teun \textbf{ten}\\
	fruit bad\\
	\glt `The fruits are bad.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD0-8}{[narr13\_4:14]}}
	}
	\label{exe:monovstat}	
	\ex
	{\gll mu \textbf{kiem}\\
		\textsc{3pl} run\\
		\glt `They run.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_15:26]}}
	}
	\label{exe:monov}
	\ex
	{\gll emun tumun=at \textbf{narorar}\\
		mother child=\textsc{obj} take\_by\_hand\\
		\glt `The mother takes the child by the hand.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA8-B}{[stim4\_4:47]}}
	}
	\label{exe:biv}
	\ex 
	{\gll mu kawir-un=at mat \textbf{naunak}\\
		\textsc{3pl} hat-\textsc{3poss=obj} \textsc{3sg.obj} show\\
		\glt `They showed him the hat.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD1-D}{[stim31\_2:30]}}
	}
	\label{exe:triv}
\end{exe}

Verbs may be part of complex predicates (Chapter~\ref{ch:svc}). (\ref{exe:svc}) shows a directional construction, with \textit{tiri} `sail' and \textit{yecie} `return' to express `sail back'. (\ref{exe:svcmod}) combines a stative verb `be far' and a directional verb `descend' to form `go down far'.

\begin{exe}
	\ex
	{\gll in se \textbf{tiri} \textbf{yecie}\\
		\textsc{1pl.excl} {\glse} sail return\\
		\glt `We sailed back.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAE-4}{[narr44\_2:56]}}
	}
	\label{exe:svc}
	\ex
	{\gll ma \textbf{kahen=i} \textbf{baran}\\
		\textsc{3sg} far={\gli} descend\\
		\glt `He goes down far.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B92-E}{[stim44\_0:32]}}
	}
	\label{exe:svcmod}
\end{exe}

All verbs can be used attributively\is{attribute}, that is, modifying a noun in a noun phrase (NP). In attributive function, verbs can carry the attributive clitic \textit{=ten} (§\ref{sec:attr}). A verb in post-nominal position can be a predicate, in which case it is bare, or it can be a modifier, in which case it carries \textit{=ten}. Consider the contrast in~(\ref{exe:rorpotma}). In~(\ref{exe:rorpotma}a), the noun \textit{ror} `wood' is incorporated in the verb \textit{potma} `cut'. In~(\ref{exe:rorpotma}b), the verb \textit{paramua} `cut' functions as a modifier of the noun and carries the attributive enclitic \textit{=ten}.

\begin{exe}
	\ex 
	\begin{xlist}
		\ex 
		{\gll ror-potma\\
			wood-cut\\
			\glt `to cut wood' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB3-0}{[narr7\_4:13]}}}
		\ex
		{\gll ror paramuan=\textbf{ten}\\
			wood cut\textsc{=at}\\
			\glt `cut wood (wood that is cut)' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_vvsa\_4]}}}
	\end{xlist}
	\label{exe:rorpotma}
\end{exe} 

Consider also the contrast in~(\ref{exe:kiemden}), where the two verbs are either used in attributive function modifying the subject, or in predicative function.

\begin{exe}
	\ex 
	\begin{xlist}
		\ex 
		{\gll tumun \textbf{ririn}=\textbf{ten} kiem\\
			child tall=\textsc{at} run\\
			\glt `The tall child runs.'}
		\ex
		{\gll tumun \textbf{kiem}=\textbf{ten} ririn\\
			child run=\textsc{at} tall\\
			\glt `The running child is tall.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_adj\_45]}}}
	\end{xlist}
	\label{exe:kiemden}
\end{exe} 

%Verbs that are predicates may also carry \textit{=ten}. An analysis of this use is presented in §\ref{sec:ten}.

The examples in~(\ref{exe:attr}) illustrate the attributive use of a monovalent stative non-agentive word (`rotten'), a monovalent dynamic non-agentive verb (`fallen'), a monovalent agentive verb (`running') and a bivalent verb (`reading').


\begin{exe}
	\ex
	\begin{xlist}
		\ex
		{\gll don \textbf{mun}=\textbf{ten} wandi=et ka bisa na=ta\\
			thing rotten=\textsc{at} like\_this={\glet} \textsc{2sg} can eat=\glta\\
			\glt `You can eat rotten things like this.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B91-5}{[narr39\_7:35]}}
		}
		\label{exe:munden2}
		\ex
		{\gll tumun \textbf{tur}=\textbf{ten} ma elao melalu\\
			child fall=\textsc{at} \textsc{3sg} down.\textsc{loc} sit\\
			\glt `The fallen child (he) sits on the floor.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_vvsa\_12]}}
		}
		\label{exe:turten}
		\ex
		{\gll tumun \textbf{tiri}=\textbf{ten} ladan kerkap=ten=at sabur\\
			child run=\textsc{at} shirt red=\textsc{at}=\textsc{obj} wear\\
			\glt `The running child wears a red shirt.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_vvsa\_13]}}
		}
		\label{exe:tiriren}
		\ex
		{\gll tumun doa=at \textbf{nabaca}=\textbf{ten} ladan kerkap=ten=at sabur\\
			child prayer=\textsc{obj} read=\textsc{at} shirt red=\textsc{at}=\textsc{obj} wear\\
			\glt `The child that is reading a prayer wears a red shirt.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_vvsa\_15]}}
		}
		\label{exe:nabacaren}
	\end{xlist}
	\label{exe:attr}
\end{exe}
%second example checked 2019: yes, correct transl.

Note that the use of \textit{=ten} on an attributively used verb is not obligatory. It is less common with monovalent stative non-agentive verbs than with other verbs, perhaps because they are semantically more fit to be used attributively. Lack of \textit{=ten} is especially common on the dimension words \textit{temun} `big' and \textit{cicaun} `small', as in~(\ref{exe:fibertemun}).

\begin{exe}
	\ex 
	{\gll mu fiber temun kerunggo\\
		\textsc{3pl} fibre\_boat big inside\\
		\glt `They are in a big fibre boat.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B93-C}{[conv14\_10:45]}}
	}
	\label{exe:fibertemun}
\end{exe}

Besides attributive \textit{=ten}, the placement of a demonstrative\is{demonstrative} can help disambiguate between an attributive and predicative reading. Noun + Verb + Demonstrative is interpreted as a NP\is{noun phrase} with an attributively used verb; Noun + Demonstrative + Verb and Pronoun + Verb is interpreted as a NP plus a predicate (with a verb in predicative use), as illustrated in~(\ref{exe:tabusik}).

\begin{exe}
	\ex 
	\begin{xlist}
		\ex {\gll {\ob}tumun tabusik wa{\cb}\textsubscript{\upshape NP}\\
			child short \textsc{prox}\\
			\glt `this short child'}
		\ex {\gll {\ob}tumun wa{\cb}\textsubscript{\upshape NP} {\ob}tabusik{\cb}\textsubscript{\upshape Pred}\\
			child \textsc{prox} short\\
			\glt `This child is short.}
		\ex {\gll {\ob}ma{\cb}\textsubscript{\upshape NP} {\ob}tabusik{\cb}\textsubscript{\upshape Pred}\\
			\textsc{3sg} short\\
			\glt `He is short.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_adj]}}}
	\end{xlist}
	\label{exe:tabusik}
\end{exe}

There are four valency-changing constructions and operations: reflexives, reciprocals, applicatives and causatives (§\ref{sec:val}. \is{verb|)}


\section{Nouns}\is{noun|(}
\label{sec:wcnoun}
The open class of nouns consists of words that can be marked with possessive morphology. Semantically, nouns refer to (animate) beings, places, substances and (abstract or inanimate) things. They head NPs, which typically function as arguments in the clause. (NPs may also occur in predicate function in nominal or equational clauses, see §\ref{sec:nonverbal}.) 

NPs function as the clausal arguments subject\is{subject} and object\is{object}. \textit{Esnem} `man' is the subject of an intransitive verb in~(\ref{exe:nSV}). \textit{Emun} `his/her mother' is the subject of a transitive verb and \textit{tumun} `child' is the object in~(\ref{exe:nAP}). Both direct and indirect object NPs are marked with the object postposition \textit{=at}. There is no gender marking and no obligatory number marking.

\begin{exe}
	\ex 
	{\gll esnem se lalat\\
		man {\glse} dead\\
		\glt `The man is dead.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC0-1}{[narr22\_2:52]}}
	}
	\label{exe:nSV}
	\ex
	{\gll emun tumun=at narorar\\
		mother.\textsc{3poss} child=\textsc{obj} take\_by\_hand\\
		\glt `The mother takes the child by the hand.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA8-B}{[stim4\_4:47]}}
	}
	\label{exe:nAP}
\end{exe}

Nouns are the head of the NP\is{noun phrase} and can be followed by one or more modifiers. (\ref{exe:towari}) and~(\ref{exe:pasaluren}) show an attributive modifier with and without attributive marker \textit{=ten}, respectively. (\ref{exe:sontumreidak2}) shows \is{quantifier}quantifier \textit{reidak} `many' modifying a noun, and in~(\ref{exe:tumtumkar}) numeral\is{numeral} quantifier \textit{karuok} `four' modifies a noun.

\begin{exe}
	\ex 
	{\gll {\ob}pebis \textbf{towari∼wari}{\cb}\textsubscript{\upshape NP} marua\\
		woman young∼\textsc{red} move\_seawards\\
		\glt `The young women went towards sea.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC1-0}{[narr19\_4:22]}}
	}
	\label{exe:towari}
		\ex
	{\gll {\ob}pasa \textbf{lu=ten}{\cb}\textsubscript{\upshape NP} mambon\\
		rice cold=\textsc{at} {\glmambon}\\
		\glt `Is there cold rice?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_3:36]}}
	}
	\label{exe:pasaluren}
	\ex 
	{\gll {\ob}sontum \textbf{reidak}{\cb}\textsubscript{\upshape NP} toni mu {\ob}...{\cb}\\
		person many say \textsc{3pl}\\
		\glt `Many people say they [...].' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA4-1}{[conv16\_1:04]}}
	}
	\label{exe:sontumreidak2}
	\ex 
	{\gll {\ob}tumtum \textbf{karuok}{\cb}\textsubscript{\upshape NP} marmar=i mia\\
		children four walk={\gli} come\\
		\glt `Four children come walking.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD1-D}{[stim31\_1:46]}}
	}
	\label{exe:tumtumkar}
\end{exe}

Some nouns require the insertion of a classifier on a numeral modifier (§\ref{sec:clf}).

Nouns are the only words that may carry possessive morphology. They can be both possessed (\textit{kewe} `house' in \ref{exe:nposd}) and possessors (\textit{kewe} `house' in \ref{exe:nposs}). The possessive suffix is always on the head and never on the dependent (§\ref{sec:possphr}).

\begin{exe}
	\ex
	\begin{xlist}
		\ex
		{\gll Malik kewe-un\\
			Malik house-\textsc{3poss}\\
			\glt `Malik's house' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE7-5}{[stim42\_10:46]}}
		}
		\label{exe:nposd}
		\ex
		{\gll kewe anggas-un\\
			house door-\textsc{3poss}\\
			\glt `the house's door' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE7-5}{[stim42\_12:29]}}
		}
		\label{exe:nposs}
	\end{xlist}
\end{exe}

If a noun inflected with a possessive suffix is in object position, the possessive suffix comes before the postposition\is{postposition}. See~(\ref{exe:eranat}).

\begin{exe}
	\ex
	\label{exe:eranat}
	{\gll et-an=at\\
		canoe-\textsc{1sg.poss=obj}\\
		\glt `my canoe' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9B-9}{[narr41\_1:05]}}
	}
\end{exe}

Nouns and NPs are described in Chapter~\ref{ch:noun}. That chapter also includes a description of Kalamang noun subclasses (§\ref{sec:nouncl}), which have some minor morphosyntactic differences but which mainly form semantic classes: proper nouns\is{count noun} and common nouns, alienable and inalienable nouns, mass and count nouns\is{count noun}, relational nouns, and kinship terms. Most kinship terms (§\ref{sec:kinterms}), for example, have a plural form. Proper nouns (§\ref{sec:prop}) differ from other nouns in that they cannot carry possessive morphology or be reduplicated. Common nouns may be reduplicated to create plurals, to indicate distributivity or extremity, or to create other nouns by associations (§\ref{sec:nounredup}).\is{noun|)}

\section{Pronouns}\is{pronoun|(}
\label{sec:wcprons}
The closed class of pronouns, like nouns, consists of NP heads which fill argument slots in the clause and which can carry postpositions\is{postposition}. Pronouns differ from nouns in four respects. Pronouns:

\begin{enumerate}
	\item take the suffixes \textit{-(a)hutak} `alone', \textit{-tain} `alone', \textit{-(nV)ninggan} `all' and prohibitive \textit{-mun};
	\item do not carry possessive suffixes;
	\item cannot be modified by an attributively used verb; %wait really? mu yalten? but seems to be the only example
	\item cannot be reduplicated.
\end{enumerate}

Kalamang pronouns distinguish first, second and third person; singular, dual\is{dual} and plural\is{plural!pronominal} number; and have a clusivity\is{clusivity} distinction in the first-person dual and plural. The pronominal paradigm is given in Table~\ref{tab:pron}.

\begin{table}
	\caption{Pronouns}
	\label{tab:pron}
	
		\begin{tabular}{l l l l}
			\lsptoprule
			&	\textsc{sg}	& \textsc{du} & \textsc{pl}\\\midrule
			1\textsc{in}&& \textit{pier} & \textit{pi}\\	
			1\textsc{ex}&	\textit{an}	& \textit{inier}  & \textit{in} \\
			2	&	\textit{ka}	& \textit{kier} &	\textit{ki}\\
			3	&	\textit{ma}	& \textit{mier} &	\textit{mu}\\ 
			\lspbottomrule
		\end{tabular}
	
\end{table}

Pronouns take several suffixes. The singular and plural forms take two restricting suffixes translatable as `alone': the restrictive/contrastive \is{focus!pronominal}focus suffix \mbox{\textit{-(a)hutak}} and the quantifying suffix \textit{-tain}. Plural pronouns take the suffix \textit{-(nV)\-ning\-gan} `all'. These and other characteristics of pronouns are described in Chapter~\ref{ch:pron}. Singular and plural forms also take prohibitive \textit{-mun} (§\ref{sec:negnondecl}).

Pronouns can substitute for a NP in subject and object position. Like nouns, they can be modified by, for example, a quantifier as in~(\ref{exe:inikon1}) or a demonstrative as in~(\ref{exe:warpas}).

\begin{exe}
	\ex 
	{\gll \textbf{in} \textbf{ikon} tok patin=nin=ten\\
		\textsc{1pl.excl} some yet wound=\textsc{neg}=\textsc{at}\\
		\glt `Some of us are not wounded yet.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC0-1}{[narr22\_7:10]}}
	}
	\label{exe:inikon1}
\ex 
{\gll \textbf{an} \textbf{yuwa} me warpas=a an=at melebor=ta\\
	\textsc{1sg} \textsc{prox} {\glme} witch=\textsc{foc} \textsc{1sg=obj} move\_to\_side={\glta}\\
	\glt `As for me, the witch shoved me aside.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE2-C}{[narr23\_4:56]}}
}
\label{exe:warpas}
\end{exe}

Pronouns do not carry possessive morphology. Instead there is a paradigm of possessive pronouns which can be used pronominally as well as adnominally (§\ref{sec:possprons}. The paradigm is given in Table~\ref{tab:posspronk} and an example with a possessive pronoun in object position is given in~(\ref{exe:possobj}).

\begin{table}
	\caption{Possessive pronouns}
	\label{tab:posspronk}
	
		\begin{tabular}{l l l}
			\lsptoprule
			&	\textsc{sg}	& \textsc{pl}\\\midrule
			1\textsc{in}&& \textit{pin}\\	
			1\textsc{ex}&	\textit{anggon}	&\textit{inggon} \\
			2	&	\textit{kain}	& \textit{kin}\\
			3	&	\textit{main}	& \textit{muin}\\ 
			\lspbottomrule
		\end{tabular}
	
\end{table}

\begin{exe}
	\ex
	{\gll ma \textbf{main=at} na\\
		\textsc{3sg} \textsc{3poss=obj} consume\\
		\glt `He drinks his [one/drink].' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA8-B}{[stim4\_2:07]}}
	}
	\label{exe:possobj}
\end{exe}

Although pronouns usually occupy the same slot as nouns, nouns and pronouns may co-occur to topicalise a referent. The pronoun can come both before and after the noun. Rather than proposing two slots here or analysing this as co-indexation, these seem to be cases of apposition\is{apposition} (see §\ref{sec:leftd}).

\begin{exe}
	\ex
	{\gll ma canam mat koluk\\
		\textsc{3sg} man \textsc{3sg.obj} meet\\
		\glt `He, the man, meets her.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE7-5}{[stim42\_3:42]}}
	}
	\label{exe:macanam}
	\ex {\gll canam ma se kaluar\\
		man \textsc{3sg} {\glse} exit\\
		\glt `The man, he has come out.'	\jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_22:55]}}
	}
	\label{exe:canamma}
\end{exe}\is{pronoun|)}
%tested 2019 if therse are interpreted differently. no. 

\section{Quantifiers}\is{quantifier|(}
\label{sec:wcquantnum}
Quantifiers are words indicating the quantity of the referent of a NP. Quantifiers can be subdivided into two groups: numerals (§\ref{sec:wcnums}) and non-numeral quantifiers like \textit{reidak} `many' and \textit{tebonggan} `all' (§\ref{sec:wcquant}). Members of this class behave as follows. Quantifiers:

\begin{enumerate}
	\item are adnominal modifiers in the NP which, if the referent is clear from the context, can also be the NP head;
	\item can be predicates in a quantifier clause;
	\item may in most cases carry the suffix \textit{-i} `\textsc{objqnt}' and \textit{-tak} `just, only'.
\end{enumerate}

(\ref{exe:numattr}) has the quantifier \textit{karuok} `four' as an adnominal modifier to \textit{kewe} `house'. Quantifiers have a fixed position in the NP (§\ref{sec:nomphr}); they follow the noun and possessive suffix, and precede possessive pronouns. In~(\ref{exe:numnom}), the nominal referent is clear from the context, and so the quantifier \textit{eir} `two' is the NP head.

\begin{exe}	
	\ex
	{\gll kewe=at \textbf{karuok-i} in paruo\\
		house=\textsc{obj} three-\textsc{objqnt} \textsc{1pl.excl} make\\
		\glt `Three houses we made.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_0:53]}}
	}
	\label{exe:numattr}
	\ex
	{\gll \textbf{eir} nau=namanghadap\\
		two \textsc{rec}=face\\	
		\glt `Two [animals] face each other.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAB-6}{[stim14\_2:09]}}
	} 
	\label{exe:numnom}		
\end{exe}

(\ref{exe:inreidak}) illustrates the use of quantifier \textit{reidak} `many' in predicate function.

\begin{exe}
	\ex 
	{\gll in \textbf{reidak}\\
		\textsc{1pl.excl} many\\
		\glt `We were many.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_1:56]}}
	}
	\label{exe:inreidak}
\end{exe}

The suffix \textit{-i} occurs on quantifiers in object\is{object} position, as illustrated in~(\ref{exe:mingtun}).

\begin{exe}
	\ex 
	{\gll mingtun=at \textbf{bolon}-\textbf{i} ko=yuon\\
		palm\_oil=\textsc{obj} little-\textsc{objqnt} \textsc{appl}=rub\\
		\glt `[You] rub a little palm oil on.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE4-8}{[narr31\_1:33]}}
	}
	\label{exe:mingtun}
\end{exe}

However, when another modifier follows the quantifier, \textit{-i} does not occur.

\begin{exe}
	\ex \gll mu {\ob}yap\_seran-un nak-\textbf{eir} met{\cb}\textsubscript{\upshape NP} toni pasor=ta eba mu nan=et\\
	\textsc{3pl} yam-\textsc{3poss} \textsc{clf\_fruit1}-two \textsc{dist.obj} want fry={\glta} then \textsc{3pl} consume={\glet}\\
	\glt `They want to fry those two yams and then they want to eat.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_16:02]}}
	\label{exe:nairmett}
\end{exe}

The suffix \textit{-tak} `just, only' is illustrated in~(\ref{exe:kodak}). When suffixed to \textit{kon} `one', it results in \textit{kodak} instead of the expected \textit{kondak}.\footnote{Note the similarity of this suffix with the last part of the pronominal suffix \textit{-autak} `alone', Chapter~\ref{ch:pron}, and the last part of \textit{reidak} `many', example~\ref{exe:inreidak}.}

\begin{exe}
	\ex
	{\gll som \textbf{kodak} pas \textbf{kodak}\\
		person only\_one woman only\_one\\	
		\glt `Only one person, only one woman.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_10:42]}}
	} 
	\label{exe:kodak}
\end{exe}

\subsection{Numerals}\is{numeral|(}
\label{sec:wcnums}
Numerals are a subclass of quantifiers with the following additional diagnostics. Numerals, but not other quantifiers:

\begin{enumerate}
	\item may be reduplicated to create distributive numerals;
	\item can carry a classifier prefix;
	\item can carry the suffix \textit{-gan} `all'.
\end{enumerate}

Kalamang numerals from one to ten are given in Table~\ref{tab:nums}. Higher numerals, information on the formation of cardinals, and other details about numerals can be found in §\ref{sec:cardnum}.

\begin{table}[ht]
	\caption{Numerals 1--10}
	\label{tab:nums}
	
		\begin{tabular}{l l}
			\lsptoprule 
			1 & \textit{kon}\\
			2 & \textit{eir}\\
			3 & \textit{karuok}\\
			4 & \textit{kansuor}\\
			5 & \textit{ap}\\
			6 & \textit{raman}\\
			7 & \textit{ramandalin}\\
			8 & \textit{irie}\\
			9 & \textit{kaninggonie}\\
			10& \textit{putkon} \\\lspbottomrule 
		\end{tabular}
	
\end{table}

Reduplication\is{reduplication} of numerals denotes distributivity (indicating that the number applies to each referent, or that the referents are divided into groups of the size the numeral indicates, \citealt{gil2013}). \is{distributive}Distributive numerals can be in attributive\is{attribute} function following the noun, as in~(\ref{exe:dstrbnum}), but may also be used predicatively. %predicative use: see q wc 2019, ror tebonggan pusun putkonkon.
\is{reduplication!numerals}

\begin{exe}
	\ex
	{\gll goni-kinkin kilo \textbf{putkon\_ba\_ap∼ap=kin} Bulog yuwane\\
		sack-small kilo fifteen∼\textsc{distr=poss} Bulog \textsc{prox}\\	
		\glt \glt `[It] comes in small sacks of fifteen kilos each, this Bulog rice.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_0:12]}}
	} 
	\label{exe:dstrbnum}		
\end{exe}

When a numeral modifies certain nouns, a classifier is used. Kalamang has sixteen classifiers associated with different individual nouns or groups of nouns, but not all nouns are associated with a classifier (§\ref{sec:clf}). The classifier is prefixed to the numeral quantifier. A classifier + numeral combination always directly follows the noun, as the fruit classifier \textit{tep} and the numeral \textit{eir} `two' in~(\ref{exe:tewirr}).

\begin{exe}
	\ex \gll im tep-eir\\
	banana \textsc{clf\_fruit2}-two\\
	\glt two bananas \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD6-8}{[stim38\_10:32]}}
	\label{exe:tewirr}
\end{exe}

Numerals, including pronouns inflected with a numeral (§\ref{sec:wcprons}), can carry the suffix \textit{-gan} `all'. This suffix is most commonly used with \textit{eir} `two' to create \textit{eirgan} `both', as in~(\ref{exe:eirgan}).

\begin{exe}
	\ex
	{\gll
		kon=a ma tan-un=at \textbf{eir}-\textbf{gan} madong\\
		one=\textsc{foc} \textsc{3sg} hand-\textsc{3poss=obj} two-all stretch\\
		\glt `One, he stretches out both his arms.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B90-D}{[stim45\_1:28]}}}
	\label{exe:eirgan}
\end{exe}
%so \textendash gan and \textendash i are incompatible, it seems. two occurrences in corpus.

Nouns cannot carry \textit{-gan}. For instance, to say `all houses' one cannot say `house-\textit{gan}'; on must instead use the quantifier \textit{tebonggan} `all', as illustrated below.

\begin{exe}
	\ex
	\label{exe:kewegan}
	\begin{xlist}
		\ex [*]
		{\gll kewe-gan temun\\
			house-all big\\
			\glt `All houses are big.' \jambox*{[elic]}
		}
		\ex []
		{\gll kewe tebonggan temun\\
			house all big\\
			\glt `All houses are big.' \jambox*{[elic]}
		}
	\end{xlist}
\end{exe}

\is{numeral|)}


\subsection{Non-numeral quantifiers}
\label{sec:wcquant}
Kalamang has six non-numeral quantifiers, listed in Table~\ref{tab:quant}. For some comments on their individual properties, see §\ref{sec:quantall}.  Unlike numeral classifiers, these cannot be reduplicated to create a \is{distributive}distributive quantifier and they cannot be preceded by a classifier. 

\begin{table}[ht]
	\caption{Non-numeral quantifiers}
	\label{tab:quant}
	
		\begin{tabular}{l l}
			\lsptoprule 
			\textit{bolon} & a little \\
			\textit{taukon, ikon} & some\\
			\textit{reidak} & much/many\\
			\textit{reingge} & not much/many\\
			\textit{tebonggan} & all \\\lspbottomrule		
		\end{tabular}
	
\end{table}

\is{quantifier|)}

\section{Demonstratives}\is{demonstrative|(}
\label{sec:wcdem}
Demonstratives form a closed class of words that locate a referent in space, time or discourse in relation to the deictic centre. Pronominally, demonstratives take the slot of the noun; identificationally, demonstratives occur in copula\is{copula*!clause} and non-verbal clauses; adnominally, demonstratives occur after the noun. There are six demonstratives, which behave differently with respect to their ability to occur in these slots. Table~\ref{tab:demwc} gives an overview of the demonstrative roots and their syntactic use. There are five demonstrative roots whose main function is to locate a referent in space in relation to the deictic centre. \textit{Wa} is generally associated with proximal reference, and \textit{me} with more distal reference. These may occur pronominally, adnominally and identificationally. A third deictic demonstrative root, far distal \textit{owa}, only occurs adnominally and identificationally, and occurs much less frequently than \textit{wa} and \textit{me}. \is{demonstrative!elevational}Elevationals \textit{yawe} `\textsc{down}' and \textit{osa} `\textsc{up}' occur adnominally and identificationally. In addition, there is an anaphoric demonstrative, \textit{opa}, which is only used adnominally. The description of demonstratives is expanded upon in Chapter~\ref{ch:dems}.%but identificational yawe/osa only as predicate, not as argument. -> rename predicative? much more gjennomsiktig.

%evt use fact that obj form has =t instead of =at as diagnostic.

\begin{table}[ht]
	\caption{Demonstratives and their syntactic distribution}
	\label{tab:demwc}
	
	\begin{tabular}{l l l l}
		\lsptoprule 
		form&pronominal&adnominal&identificational\\
		\midrule
		\textit{wa} proximal& $+$ & $+$ & $+$\\
		\textit{me} distal&  $+$ & $+$ & $+$\\
		\textit{owa} far distal&&$+$&$+$\\
		\textit{yawe} `\textsc{down}' &&$+$&$+$\\
		\textit{osa} `\textsc{up}' &&$+$&$+$\\
		\textit{opa} anaphoric &  & $+$&\\ \lspbottomrule		
	\end{tabular}

\end{table}

The same root is used in all positions, as illustrated below for proximal \textit{wa}. \textit{Wa} and \textit{me} as adnominal and pronominal demonstratives are infrequent; almost all corpus occurrences of \textit{wa} and \textit{me} are identificational. The three uses are illustrated in (\ref{exe:demad}) (adnominal), (\ref{exe:demad2}) (pronominal), (\ref{exe:demid}) (identificational in argument function) and~(\ref{exe:demidpred}) (identificational in predicate function). When used adnominally or identificationally in argument function, the demonstrative is almost invariably followed by \textit{me} `\glme'.

\begin{exe}
	\ex
	{\gll goras \textbf{wa} me bara=te kul nerunggo sor=at jie\\
		crow \textsc{prox} {\glme} descend={\glte} basket inside fish=\textsc{obj} get\\	
		\glt `This crow went down into the basket and got a fish.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C9B-8}{[stim3\_0:05]}}
	}
	\label{exe:demad}	
	\ex
	{\gll ma \textbf{wat} paruo\\
		\textsc{3sg} \textsc{prox.obj} make\\	
		\glt `She made this.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB5-B}{[conv17\_8:37]}}
	}
	\label{exe:demad2}			
	\ex
	{\gll \textbf{wa} me yartep\\
		\textsc{prox} {\glme} sardine\\	
		\glt `This is a sardine.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C75-D}{[stim15\_2:03]}}
	}
	\label{exe:demid}
	\ex 
	{\gll	namun=a \textbf{wa} kiun=a \textbf{wa} tumun-un=a \textbf{wa}\\
		husband\textsc{.3poss=foc} \textsc{prox} wife\textsc{.3poss=foc} \textsc{prox} child\textsc{-3poss=foc} \textsc{prox}\\
		\glt `Her husband is this, his wife is this, their child is this.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA9-9}{[stim6\_19:23]}}}
	\label{exe:demidpred}	
\end{exe}

The \is{demonstrative!elevational}elevationals \textit{yawe} `\textsc{down}' and \textit{osa} `\textsc{up}' can only be used adnominally and identificationally. This is illustrated for \textit{yawe} `\textsc{down}' in~(\ref{exe:yawekom}) (adnominal) and~(\ref{exe:lempyawe}) (identificational).

\begin{exe}
	\ex
	{\gll ma tama ka bo minggalot-an \textbf{yawe} komere\\
		\textsc{3sg} where \textsc{2sg} go bedroom-\textsc{1sg.poss} \textsc{down} look\\	
		\glt `Where is it? You go look in my bedroom down there.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB5-B}{[conv17\_3:42]}}
	}
	\label{exe:yawekom}
	\ex \gll lempuang \textbf{yawe}\\
		island \textsc{down}\\	
		\glt \glt `The island is down there.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE8-0}{[stim43\_2:22]}}
	\label{exe:lempyawe}
\end{exe}

The demonstrative roots \textit{wa}, \textit{me}, \textit{owa}, \textit{yawe} and \textit{osa} can take the lative enclitic \textit{=ka} and the locative \textit{=ko} (§\ref{sec:demcase}). The roots \textit{wa} and \textit{me} are also used to create manner, quality, quantity and degree demonstratives (§\ref{sec:mqqd}).

Finally, there is a demonstrative \textit{opa}, which is always used adnominally. It signals that a referent represents shared knowledge between the speaker and addressee. This demonstrative is further described in §\ref{sec:demopa} and~\textcite{visser2020}. An example is given in~(\ref{exe:udangopa}).


\begin{exe}
	\ex {\gll udang \textbf{opa} ka kuet=nin\\
		lobster {\glopa} \textsc{2sg} get=\textsc{neg}\\
		\glt `Didn't you get that lobster (we talked about earlier)?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCB-5}{[conv1\_6:05]}}}
	\label{exe:udangopa}
\end{exe}

\is{demonstrative|)}


\section{Postpositions}
\label{sec:wcpost}
Postpositions\footnote{These are analysed as postpositions, not case markers, because of the domain of attachment (the NP, not the noun), a lack of agreement or declension paradigms, and because the forms are almost mutually exclusive. The author has no strong preference for either analysis, and encourages the reader to replace ``postposition'' with ``case marker'' in their mind if they wish.}\is{postposition} mark the function of core and peripheral argument noun phrases.\is{argument!core}\is{argument!peripheral} Kalamang has nine postpositions: object \textit{=at}, comitative \textit{=bon}, instrumental \textit{=ki}, benefactive \textit{=ki}, similative \textit{=kap}, locative \textit{=ko}, animate locative \textit{=konggo}, lative \textit{=ka} and animate lative \textit{=kongga}. They are enclitics that attach to the right edge of the NP.\is{noun phrase} The postpositions are further described in Chapter~\ref{ch:noun}. References to the specific sections are given in Table~\ref{tab:wcpos}.


\begin{table}[!ht]
	\caption{Postpositions}
	\label{tab:wcpos}

		\begin{tabular}{llll}
			\lsptoprule 
			form & function & gloss & reference\\ \midrule 
			∅ & subject (S, A) & \textendash   & \textendash \\
			\textit{=at} & object (O) & \textsc{obj} &   §\ref{sec:at}\\
			\textit{=bon} & comitative & \textsc{com} & §\ref{sec:comi}\\
			\textit{=ki} & instrumental, & \textsc{ins} & §\ref{sec:ins}\\
			& benefactive& \textsc{ben}& §\ref{sec:ben}\\
			\textit{=kap} & similative & \textsc{sim} & §\ref{sec:simcase}\\
			\textit{=ko} & locative & \textsc{loc} & §\ref{sec:loc}\\
			\textit{=ka} & allative, ablative & \textsc{lat} & §\ref{sec:lat}\\	
			\textit{=konggo} & animate locative & \textsc{an.loc} & §\ref{sec:animloclat}\\		 			
			\textit{=kongga} & animate allative, ablative & \textsc{an.lat} & §\ref{sec:animloclat}\\
			\lspbottomrule
		\end{tabular}
	
\end{table}

\section{Adverbial modifiers}\is{modifier!adverbial}
\label{sec:wcadv}
Adverbial modifiers is a cover term for morphemes that modify the clause by changing the mood, aspect or mode of a predicate or clause; or specify the manner, temporal setting, degree or other characteristics of the state or event expressed by the verb, such as \is{repetition}repetition or exclusivity. They are independent words, proclitics, enclitics and suffixes that occupy different slots in the clause. They do not form a real word class, but are given here for ease of reference. Adverbial modifiers can be divided into functional groups such as modal adverbials or manner adverbials. See Chapter~\ref{ch:clausmod} for an overview of all morphemes and their positions in the clause. One important position is the post-subject position, which is taken by \is{iamitive}iamitive \textit{se}, \is{nondum}nondum \textit{tok},\footnote{See \textcite{olsson2013} and \textcite{auwera2013} for a definition of and information about the categories iamitive and nondum, respectively.} \textit{gen} `maybe' and some modal adverbials. The predicate can host a number of clitics, including mood and aspect markers. Clausal negation is also marked as a clitic on the predicate (described in §\ref{sec:clauneg}).

The following examples illustrate post-subject iamitive \textit{se} and \textit{koi} `again' in (\ref{exe:hekoi}), clause-final \textit{reon} `maybe' in (\ref{exe:gongreon}), clause-initial temporal adverbials and progressive \textit{=teba} and volitional \textit{=kin} on the predicate in (\ref{exe:kasurm}).

\begin{exe}
	\ex 
	{\gll mu \textbf{se} \textbf{koi} yal\\
		\textsc{3pl} {\glse} again paddle\\
		\glt `They already paddled again.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC1-0}{[narr19\_10:06]}}
	}
	\label{exe:hekoi}
	\ex 
{\gll esa anggon temun=a an=at gonggung \textbf{reon}\\
	father \textsc{1sg.poss} big=\textsc{foc} \textsc{1sg=obj} call maybe\\
	\glt `Maybe my uncle called me.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_9:20]}}
}
\label{exe:gongreon}
\ex
{\gll ma toni \textbf{opa\_yuwa} an dodon waruo=\textbf{teba} \textbf{kasur} mu kolak=ka bot=\textbf{kin}\\
	\textsc{3sg} say today \textsc{1sg} clothes wash={\glteba} tomorrow \textsc{3pl} mountain=\textsc{lat} go={\glkin}\\
	\glt `She said: ``Today I'm washing clothes.'' Tomorrow they want to go to the mountains.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA2-F}{[conv11\_6:15]}}
}
\label{exe:kasurm}
\end{exe}

\section{Question words}\is{question word|(}
\label{sec:wcq}
Question words or content interrogatives are words that are used to make non-polar questions, indicating the part of the proposition that the speaker is asking about. They form a closed class. Question words take the slot of the constituent that the speaker is asking about. They cut across several word classes \parencite[][33]{schachter2007}, with Kalamang having nominal, adnominal, predicative and adverbial question words, and one quantifier question word. Their main shared feature is that they structurally replace other constituents, which is why they are discussed together here. As for their form, all question words but \textit{neba} `what' contain a sequence \textit{ama}, and all but \textit{neba} and \textit{naman} `who' contain the sequence \textit{tama} or its allophonic variant \textit{rama}. The question words are listed in Table~\ref{tab:wcques}. \textit{Neba} `what' is a very generic question word which can question both items and actions. It also functions as a \is{placeholder}placeholder for any constituent, see §\ref{sec:ph}.\is{wh-question|see {question-word question}}\is{interrogative}\is{placeholder}\is{question word question}

\begin{table}
	\caption{Question words}
	\label{tab:wcques}
	\fittable{
	\begin{tabular}{l l l l}
		\lsptoprule 
		&& questions & syntactic identity\\ \midrule
		\textit{naman} & who & person & pronominal\\
		\textit{neba} & what & generic & pronominal\\
		\textit{tama} & which & item/thing & adnominal\\
		\textit{tamangga}& from/to where & goal/source & predicative\\
		&&& (with other verb, §\ref{sec:lat}, §\ref{sec:mvcgoal})\\
		\textit{tamatko} & where & location & predicative\\
		\textit{tamandi} & why/how & reason/manner & adverbial\\
		\textit{puraman} & how many & quantity & quantifier\\ \lspbottomrule		
	\end{tabular}
}
\end{table}

\textit{Naman} `who' and \textit{neba} `what' are nominal, taking the place of subjects and objects\is{in situ}, with the ability to carry postpositions\is{postposition} (§\ref{sec:wcnoun}). \textit{Naman} replaces pronouns, being unable to carry possessive morphology, but having the possibility to carry generic possessive \textit{=kin} or animate lative and locative postpositions \textit{=kongga} and \textit{=konggo}, respectively. \textit{Naman} and \textit{neba}, as well as \textit{tama} below, can be followed by a \is{focus}focus marker \textit{=ba}. In question-answer pairs, the answer may then also be followed by \textit{=ba}, as in~(\ref{exe:namanba}).


\begin{exe}
	\ex 
	\begin{xlist}
		\exi{A:}
		{\gll \textbf{naman}=ba\\
			who={\glba}\\
			\glt `Who?'
		}
		\exi{B:}
		{\gll an=ba\\
			\textsc{1sg}={\glba}\\
			\glt `Me.' \jambox*{[overheard]}
		}
	\end{xlist}
	\label{exe:namanba}
\end{exe}

\begin{exe}
	\ex \gll mena mu \textbf{neba}=at=a paruot=kin\\
	then \textsc{3pl} what=\textsc{obj=foc} do={\glkin}\\
	\glt `Then, what do they want to do?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B98-6}{[narr45\_3:01]}}
	\label{exe:menamu}
\end{exe}	

\textit{Tama} `which' usually modifies the noun in the NP, replacing a demonstrative. \textit{Neba} does not occur replacing demonstratives.

\begin{exe}
	\ex 
	\begin{xlist}
		\exi{A:}
		{\gll anggas=at tarat-un eranun\\
			door=\textsc{obj} close-\textsc{nmlz} cannot\\
			\glt `Couldn't you close the door?'
		}
		\exi{B:}
		{\gll anggas \textbf{tama}\\
			door which\\
			\glt `Which door?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_17:05]}}
		}
	\end{xlist}
	\label{exe:tama}
\end{exe}

\textit{Tamangga} `to/from where' and \textit{tamatko} `where' are formed with \textit{tama} plus the lative or the locative enclitic, respectively (for the morphosyntax see §\ref{sec:problems}). They have verbal properties. \textit{Tamangga}, like other words carrying lative \textit{=ka}, is primarily used in \is{predicate!complex} complex predicates (see Chapter~\ref{ch:svc}), for example with \textit{jie} `get' in~(\ref{exe:tamjie}). \textit{Tamatko}, like other words carrying locative \textit{=ko}, is usually in predicate function. It can stand on its own, as illustrated in~(\ref{exe:tamatko}), or be accompanied by another verb. A short form of \textit{tamatko} is \textit{tama}. These suffixes are also described in §\ref{sec:lat} and~\ref{sec:loc}.\is{question word}\is{location}

\begin{exe}
	\ex 
	{\gll ka \textbf{tamangga}=ta jie\\
	\textsc{2sg} from\_where={\glta} get\\
	\glt `Where did you get [it] from?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCB-5}{[conv1\_5:23]}}
}
\label{exe:tamjie}
	\ex
{\gll ema \textbf{tamatko}\\
	mother where\\	
	\glt `Where is mother?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_5:19]}}
}
\label{exe:tamatko}
\end{exe}

\textit{Tamandi} is an adverbial question word. It is illustrated modifying \textit{komele} `burn' in~(\ref{exe:qadv}), and \textit{paruo} `do' in~(\ref{exe:babut}) below. In addition, it is used as a \is{greeting}greeting with the meaning `how are you doing?'. It seems to contain the suffix \textit{-ndi}, which is also found on manner demonstratives (§\ref{sec:mqqd}).


\begin{exe}
	\ex
	{\gll din saerak in \textbf{tamandi}=a lampur=at komele\\
		fire \textsc{neg\_exist} \textsc{1pl.excl} how=\textsc{foc} lamp=\textsc{obj} burn\\	
		\glt `There was no fire, how could we burn the lamps?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_0:33]}}
	}
	\label{exe:qadv}		
\end{exe}

\textit{Puraman} `how many' replaces a quantifier. (\ref{exe:pel}) illustrates \textit{puraman} following a classifier (§\ref{sec:clf}).

\begin{exe}
	\ex 
	{\gll wat \textbf{nak}-\textbf{puraman-i} mindi kajie\\
		coconut \textsc{clf\_fruit1}-how\_many-\textsc{objqnt} like\_that pick\\
		\glt `How many coconuts did [we] pick like that?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA2-F}{[conv11\_4:50]}}
	}
	\label{exe:pel}
\end{exe}


Note that there is no question word for time. \textit{Yuol tama} `which day' was elicited, but does not occur in the natural speech corpus. There is one occurrence of \textit{yuol puraman} `which day'.

\begin{exe}
	\ex \gll ma toni yuol puraman usar=et\\
	\textsc{3sg} say day how\_many build\_frame\_house={\glet}\\
	\glt \glt `Which day do [we] install the frame of the house?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C78-C}{[narr46\_2:11]}}
\end{exe}

The syntax of questions with question words is further described in §\ref{sec:opq}.\is{question word|)}

\section{Conjunctions}\is{conjunction|(}
\label{sec:wconj}
Conjunctions are words that connect words, phrases or clauses \parencite[][45]{schachter2007}. Kalamang has sequential, disjunctive, adversative, consequence and reason, concessive and conditional conjunctions. Coordination achieved with `and' in English is done by apposition\is{apposition} of two words, phrases or clauses in Kalamang, or, in the case of noun coordination, with comitative postposition\is{postposition!comitative} \textit{=bon} (§\ref{sec:comi}). Clause conjunction is described in depth in §\ref{sec:clauseconj}. The three most common conjunctions are introduced here.

\textit{Ba} `but' is an adversative coordinator\is{conjunction!adversative}, as illustrated in~(\ref{exe:babut}). It is also used to conjoin numerals\is{numeral} between 11 and 29, as described in §\ref{sec:cardnum}.

\begin{exe}
	\ex
{\gll ma muawese \textbf{ba} ma tamandi paruot=et\\
	\textsc{3sg} hungry but \textsc{3sg} how do={\glet}\\	
	\glt `He's hungry but what can he do?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBA-8}{[stim2\_4:31]}}
}
\label{exe:babut}
\end{exe}		

\textit{Ye} `or' is a disjunctive coordinator\is{conjunction!disjunctive}. Intonationally, it belongs to the first clause. In addition to its coordinative function, it can also be used as a tag `maybe'. Both uses are illustrated in~(\ref{exe:ye}).

\begin{exe}
	\ex
	{\glll Kalau sontum tur \textbf{ye}, don muat pue \textbf{ye}.\\
		kalau sontum tur ye don mu=at pue ye\\
		if person fall or thing \textsc{3pl=obj} hit maybe\\	
		\glt `If a person falls, or maybe a thing hits him...' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB1-3}{[narr34\_4:31]}}
	}
	\label{exe:ye}
\end{exe}

\textit{Eba} `then' is used to connect two clauses that refer to sequential events\is{conjunction!sequential}. Intonationally, it belongs to the second clause. \textit{Koi} `then' also expresses sequential meaning, but is often placed after the subject, although it can also be clause-initial when the subject is elided. There is no restriction on the combination of \textit{eba} and \textit{koi}. (\ref{exe:eba}) illustrates post-subject \textit{koi} and \textit{eba}. (\ref{exe:koicllnk}) illustrates \textit{koi} clause-initially in an example where there is no subject because it is from a procedural text explaining how to build a traditional house. \textit{Koi} is also an adverbial that means `again', see §\ref{sec:wcadv}.

\begin{exe}
	\ex
	{\glll In \textbf{koi} timunat potma, timunat potma bara melalu, \textbf{eba} in sensurgi kaborunat parair.\\
		in koi timun=at potma timun=at potma bara melalu eba in sensur=ki kabor-un=at parair\\
		\textsc{1pl.excl} then tip=\textsc{obj} cut tip=\textsc{obj} cut descend sit then \textsc{1pl.excl} chainsaw=\textsc{ins} stomach-\textsc{3poss=obj} chop\\	
		\glt `Then we cut the tips, cut the tips and put [the canoe] down, then we chop its stomach with the chainsaw.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB4-6}{[narr14\_1:39]}}
	}
	\label{exe:eba}
	\ex {\gll saban potmat=et \textbf{koi} kanggaran paruot=et\\
	bamboo cut={\glet} then bamboo\_floor make={\glet}\\
	\glt `[One] cuts bamboo, then [one] makes the bamboo floor.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB2-1}{[narr6\_4:51]}}
}
\label{exe:koicllnk}
\end{exe}

\is{conjunction|)}


\section{Interjections}\is{interjection|(}
\label{sec:wcyesno}
Interjections are conventionalised utterances expressing spontaneous emotions or reactions. They are extra-grammatical and can be utterances on their own or they can be apposed to a sentence. They include cries of disgust or surprise, as well as greetings and words like `yes' and `no'.

Kalamang has a number of interjections that I have classified in a few semantic (and partly phonological) subcategories. These are listed here, together with their gloss used in the corpus, some words on their use and their phonological form. A non-exhaustive list of common interjections, their approximate form and their gloss is given in Table~\ref{tab:inter}. More information about the phonetics of Kalamang interjections can be found in §\ref{sec:int}. They are further described and illustrated in §\ref{sec:discinterj}.

\begin{table}
	\caption{Interjections and their gloss}
	\label{tab:inter}
		\begin{tabularx}{\textwidth}{XXl}
			\lsptoprule
			(typical) form(s) & function &  gloss \\ \midrule
			\textit{a}, \textit{e} & filler	&  \textsc{fil}\\
			\textit{a (a)}, \textit{m (m)}, \textit{yo}, \textit{ya} & agreement &  yes\\
			\textit{adi(h)} & pain, discomfort & \textsc{pain}\\	
			\textit{(a)dih}, \textit{(a)deh} & contempt, dissatisfaction  & \textsc{int.pej}\\			
			\textit{e} & various & \textsc{int.e}\\
			\textit{eh} & introduce quote & \textsc{quot}\\	
			\textit{ema} & surprise, contempt &  \textsc{surpr}\\	
			\textit{ge} & disagreement & no(t)\\	
			\textit{ha}	& repair initiator & what\\
			\textit{hi} & enjoyment & yay\\
			\textit{i(h)}, \textit{e(h)} & confirmation-seeking & \textsc{tag}\\
			\textit{inye} & contempt, dissatisfaction & \textsc{int.pej}\\	
			\textit{kan}, \textit{adu(h)} & various Malay loans &  \textsc{int.mly} \\		
			\textit{mera} & downplay, obviousness & \textsc{int}\\ %sudah mu			
			\textit{mo} & softener & \textsc{soft}\\
			\textit{o} & emphatic &  \textsc{emph}\\
			\textit{o(h)} & surprise & \textsc{surpr}\\
			\textit{some} & (annoyed) encouragement & \textsc{enc}\\ %sudah mu
			\textit{uei} & surprise &  \textsc{surpr}\\
			\textit{ya aula} & (feigned) incredulity & my god\\
			\lspbottomrule
		\end{tabularx}

\end{table}

\newpage
Below I elaborate on the function of some of these interjections.\is{filler}

\begin{itemize}
	\item Filler interjections (\textsc{fil}) serve to keep the floor while thinking.
	\item Agreement interjections (`yes') serve to express agreement. Sometimes translatable as `yes', often more appropriately `yeah', `OK', `go on (I'm listening)'. 
	\item The confirmation-seeking tag (\textsc{tag}) is described in §\ref{sec:polq}.
	\item Emphatic \textit{o} `\textsc{emph}' is typically clause-final, with an \is{intensification}intensifying or reinforcing function.
	\item Multi-functional interjections glossed as \textsc{int.e} express among other things resignment or encouragement and are used for \is{intensification}intensifying commands or statements, like \textit{o} `\textsc{emph}'.
	\item The gloss \textsc{int.mly} is used for interjections borrowed from Malay, usually \textit{kan} `right' or \textit{adu(h)}, an expression of pain or disappointed surprise. Confirmation-seeking tag \textit{to(h)} is glossed as `right' and is treated in §\ref{sec:polq}.
\end{itemize}

The remainder of interjections in the corpus that are not listed here are simply glossed as \textsc{int}, and include interjections of surprise like [oi] or [ui], calls like [ei], and a number of [o]'s indicating relief or recognition. 

Conventions around greeting each other, asking for permission to leave, giving thanks and ending a narrative, among other things, are described in Chapter~\ref{ch:inf}.\is{interjection|)}\is{word class|)}
