\chapter{Pronouns and person reference and address}\is{person reference|(}
\label{ch:pron}
This chapter examines person reference and address. It is divided between pronominal and non-pronominal reference and address. Pronouns were introduced in §\ref{sec:wcprons}. They are NP heads that can carry postpositions and a number of pronoun-specific suffixes. They do not carry possessive suffixes and are not reduplicated. In §\ref{sec:pronref}, the different Kalamang pronominal paradigms are given, with examples of their use. Non-pronominal person reference and address involves either kinship terms or names as well as subclasses of nouns, and is described in §\ref{sec:nonpronref}.

\section{Pronominal person reference and address}\is{pronoun|(}
\label{sec:pronref}
Kalamang pronouns distinguish between first, second and third person and between singular, dual\is{dual} and plural\is{plural!pronominal} number. They also have a clusivity\is{clusivity} distinction in the first-person dual and plural. There are five series of pronouns, listed in Table~\ref{tab:pronouns}: basic pronouns, quantifying pronouns, restricting pronouns, collective pronouns and possessive pronouns. The basic pronouns are free-standing forms; person and number are not marked on the verb in general. Dual pronouns are formed by suffixing \textit{-ier} to the plural basic pronouns. The basic pronouns are the only pronouns that have a dual\is{dual} form. There are two series of pronouns made with suffixes that roughly mean `alone' or `self'. The \textit{-(ah)utak} series has a quantifying reading, whereas the \textit{-tain} series has a restrictive or contrastive focus meaning. Collective pronouns are only available for plural\is{plural!pronominal} forms, and are formed by suffixing \textit{-(nV)ninggan} to the basic pronouns. Kalamang possessive pronouns consist of the basic pronouns plus an element \textit{-gon} (first-person singular and first-person plural exclusive) or \textit{-in} (other forms).

\begin{table}
	\caption{Pronominal paradigms}
	\label{tab:pronouns}
\subfigure[Basic]{
		\begin{tabularx}{.45\textwidth}{l Q Q l}
			\lsptoprule
			&	\textsc{sg}	& \textsc{du} & \textsc{pl}\\			\midrule
			1	&	\textit{an}	& \textsc{ex}: \textit{inier}  &	\textsc{ex}: \textit{in} \\
			&& \textsc{in}: \textit{pier} & \textsc{in}: \textit{pi}\\
			2	&	\textit{ka}	& \textit{kier} &	\textit{ki}\\
			3	&	\textit{ma}	& \textit{mier} &	\textit{mu}\\
			\lspbottomrule
                \end{tabularx}
}
\subfigure[Quantifying \textit{-(ah)utak}]{
		\begin{tabularx}{.45\textwidth}{l Q Q}
			\lsptoprule
			& \textsc{sg} &\textsc{pl} \\
			\midrule
			1 & \textit{an(ah)utak} & \textsc{ex}: \textit{in(h)utak} \\
			&& \textsc{in}: \textit{pi(h)utak}\\
			2 & \textit{ka(h)utak} & \textit{ki(h)utak} \\
			3 & \textit{ma(h)utak} & \textit{mu(h)utak} \\
			\lspbottomrule
			\end{tabularx}
}
\subfigure[Restrictive/contrastive focus \textit{-tain}]{
		\begin{tabularx}{.45\textwidth}{l Ql}
			\lsptoprule
			& \textsc{sg} &\textsc{pl} \\
			\midrule
			1 & \textit{andain} & \textsc{ex}: \textit{indain} \\
			&& \textsc{in}: \textit{pirain}\\
			2 & \textit{karain} & \textit{kirain}\\
			3 & \textit{marain} & \textit{murain}\\
			\arrayrulecolor{black}
			\lspbottomrule
			\end{tabularx}
}
\subfigure[Collective]{
		\begin{tabularx}{.45\textwidth}{l Q l}
		\lsptoprule
			& \textsc{sg} &\textsc{pl} \\
			\midrule
			1 && \textsc{ex}: \textit{inaninggan}\\
			&& \textsc{in}: \textit{pinaninggan}\\
			2 && \textit{kinaninggan}\\
			3 && \textit{munaninggan}\\
			\lspbottomrule
                \end{tabularx}
}
\subfigure[Possessive]{
		\begin{tabularx}{.45\textwidth}{l Q l}
		\lsptoprule
			& \textsc{sg} &\textsc{pl} \\ \midrule
			1	&	\textit{anggon}	& 	\textsc{ex}: \textit{inggon} \\
			& & \textsc{in}: \textit{pin}\\
			2	&	\textit{kain}	&	\textit{kin}\\
			3	&	\textit{main}	&	\textit{muin}\\ \arrayrulecolor{black}
			\lspbottomrule
		\end{tabularx}
}
\end{table}

The phonological form of basic pronouns shows a pattern across number/person. First-person forms end in /n/ (except for the first-person inclusive, which has initial /p/), second-person forms start with /k/ and third person forms start with /m/. All singular pronouns have the vowel /a/. Plural pronouns have /i/ or /u/.

Note that the restricting series with \textit{-(ah)utak} has variable pronunciation with /h/ and /s/, such that the suffix may be [aˈhutak̚] or [aˈsutak̚] (see also §\ref{sec:debucc}). Pronunciation with /h/ is more common than with /s/. /h/ or /a/ and /h/ may also be omitted, resulting in [aˈutak̚] or [ˈutak̚], which are the more common realisations of the suffix. Variable pronunciation is also found for the collective pronouns, where the suffixed part may be [naˈniŋgan], [aˈniŋgan], [niˈniŋgan] or [ˈniŋgan].

Third-person singular \textit{ma} may also refer to inanimates. Third-person plural \textit{mu} does not occur with reference to inanimates.\is{animacy}

\begin{exe}
	\ex \gll padanual=at potman \textbf{mat} parair\\
	pandanus=\textsc{obj} cut \textsc{3sg.obj} cut\\
	\glt `[You] cut the pandanus, split it...' \jambox* {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB8-C}{[narr11\_1:56]}}
	\label{exe:manmat}
\end{exe}

%tested 2019: -autak, -tain and =nininggan can really not occur on nouns.

\subsection{Morphosyntactic properties}
At least the basic pronouns, the \textit{-tain} series of restricting pronouns and possessive pronouns can be marked with the following clitics in different syntactic and pragmatic roles: object \textit{=at} (§\ref{sec:at}), \is{causative}causative \textit{di=} (in give-constructions, §\ref{sec:give}), comitative \textit{=bon} (§\ref{sec:comi}) and similative \textit{=kap} (§\ref{sec:simcase}). Lative \textit{=ka} and locative \textit{=ko} cannot be suffixed to pronouns, having dedicated animate forms instead that can be used with pronouns referring to animates: \textit{=kongga} and \textit{=konggo}, respectively (§\ref{sec:animloclat}). Benefactive marker \textit{=ki} is the only case enclitic that cannot be attached to pronouns (see §\ref{sec:give}). An instance of instrumental \textit{=ki} was elicited, see~(\ref{exe:anggonggi}). The comitative and similative postpositions\is{postposition!on pronouns} are illustrated in~(\ref{exe:proncom}) and~(\ref{exe:pronkap}), respectively.

\begin{exe}
		\ex
	{\gll an \textbf{anggon}=\textbf{ki} kawotman ma \textbf{main}=\textbf{ki} kawotman\\
		\textsc{1sg} \textsc{1sg.poss=ins} cut \textsc{3sg} \textsc{3poss=ins} cut\\
		\glt `I cut with mine, he cuts with his.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_wc19\_86]}}
	}
	\label{exe:anggonggi}
	\ex 
		{\gll mu se yal \textbf{marain=bon}\\
		\textsc{3pl} {\glse} paddle \textsc{3sg.alone=com}\\
		\glt `They paddled with him alone.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDB-C}{[narr28\_11:26]}}
	}
	\label{exe:proncom}
		\ex
	{\gll me sama=i nain \textbf{an=kap}\\
		\textsc{dist} same={\gli} like \textsc{1sg=sim}\\
		\glt `That's the same as me.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC1-0}{[narr19\_12:22]}}
	}
	\label{exe:pronkap}	
\end{exe}

All pronouns in Table~\ref{tab:pronouns} must carry object marker \textit{=at} when they function as the (direct or indirect) object of a clause. (\ref{exe:bandrol}) and~(\ref{exe:naluar}) illustrate different kinds of pronouns in object function.

\begin{exe}
	\ex 
	{\gll an bo bandrol-un=at jie \textbf{mu=at} naunak\\
	\textsc{1sg} go tobacco\_pouch-\textsc{3poss=obj} get \textsc{3sg=obj} show\\
    \glt `I went to get the tobacco's pouch [to] show them.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC5-7}{[narr16\_3:38]}}}
\label{exe:bandrol}
    \ex
    {\gll an \textbf{anggon=at} naluar=teba\\
    \textsc{1sg} \textsc{1sg.poss=obj} slacken=\glteba\\
    \glt `I'm slackening mine.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCD-C}{[conv3\_4:16]}}}
\label{exe:naluar}
\end{exe}

%(other forms to be tested 2020)
%\textbf{Not all combinations of pronominal suffixes and case markers are found, but that is probably just because they are not so common. Found: marainbon, nembarainggap.}

Other nominal morphology that is also carried by pronouns is \textit{=nan} `too', illustrated in~(\ref{exe:pronnan}).

\begin{exe}
	\ex 
	{\gll \textbf{ma=nan} koyan amdir-un=ko\\
		\textsc{3sg}=too plant garden-\textsc{3poss=loc}\\
		\glt `He also plants in the garden.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB0-D}{[stim12\_0:42]}}
	}
	\label{exe:pronnan}
\end{exe}

The enclitic \textit{=saet} `only, exclusively', found also on nouns, was not found on pronouns in the natural spoken corpus, probably because the restricting pronouns take over that function.

\subsection{Clusivity}\is{clusivity}
\label{sec:clus}
In all pronominal paradigms, there is a distinction between inclusive and exclusive forms. Forms containing \textit{in} are first-person exclusive (referring to the speaker and other people, excluding the addressee), while forms containing \textit{pi} are inclusive (referring to the speaker and the addressee, and (an)other person(s)). The dual form \textit{inier} refers to the speaker and another person, and the dual form \textit{pier} refers to the speaker and the addressee. Clusivity is illustrated for the basic pronouns in~(\ref{exe:kimuninbon}) and~(\ref{exe:pibuokbuok}).\footnote{Contra the preliminary analysis of the Kalamang pronominal system in \textcite{visser2016}, clusivity is a well-established distinction in Kalamang.}

\begin{exe}
	\ex
	{\gll ki-mun \textbf{in}=bon ewa∼wa=in\\
		\textsc{2pl-proh} \textsc{1pl.excl=com} speak∼\textsc{prog=proh}\\
		\glt `Don't you guys speak to us!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_31:04]}}}
	\label{exe:kimuninbon}
	\ex
	{\gll \textbf{pi} buokbuok=et eba \textbf{pi} garung=et\\
		\textsc{1pl.incl} chew\_betel={\glet} then \textsc{1pl.incl} talk={\glet}\\
		\glt `We chew betel and then we talk.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_00:10]}}}
	\label{exe:pibuokbuok}
\end{exe}


\subsection{Number}\is{number!pronominal}
Dual number is created by suffixing \textit{-ier} to the basic pronouns. This might be a metathesis\is{metathesis} of \textit{eir} `two'. Dual forms are not obligatory, but are used in the great majority of the cases where a pronoun refers to two referents. When there are more than two referents, the speaker can choose between a basic pronoun with or without a suffixed \is{numeral}numeral. The higher the number of referents, the lower the share of pronouns with a suffixed numeral.\footnote{This observation is impressionistic. It is based on a corpus search for pronouns with a suffixed numeral. A confounding factor is that lower numerals are more common than higher numerals.} (\ref{exe:mier}) shows dual number, and~(\ref{exe:munggansuor}) shows the numeral \textit{kansuor} `four' suffixed to the third-person plural.\is{metathesis}\is{number!pronominal}

\begin{exe}
	\ex
	{\gll \textbf{inier} et-putkaruok koluk\\
		\textsc{1du.ex} \textsc{clf\_an}-thirteen find\\
		\glt `We caught thirteen pieces.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAE-4}{[narr44\_5:17]}}
	}
	\label{exe:mier}
	\ex
	{\gll \textbf{munggansuor} belajar=teba\\
		they\_four study={\glteba}\\
		\glt `They four are studying.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA8-B}{[stim4\_4:05]}}
	}
	\label{exe:munggansuor}
\end{exe}

Although suffixing a numeral to a pronoun is less and less common the higher the number of referents, there are several instances in the corpus where speakers count the number of referents to make sure they refer to them in the correct way (as in~\ref{exe:muraman}, where the number of people in a picture is counted), or check with their addressee whether they are referring to the correct number of people, as in~(\ref{exe:mupuraman}). In~(\ref{exe:muraman}) it is unclear whether \textit{raman=a} should be analysed as a suffix to the pronoun or as an independent numeral, but since the corpus contains examples with `two', `three' and `four' that are suffixed, I take it that `six' is, too.

\begin{exe}
	\ex
	{\gll kon eir karuok kansuor \textbf{mu-raman=a} melelu\\
		one two three four \textsc{3pl}-six=\textsc{foc} sit\\
		\glt `One, two, three, four, they six sit.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE7-5}{[stim42\_12:42]}}
	}
	\label{exe:muraman}
\end{exe}	

\begin{exe}	
	\ex
	\begin{xlist}
		\exi{A:}
		{\gll mu puraman\\
			\textsc{3pl} how\_many\\
			\glt `How many are they?'
		}
		\exi{B:}
		{\gll munggaruok\\
			they\_three\\
			\glt `They [are] three.'
		}
		\exi{A:}
		{\gll munggaruok mat rup=te kajie\\
			they\_three \textsc{3sg.obj} help={\glte} pick\\
			\glt `They three help him pick.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD4-C}{[stim29\_0:49]}}
		}
	\end{xlist}
	\label{exe:mupuraman}
\end{exe}

A common construction with \textit{inier}, which includes the speaker, is \textit{inier N=bon}, literally `we two and X'. Rather than referring to three people, it refers to the speaker and X, as illustrated in~(\ref{exe:inierbon}). This is what \textcite{singer2001} refers to as  an inclusory construction.

\begin{exe}
	\ex
	{\gll \textbf{inier} \textbf{Keca=bon} bo war=kin\\
		\textsc{1du.ex} Keca=\textsc{com} go fish={\glkin}\\
		\glt `Me and Keca wanted to go fishing.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAE-4}{[narr44\_0:05]}}
	}
	\label{exe:inierbon}
\end{exe}

\subsection{`Alone' pronouns}\is{focus!pronominal}\is{pronoun!restricting}
\label{sec:alonepron}
There are two series of `alone' pronouns. The \textit{-tain} series has a restrictive or contrastive focus\is{focus}\is{pronoun!focus} meaning, whereas the \textit{-(a)hutak} series refers to literally no more (or fewer) referents than stated in the pronoun, and is a quantifying pronoun series. The differences are slight, and not every corpus example fits this analysis entirely, but the following examples are representative of the use of these pronouns. Both suffixes can be freely translated as `alone' or `self'.

(\ref{exe:marain}) and~(\ref{exe:andain}) illustrate the \textit{-tain} series with restrictive or contrastive focus\is{pronoun!focus}. In~(\ref{exe:marain}), where people describe a picture from the \textit{Family problems picture task}\is{picture-matching task} \parencite{carroll2009}, there are clearly more referents to choose from, but the speaker wishes to focus on `him alone'. In~(\ref{exe:andain}), from the same recording, the speaker explains that the man in the task, who has come out of jail and has become a better person, improved himself on his own. There were perhaps more candidates for the source of his improvement, but the speaker wishes to convey that the man did it all alone. Strikingly, \textit{-tain} pronouns often co-occur with the focus enclitic \textit{=a}.

\begin{exe}
	\ex
	{\gll \textbf{ma-tain} taikonggo kiun taikonggo\\
		\textsc{3sg}-alone side.\textsc{loc} wife.\textsc{3poss} side.\textsc{loc}\\
		\glt `He (alone/himself) on one side, his wife on the other side.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_16:27]}}
	}
	\label{exe:marain}
	\ex
	{\gll \textbf{an-tain=a} paruo\\
		\textsc{1sg}-alone=\textsc{foc} do\\
		\glt `It was I who did it.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_16:27]}}
	}	
	\label{exe:andain}
\end{exe}

The \textit{-tain} series may also be used in \is{reflexive}reflexive constructions, as in~(\ref{exe:undeirii}). See §\ref{sec:rflx} for a further description.

\begin{exe}
	\ex \gll ma-\textbf{tain} se un-deir=i luk\\
	\textsc{3sg}-alone {\glse} \textsc{refl}-bring={\gli} come\\
	\glt `She came herself.' (Lit. `She brought herself coming.') \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBC-4}{[narr24\_5:33]}}
	\label{exe:undeirii}
\end{exe}	

(\ref{exe:kahutak}) and~(\ref{exe:mahutak}) illustrate the quantifying \textit{-(a)hutak} series\is{pronoun!quantifying}. In~(\ref{exe:kahutak}), one speaker leaves the other during a recording session and orders her to speak further on her own. There are no others around, so the speaker is really only talking about the addressee. In~(\ref{exe:mahutak}), people are discussing a picture with a girl named Ramina on it, and no-one else.

\begin{exe}
	\ex
	{\gll \textbf{ka-hutak} watko ewa=te\\
		\textsc{2sg}-alone \textsc{prox.loc} speak=\textsc{imp}\\
		\glt `You talk on your own here!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_21:23]}}
	}
	\label{exe:kahutak}
	\ex
	{\gll Ramina \textbf{ma-hutak} bara\\
		Ramina \textsc{3sg}-alone descend\\
		\glt `Ramina alone is coming back.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE7-5}{[stim42\_15:46]}}
	}
	\label{exe:mahutak}
\end{exe}

The difference of these last two examples with the \textit{-tain} examples above is that there are no more referents to choose from. A negated \textit{-(a)hutak} example further strengthens this hypothesis. In~(\ref{exe:inahutak}), the speaker explains that he and his travel company were not alone (drinking tea at a funeral), but that the other invited people did so as well. By negating a quantifying pronoun, the speaker indicates that there were more people present than just those referred to with `we'. The \textit{-(a)hutak} pronouns contrast with the \textit{-(V)ninggan} series described in the next section, also illustrated in the example below.

\begin{exe}
	\ex
	{\gll bukan \textbf{in-ahutak} ge {\ob}...{\cb} \textbf{in-ininggan} mindi\\
		not \textsc{1pl.excl}-alone no {} \textsc{1pl.excl}-all like\_that\\
		\glt `Not just us, we all did like that.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD8-4}{[narr1\_00:47]}}
	}
	\label{exe:inahutak}
\end{exe}

The following elicited example shows how the two pronouns are not interchangeable in some contexts. In~(\ref{exe:anahutak}), the form \textit{anahutak} is the only correct one, because there was only one person swimming. \textit{Andain} would have been correct if the speaker intended to communicate that she swam without help, but that would have required a different context.

\begin{exe}
	\ex
	{\gll sontum reidak osep=ko \textbf{an-ahutak/*-tain} jie\\
		person many beach=\textsc{loc} \textsc{1sg}-alone/-alone swim\\
		\glt `There were many people at the beach [but] only I was swimming.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_pro\_10]}}
	}
	\label{exe:anahutak}
\end{exe}

Note also that the \textit{-(a)hutak} forms seem to include the clitic \textit{=tak} `just, only'. Dual forms cannot be suffixed with \textit{-(a)hutak}, only with \textit{-tak}, e.g. \textit{iniertak} means `only us two'. Perhaps these forms are comparable to the \textit{-(a)hutak} forms. The dual forms with \textit{-tak} give the speaker the possibility to specify that there were only two people, no more, no less. Forms like *\textit{inggaruoktak} `only us three' and *\textit{inggansuortak} `only us four' are not found in the corpus.

Dual forms with \textit{-tain} were elicited, but their semantics remain unclear. For the third-person dual, \textit{miertain}, a speaker remarked that it means `they two have'. Dual pronouns with \textit{-tain} can perhaps be analysed as pronoun+\textit{tak}+\textit{kin}, where \textit{-kin} is the associated plural or a possessive form indicating part-whole relations (see §\ref{sec:asspl}).

\subsection{Collective pronouns}\is{pronoun!quantifying}
\label{sec:allprons}
Another series of quantifying pronouns are the collective pronouns. The use of a collective pronoun stresses that all referents referred to with a pronoun were partaking in the action described. It contrasts with the restricting \textit{-(a)hutak} series described above. The suffix, which is only added to plural pronouns, varies between the forms \textit{-naninggan}, \textit{-nininggan} and \textit{-ninggan}. The \textit{-naninggan} form perhaps contains the morpheme \textit{-nan} `too', but the three forms mentioned here seem to be used interchangeably.

\begin{exe}
	\ex 
	{\gll Bobi emun=bon tumtum-un=bon \textbf{in-naninggan=a} se mara metko\\
		Bobi mother.\textsc{3poss=com} children-\textsc{3poss=com} \textsc{1pl.excl}-all=\textsc{foc} {\glse} move\_landwards \textsc{dist.loc}\\
		\glt `We together with Bobi's mother and her children, we all went there.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_7:32]}}
	}
\end{exe}

For the dual forms, the suffix just takes the form \textit{-gan}, resulting in the meaning `we both, you both, they both', as shown in~(\ref{exe:miergan2}). This suffix is also used on \is{numeral}numerals with the meaning `both' or `all'. Note that the morpheme \textit{-gan} also seems to be part of the word \textit{tebonggan} `all'. \textit{Tebon} does not have a meaning on its own.

\begin{exe}
	\ex 
	{\gll \textbf{mier-gan} nakal-un elak-pis\\
		\textsc{3pl.du}-both head-\textsc{3poss} down-side\\
		\glt `They both have their heads down'. \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD6-8}{[stim38\_2:46]}}
	}
	\label{exe:miergan2}
\end{exe}

\subsection{Possessive pronouns}\is{possessive}
\label{sec:possprons}
Possessive pronouns, listed in Table~\ref{tab:pronouns}, are used in two ways: pronominally and adnominally. A pronominal example is given in~(\ref{exe:kainme}). The possessive pronouns refer to the two ends of a fishing net.

\begin{exe}
	\ex 
	{\gll an \textbf{anggon=at=a} naluar \textbf{kain} me ka narorar=i bo tamatko=et eba naluar=te\\
		\textsc{1sg} \textsc{1sg.poss=obj=foc} slacken \textsc{2sg.poss} {\glme} \textsc{2sg} drag={\gli} go where={\glet} then slacken=\textsc{imp}\\
		\glt `I slacken mine, as for yours, you drag it over there, then you slacken!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC9-2}{[conv5\_1:02]}}
	}
\label{exe:kainme}
\end{exe}


Possessive constructions and possessive morphology is further described in Chapter~\ref{ch:poss}.


\subsection{Pronominal address}\is{person address}
\label{sec:pronaddr}
Pronominal address with the second-person singular \textit{ka} and plural \textit{ki} is a common form of address in everyday speech. It is considered very informal, and is used among peers, among spouses and towards children, typically for commands. In a more polite variant, it is used in combination with non-pronominal address. (\ref{exe:kaaddr}) is taken from a recording of two friends discussing fishing gear. The speaker addresses her friend with the second-person singular \textit{ka} to tell her when to speak. (\ref{exe:munka}) is taken from a recording from a boat trip with two distant relatives. The speaker uses a non-pronominal address form first before he uses the pronoun. There are no formal pronominal terms of address. Non-pronominal address is further described in §\ref{sec:usage}.

\begin{exe}
	\ex \gll an ewa=et me \textbf{ka} tok\_tok=ta\\
	\textsc{1sg} speak={\glet} {\glme} \textsc{2sg} not\_yet=\glta\\
	\glt `When I speak you don't [speak] yet.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C75-D}{[stim15\_0:36]}}
\label{exe:kaaddr}
	\ex \gll \textbf{Binkur} \textbf{emun} \textbf{ka} tok per=at di=mu ∅=te\\
	Binkur mother.\textsc{3poss} \textsc{2sg} first water=\textsc{obj} \textsc{caus=3pl} give=\textsc{imp}\\
	\glt `Binkur's mother, you give them water first.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B6C-1}{[conv27\_5:28]}}
	\label{exe:munka}
\end{exe}

\is{pronoun|)}

\section{Non-pronominal person reference and address}\is{person reference}\is{onomastics}
\label{sec:nonpronref}
Kalamang has a broad array of kinship\is{kinship} terms and names that are used to refer to and address people without using a pronoun. Every Kalamang community member can be referred to and addressed with more than one term or name. Formal terms include kinship terms and teknonyms. Given names are very informal or even taboo. Nicknames are very common, and their degree of formality is in between formal and informal. First, I list kinship terms of reference in §\ref{sec:kinterms}, and then names, nicknames and teknonyms in §\ref{sec:names}. The terms of address and their degrees of formality are discussed in §\ref{sec:usage}. Table~\ref{tab:address} gives an overview of the forms treated in this section.

\begin{table}
	\caption[Terms of non-pronominal reference and address]{Terms of non-pronominal reference and address, split between more and less formal forms. SS = same-sex, OS = opposite-sex, n.a. = not applicable. A dash means not possible/taboo.}
	\label{tab:address}
	\small
		\begin{tabularx}{\textwidth}{l l Q p{2cm}}
			\lsptoprule
			&reference& address,\newline formal &address,\newline informal\\
			\midrule
			SS sibling & \textit{kia-} & teknonym & name\\
			OS sibling & \textit{dun-} & teknonym & name\\
			cross-cousin & \textit{korap} & teknonym & name\\
			SS parallel cousin & \textit{kia-} & teknonym & name\\
			OS parallel cousin & \textit{dudan} & teknonym & name\\
			father & \textit{esa} & \textit{esa}\newline teknonym & \textendash \\
			mother & \textit{ema} & \textit{ema}\newline teknonym & \textendash\\
			father's brother & \textit{esa temun/caun} & \textit{esa (temun/caun)}\newline teknonym & \textit{temun/caun}\\
			father's sister & \textit{wowa temun/caun} & \textit{wowa (temun/caun)}\newline teknonym & \textit{temun/caun}\\
			mother's brother & \textit{esa temun/caun} & \textit{esa (temun/caun)}\newline teknonym & \textit{temun/caun}\\
			mother's sister & \textit{wowa temun/caun} & \textit{wowa (temun/caun)}\newline teknonym & \textit{temun/caun}\\
			great-grandfather & \textit{tatanus} & n.a. & n.a.\\
			great-grandmother & \textit{ninanus} & n.a. & n.a.\\
			grandfather & \textit{esnem/tata} & \textit{esnem/tata}\newline teknonym & \textendash\\
			grandmother & \textit{emnem/nina} & \textit{emnem/nina}\newline teknonym & \textendash \\
			grandparent & \textit{tara-} & as grandfather/grandmother & \textendash\\
			grandchild & \textit{tara-} & teknonym & name\\
			son & \textit{tumun (canam)} & teknonym & name\\
			daughter & \textit{tumun (pas)} & teknonym & name\\
			nephew & \textit{tumun (canam)} & teknonym & name\\
			niece & \textit{tumun (pas)} & teknonym & name\\
			husband & \textit{nam-} & teknonym & name\\
			wife & \textit{kia-, kiar-, kie-} & teknonym & name\\
			parent-in-law & \textit{ketan} & \textit{esa, ema}\newline teknonym & \textendash \\
			child-in-law& \textit{ketan} & teknonym & name\\	
			sibling-in-law & \textit{dauk} & teknonym & name?\\		
			non-kin & teknonym & teknonym & name\\
			\lspbottomrule
		\end{tabularx}
\end{table}

In addition to the specifications \textit{temun} `elder' and \textit{caun} `younger', both terms of reference and terms of address for aunts and uncles can be specified with \textit{raor} `middle'.

\subsection{Kinship terms}\is{kinship}
\label{sec:kinterms}
Kinship terms form a subclass of nouns, some of which are inalienable\is{alienability} (\textit{kiar-/kie-} `wife', \textit{nam-} `husband', \textit{kia-} `same-sex sibling', \textit{dun-} `opposite-sex sibling' and \textit{tara-} `grandparent; grandchild'), and many of which have a plural\is{plural!nominal} form with \textit{-mur} (§\ref{sec:kinnouns}). %Kalamang has an Iroquois kinship system in the classification made by \textcite{morgan1871}.

Kinship terms are presented in the following groups: siblings; cousins; parents, aunts and uncles; grandparents and grandchildren; sons and daughters, nieces and nephews (consanguines); and inlaws (affines). For each group, the terms are presented in a table. An overview of all terms of reference, alongside terms of address, is given in Table~\ref{tab:address} in §\ref{sec:usage}.

\subsubsection{Kinship terms: consanguines}
Consanguines are those family members with whom one shares an ancestor.

For siblings there are different terms used depending on the sex of the referee, with the inalienable\is{alienability} terms \textit{kia-} for `same-sex sibling' and \textit{dun-} for `opposite-sex sibling'. Terms like \textit{canam/pas} `male/female' and \textit{temun/cicaun} `big/small' (here: elder/younger) can be added as terms of reference if one wants to be precise, but these are not generally used. The forms are presented in Table~\ref{tab:siblings}.

\begin{table}[ht]
	\caption[Kinship terms: siblings]{Kinship terms: siblings}
	\label{tab:siblings}
	
		\begin{tabularx}{.66\textwidth}{XXl}
			\lsptoprule
			& same-sex & opposite-sex\\ \midrule
			ego & \textit{kia-} & \textit{dun-}\\
			\lspbottomrule 
		\end{tabularx}
	
\end{table}

Terms for cousins distinguish between parallel and cross-cousins (\textit{korap}). A parallel cousin is a cousin from a parent's same-sex sibling, while a cross-cousin is from a parent's opposite-sex sibling. Within parallel cousins, there is a distinction between same-sex (\textit{kia-}, the same term used for same-sex siblings) and opposite-sex cousins (\textit{dudan}, cf. \textit{dun-} for opposite-sex siblings). Table~\ref{tab:cousins} gives an overview of the terms.

\begin{table}[ht]
	\caption[Kinship terms: cousins]{Kinship terms: cousins}
	\label{tab:cousins}
	
		\begin{tabularx}{.66\textwidth}{XXll}
			\lsptoprule
			& \multicolumn{2}{c}{parallel}  & cross\\
			\cmidrule{2-3}
			& same-sex & opposite-sex & \\ \midrule
			ego & \textit{kia-} & \textit{dudan} & \textit{korap}\\
			\lspbottomrule 
		\end{tabularx}
	
\end{table}

Father and mother are called \textit{esa} and \textit{ema}, respectively. A distinction is made between uncles and aunts who are of the same sex as ego's parents and those that are of the opposite sex of ego's parents. Father's brothers and mother's sisters are  \textit{esa} and \textit{ema}, respectively. Father's sisters are \textit{wowa}, \textit{ulan} or \textit{u} (in order of commonness) and mother's brothers are \textit{mama}. Aunts and uncles who are older than ego's parent get the optional addition \textit{temun} `elder'. Aunts and uncles younger than ego's parent get the optional addition \textit{cicaun} `younger', usually shortened to \textit{caun}.\footnote{This is parallel to the local Malay, where \textit{mama tua} and \textit{bapak tua} are father or mother's older sister and brother, respectively. A parent's younger sister can be called \textit{(mama) ci}, but I am not aware of a corresponding term for men in use on Karas.} For example, ego's mother's elder sister (MeZ)\footnote{Using the established abbreviations for kinship terms: e = elder,	y = younger, F = father, M = mother,	B = brother, Z = sister, S = son, D = daughter, H = husband, W = wife. I.e. FeZ means `father's elder sister'.} is called \textit{ema temun}. Table~\ref{tab:parents} also shows that the terms are the same for male and female speakers.

\begin{table}[ht]
	\caption{Kinship terms: parents, aunts and uncles}
	\label{tab:parents}
	
		\begin{tabularx}{.8\textwidth}{XXXXl}
			\lsptoprule
			& F/FB & M/MZ & MB & FZ \\ \midrule
			ego & \textit{esa} & \textit{ema} & \textit{mama} & \textit{wowa}\\
			\lspbottomrule 
		\end{tabularx}
	
\end{table}

For grandparents there are two sets of terms that seem to be used interchangeably. \textit{Esnem} `grandfather' and \textit{emnem} `grandmother' are indigenous terms, where\-as \textit{tata} and \textit{nina} seem to be loans from an East Malukan language (see §\ref{sec:kinloan} below). A general term for grandparent regardless of sex is inalienable \textit{tara-}. This same term is applied to grandchildren,\footnote{In other words, lineal kin of both two generations up and two generations down from ego is referred to as \textit{tara-}} with the optional addition of \textit{canam} `male' or \textit{pas} `female'. All terms are shown in Table~\ref{tab:grand}. Great-grandparents are called \textit{tatanus} `great-grandfather' and \textit{ninanus} `great-grandmother'.

\begin{table}[ht]
	\caption[Kinship terms: grandparents and grandchildren]{Kinship terms: grandparents and grandchildren}
	\label{tab:grand}
		\begin{tabularx}{.8\textwidth}{Xllll}
			\lsptoprule
			& FF/MF & FM/MM & SS/DS & SD/DD \\ \midrule
			ego & \textit{tata/esnem} & \textit{nina/emnem} & \textit{tara- canam} & \textit{tara- pas}\\
			\lspbottomrule 
		\end{tabularx}
	
\end{table}

Sons, daughters, nephews and nieces are all called \textit{tumun} `child', with the optional addition of \textit{canam} for the boys and \textit{pas} for the girls. Table~\ref{tab:sons} shows the relatives that are covered by these terms.

\begin{table}[ht]
	\caption{Kinship terms: sons, daughters, nephews and nieces}
	\label{tab:sons}
	
		\begin{tabularx}{.8\textwidth}{X l l }
			\lsptoprule
			& S/eBS/yBS/eZS/yZS & D/eBD/yBD/eZD/yZD \\ \midrule
			ego & \textit{tumun (canam)} & \textit{tumun (pas)}\\
			\lspbottomrule
		\end{tabularx}
	
\end{table}

The middle child or grandchild, as well as the middle aunt or uncle, can be designated as \textit{tumun/tara-/wowa/mama raor}, with \textit{raor} meaning `middle'.

\subsubsection{Kinship terms: affines}
Affines are relatives by marriage. The terms for `husband' and `wife' are inalienable. `Husband' has the same root for first-, second-, and third-person singular inflection: \textit{nam-}. For `wife', first and second-person have the root \textit{kia-} or \textit{kiar-}, whereas the third-person has the root \textit{kie-}. The inflected forms are given in Table~\ref{tab:married}. Note that the term \textit{kian} `my wife' is homophonous with \textit{kian} `my same-sex sibling'.

\begin{table}[ht]
	\caption{Kinship terms: inflected terms for wives and husbands}
	\label{tab:married}
	
		\begin{tabularx}{.8\textwidth}{X X l }
			\lsptoprule
			& wife & husband\\\midrule
			\textsc{1sg} & \textit{kian, kiaran} & \textit{naman}\\
			\textsc{2sg} & \textit{kiar, kiarca} & \textit{namca}\\
			\textsc{3sg} & \textit{kieun} & \textit{namun}\\
			\lspbottomrule
		\end{tabularx}
	
\end{table}

The specification \textit{caun} `small' can be used for a second (or third, or fourth) wife, whether polygamous\footnote{Polygamy is legal in Indonesia but not currently practiced on Karas.} or after the first wife's or husband's death. A bachelor is called \textit{toari}, although this is also a more general term for `youngster'.

In-laws, both parents-in-law and children-in-law, are termed \textit{ketan}. Siblings-in-law are referred to as \textit{dauk}.


\subsubsection{Kin term borrowings}
\label{sec:kinloan}
A number of kinship terms are borrowings from Austronesian\is{Austronesian!loan} languages, reflecting contact with other Austronesian groups. \textit{Wowa} is likely a borrowing from the neighbouring Austronesian language Uruangnirin\il{Uruangnirin} \parencite{uruangnirinLR}, where it also means `father's sister'. It is also found in East Central Maluku languages such as \ili{Geser-Gorom} \parencite{goromLR}, \ili{Watubela} \parencite{collins1986} and \ili{Masiwang} \parencite{cocq1901}. It has the same form (\textit{wowa}) in all these languages.

The terms \textit{tata} `grandfather' and \textit{nina} `grandmother' are also likely borrowed from Uruangnirin\il{Uruangnirin} \textit{tatan} and \textit{nina}, respectively. Again, East Central Malaku languages such as Geser-Gorom\il{Geser-Gorom} have similar terminology (\textit{tatanusi} `ancestor', \citealt{chlenova2010}, \textit{tata ena} `grandfather' and \textit{tata nina} `grandmother', \citealt{uruangnirinLR}. Note also the polite address form for female strangers in Kalamang, \textit{tatanina}.

The terms for great-grandparents are likely borrowings from Geser-Gorom, with Kalamang \textit{tatanus} `great-grandfather' and \textit{ninanus} `great-grandmother'.\is{Moluccas}

\subsection{Names}\is{names!personal names|(}
\label{sec:names}
Besides kinship terms, personal proper names are a common strategy for non-pronominal reference and address. Because it is customary to avoid first names, most people are known by a variety of names. Names can be divided into three categories: given names, nicknames and teknonyms.\is{names!teknonymy}\is{names!nicknames}

Most Kalamang speakers have an Arabic name as their given name. Some non-Arabic names of community members are Erna, Sabtu (Malay `Saturday') and Bini (Malay `wife'). I am not aware of any names that are regarded as Kalamang proper names. 

Nicknames can be of many kinds, and can change throughout a person's lifetime. Nicknames may be regular first names, such as Manto, which was given to a Kalamang speaker by visiting fishermen. Nicknames may also originate in other entities, such as Mester, the name of the boat on which that person was born. One boy is called Kalamang, after the name of his people. Nicknames also commonly refer to someone's place of origin, especially if they are the first or only person from that place: \textit{Kei-ca} stands for Kei-man \textit{Walaka-ca} for Gorom-man.

As soon as someone becomes a parent, that person can be referred to and addressed with their teknonym\is{names!teknonymy} (the name of their first child) followed by \textit{esun} `father.\textsc{3poss}' or \textit{emun} `mother.\textsc{3poss}'. This is exemplified with \textit{Nyong esun} `Nyong's father' in~(\ref{exe:nyong}). 

\begin{exe}
	\ex 
	{\gll \textbf{Nyong} \textbf{esun}=a marua yuwa\\
		Nyong father.\textsc{3sg=foc} move\_seawards \textsc{prox}\\
		\glt `Nyong's father comes towards sea here.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCB-5}{[conv1\_3:49]}}
	}
	\label{exe:nyong}
\end{exe}

When someone becomes a grandparent, the teknonym is usually replaced with the name of the first grandchild, although the name of the first child may also still be used. Someone can thus be known as `X's grandmother' and `Y's mother' at the same time. For married people without children, the name of the spouse may be used: `X's wife', or `Y's husband'. If the spouse is not Kalamang, his or her name is generally avoided, and the person is referred to with a nickname. \is{names!personal names|)}

\subsection{Non-pronominal address}\is{person address}
\label{sec:usage}
In everyday situations, pronouns are very common terms of address (§\ref{sec:pronaddr}), but non-pronominal terms of address are also very conventionalised. All non-pron\-om\-i\-nal terms were given in Table~\ref{tab:address}. Terms of address can be put on a cline of politeness (here, for the sake of ease, encompassing also formality and respectfulness). Towards the polite\is{polite speech} side there are kinship terms and teknonyms, and names are towards the less polite side. Nicknames cover the middle ground.\is{formal speech}\is{informal speech} Vocatives\is{vocative} are briefly described in §\ref{sec:wcyesno} and §\ref{sec:discinterj}.

Kinship terms are used as terms of address for parents, parents' siblings, grandparents, parents-in-law and strangers or foreigners. Such terms of address are deemed polite. Teknonyms and names of spouses are equally polite ways of addressing someone as kinship terms. Moreover, the use of teknonyms and names of spouses as a term of address is available for all kin and non-kin relations. Names are the common way of addressing kin and non-kin for and amongst children until they are married. After that, names may only be used in (very) informal settings, and between people of the same generation, or when addressing someone from a younger generation than yourself. It is deemed very disrespectful, or even taboo, to address or even refer to a parent, grandparent or parent-in-law by their first name. Even between friends it is customary to use teknonyms rather than first names. Teknonyms without the addition of the word `(grand)parent' may be used as a kind of collective term of address for a group of people, usually sharing a house. This can be used, for example, when someone standing outside the house calls to see if someone's there, or to refer to that group of people (in that case with the addition of \textit{mu} \textsc{3pl}, as in~(\ref{exe:mayor})\is{associative plural}.)

\begin{exe}
	\ex 
	{\gll {\ob}\textbf{Mayor} \textbf{mu}{\cb} amdir-un=ka=ta marua\\
		Mayor \textsc{3pl} garden-\textsc{3poss=lat}={\glta} move\_seawards\\
		\glt `Mayor and associates go to their garden moving towards the sea.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCE-D}{[conv4\_7:16]}}
	}
	\label{exe:mayor}
\end{exe}

Nicknames and titles occupy the middle ground in terms of politeness. If a nickname is honorary, such as Mayor, this may be the main term of address for that person. People with an administrative function may be addressed according to their title (in Malay): \textit{Pak Sekretaris} `Mister Secretary' and \textit{Ibu Guru} `Mistress Teacher'. Disabled people may also be first and foremost known by their nicknames (such as a deaf-mute person referred to as \textit{Toki}, perhaps from \textit{kelkam toktok} `deaf'). Note, however, that if a person is known to be very informal, they may be primarily known by their first name, such as the often-joking father of three children called Malik. Self-reference is usually obtained with a teknonym. If people utter their own given name, they usually do so with a quieter voice.

Strangers or foreigners may be both referred to and addressed with polite Kalamang kinship terms, preferably \textit{esnem} or \textit{esa} for a man, and \textit{emnem} or \textit{ema} for a women. However, terms of address from other (Indonesian) languages are also commonly used. A visiting man from Java is referred to and addressed as Mas; a Javanese woman is Mba (after the Javanese terms of address). A foreigner may be referred to as `mister' (pronounced \textit{mester}) or `mistress/miss' (\textit{misis/mis}). If necessary, a specification is given after the title: Mas Karamba, for example, is used for the Javanese men who work in the live-fish storage place called \textit{karamba} in Indonesian. Indonesian Bapak/Pak and Ibu can be applied to any stranger. The borrowing \textit{tatanina} may also be used for foreign women, and was a popular term of reference and address for the researcher, as well as \textit{mis(is)}.

Kalamang society practices kinship-based fostering, which means that children may be raised by a family member, usually one who does not have children him/herself.\footnote{\textit{Anak piara }in local Malay. The parents of such a child are typically alive and capable of taking care of children, and are in contact with the child they have fostered out. This may be a practice that takes care of individuals or families that do not or cannot have children themselves, while at the same time lifting the burden from families with many children.} These children have and use the same terms of reference and address as biological children.

\is{person reference|)}
