\chapter{Possessive and associative constructions}\is{possessive|(}
\label{ch:poss}
Possessive constructions express the relation between the possessor and the possessed. This chapter deals with the ways these relations are expressed in Kalamang. It also includes a discussion of associative relations, which are expressed with the enclitic \textit{=kin}.

\section{Overview}
Kalamang has two series of adnominal markers of possession: one with suffixes and one with pronouns. Possessive pronouns can be used pronominally or (much more rarely) follow the possessed noun they modify. In possessive constructions where the possessor is a noun rather than a pronoun, the order is reversed (§\ref{sec:possphr}). Possessive suffixes and pronouns are given in Table~\ref{tab:posspron}, together with the basic pronouns for comparison.


\begin{table}
	\caption{Basic pronouns, possessive suffixes and possessive pronouns}
	\label{tab:posspron}
	
		\begin{tabular}{llll}
			\lsptoprule 
			& pronoun & possessive suffix & possessive pronoun\\
			\midrule
			\textsc{1sg} & \textit{an} & \textit{-an} & \textit{anggon}\\
			\textsc{2sg} & \textit{ka} & \textit{-ca} & \textit{kain}\\
			\textsc{3sg} & \textit{ma} & \textit{-un} & \textit{main}\\
			\textsc{1pl.excl} & \textit{in} & \textit{-un} & \textit{inggon}\\
			\textsc{1pl.incl} & \textit{pi} & \textit{-pe} & \textit{pin}\\
			\textsc{2pl} & \textit{ki} & \textit{-ce} & \textit{kin}\\
			\textsc{3pl} & \textit{mu} & \textit{-un} & \textit{muin}\\\lspbottomrule 		
		\end{tabular}
	
\end{table} 

Ordinary possessive constructions, the most common possessive construction type, are made by attaching a possessive suffix to the possessed (the head noun in the NP), as illustrated in~(\ref{exe:posssuff}). 

\begin{exe}
	\ex
	{\gll an bo lembaga nerun {\ob}tumun-\textbf{an}{\cb}\textsubscript{\upshape NP} se bo temun\\
		\textsc{1sg} go prison inside child-\textsc{1sg.poss} {\glse} go big\\
		\glt `I went to prison, and my child has grown big.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_29:09]}}
	}
	\label{exe:posssuff}
\end{exe}
	
Three other, much less common constructions are available. First, possessive pronouns can be used as modifiers of the possessed head noun in the NP to create a possessive construction. The possessive pronouns occur adnominally only rarely, and mostly following the kinship\is{kinship} terms \textit{esa} `father' and \textit{ema} `mother' as in~(\ref{exe:posspron}). Second, a combination of a possessive suffix on the possessed noun and a possessive pronoun following the possessed noun also creates a possessive construction, as in~(\ref{exe:possdouble}). Third, a combination of pronoun, possessed noun and possessive suffix is found, as illustrated in~(\ref{exe:pusunrasa}).

\begin{exe}
		\ex
	{\gll ma sem=nin {\ob}ema \textbf{anggon}{\cb}\textsubscript{\upshape NP} me sem=nin\\
	\textsc{3sg} afraid=\textsc{neg} mother \textsc{1sg.poss} {\glme} afraid=\textsc{neg}\\
	\glt `She wasn't afraid, my mother wasn't afraid.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_16:35]}}
	}	
	\label{exe:posspron}
		\ex 
	{\gll wa me {\ob}taman-\textbf{un} \textbf{main}{\cb}\textsubscript{\upshape NP}=a\\
	\textsc{prox} {\glme} friend\textsc{.mly-3poss} \textsc{3poss=foc}\\
	\glt `This is his friend.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA9-9}{[stim6\_6:38]}}
	}
	\label{exe:possdouble}
	\ex \gll {\ob}ma pus-\textbf{un}{\cb}\textsubscript{\upshape NP} rasa\\
	\textsc{3sg} flower-\textsc{3poss} good\\
	\glt `It will have good flowers.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD0-8}{[narr13\_2:10]}}
	\label{exe:pusunrasa}
\end{exe}

The four different constructions are described in §\ref{sec:posssuf} to §\ref{sec:possnew}.

The freestanding possessive pronouns can be used both as NP heads and as modifiers, whereas the suffixes always have to be attached to the NP head (the possessed noun). Adnominal use of the freestanding possessive pronoun was illustrated in~(\ref{exe:posspron}) and~(\ref{exe:possdouble}) above. A pronominal example is given in~(\ref{exe:anggonawa}).
	
\begin{exe}
	\ex
	{\gll \textbf{anggon}=a wa lolok tak-karuok\\
		\textsc{1sg.poss}=\textsc{foc} \textsc{prox} leaf \textsc{clf\_leaf}-three \\
		\glt `This is mine. It has three leaves.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD6-8}{[stim38\_0:36]}}
	} 
\label{exe:anggonawa}
\end{exe}

Kalamang has one possessive-like construction to express associative relations, for which clitic \textit{=kin} is used. This is described in §\ref{sec:asspl}.

Possessives and other NP modifiers, and their behaviour in relation to each other, are described in §\ref{sec:nomphr}.

\section{Possessive constructions with nominal possessors}\is{possessor-possessum order}
\label{sec:possphr}
In possessive constructions with nominal possessors, the NP head (the possessed noun) is preceded by the possessor. It is ungrammatical to swap the order. The third-person possessive marker is suffixed to the NP head, as shown in~(\ref{exe:malikkeweun}) and~(\ref{exe:emadidirasun}).

\begin{exe}
\ex
{\gll Malik kewe-un\\
	Malik house-\textsc{3poss}\\
	\glt `Malik's house' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C9A-E}{[narr37\_1:51]}}
}
\label{exe:malikkeweun}
\end{exe}

\begin{exe}
	\ex
	{\gll ema didiras-un\\
		mother kitchen-\textsc{3poss}\\
		\glt `Mother's kitchen' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_gen\_3]}}
	}
	\label{exe:emadidirasun}
\end{exe}

Note also the contrasting meaning between~(\ref{exe:childdog}), where \textit{bal} `dog' is the NP head, and~(\ref{exe:dogchild}), where \textit{tumun} `child' is the head.

\begin{exe}
	\ex
	\begin{xlist}
		\ex
		{\gll tumun bal-un\\
			child dog-\textsc{3poss}\\
			\glt `The child's dog.'
		}
		\label{exe:childdog}
		\ex
		{\gll bal tumun-un\\
			dog child-\textsc{3poss}\\
			\glt `The dog's child (puppy).' \jambox*{[elic]}
		}
		\label{exe:dogchild}
	\end{xlist}
\end{exe}

%test 2020 possessor + dem e.g. this mother's kitchen. this child's dog. this dog's puppy.

NPs may be embedded in prenominal possessor position to modify the NP head. In~(\ref{exe:temankeweun}), the head \textit{sanong} `roof' is modified by \textit{kewe} `house', which is in turn modified by `friend'.

\begin{exe}
	\ex {\gll teman-an kewe-un sanong-un\\
		friend-\textsc{1sg.poss} house-\textsc{3poss} roof-\textsc{3poss}\\
		\glt `My friend's house's roof.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_gen\_20]}}
	}
	\label{exe:temankeweun}
\end{exe}

Although both third-person possession and nominalised attributes are marked with the suffix \textit{-un} (see §\ref{sec:nomder}), there is no ambiguity between a possessive phrase with an attribute, as in~(\ref{exe:possnom}a), and a phrase with a nominalised attribute, as in~(\ref{exe:possnom}b).

\begin{exe}
	\ex 
	\begin{xlist}
		\ex
		{\gll kewe-un temun\\
			house-\textsc{3poss} big\\
			\glt `his big house (or: his house is big)' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_adj\_30]}}
		}
		\ex
		{\gll kewe temun-un\\
			house big-\textsc{nmlz}\\
			\glt `the size of the house' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB2-1}{[narr6\_3:07]}}
		}
	\end{xlist}
	\label{exe:possnom}
\end{exe}

\section{Basic possessive constructions}
\label{sec:posssuf}
The most common possessive construction is created by means of the possessive suffix. I refer to this as the basic possessive construction. This construction can be used for both animate and inanimate possessors and possessed. See~(\ref{exe:tumunan}) with \textit{tumun} `child' and~(\ref{exe:dowiwe}) with \textit{dowi} `seed' as the possessed, respectively.

\begin{exe}
		\ex
		{\gll an bo lembaga nerun tumun-\textbf{an} se bo temun\\
			\textsc{1sg} go prison inside child-\textsc{1sg.poss} {\glse} go big\\
			\glt `I went to prison, and my child has grown big.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_29:09]}}
		}
\label{exe:tumunan}
		\ex
		{\gll nain dowi-\textbf{pe} opa me\\
			like seed-\textsc{1pl.poss} \textsc{ana} {\glme}\\
			\glt `Like those seeds of ours.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD0-8}{[narr13\_0:42]}}
		}
\label{exe:dowiwe}
\end{exe}

The third-person possessive suffix, in its singular reading, is used for a variety of functions with NPs that contain both possessor and possessed including teknonyms\is{names!teknonymy}, part-whole relations and place names. These are described in the following paragraphs.

Kalamang has five nouns that are morphophonologically inalienable\is{alienability} (§\ref{sec:inal}). These are \textit{kiar-/kie-} `wife', \textit{nam-} `husband', \textit{kia-} `same-sex sibling', \textit{dun-} `opposite-sex sibling' and \textit{tara-} `grandparent; grandchild'. These five roots cannot occur without a possessive suffix. Alienability does not, however, play a further role in the choice of possessive construction. For a further discussion, see §\ref{sec:posspron}.\is{alienability}

There are a number of conventionalised constructions that can or must be made by means of the third-person possessive suffix \textit{-un}. First, teknonyms, a common term of address, are made with this possessive suffix. Teknonyms\is{names!teknonymy} are formed with the name of a firstborn and follow the template [name] [kinship term]-[\textit{un}], where the kinship term can be \textit{esa} `father', \textit{ema} `mother' or \textit{tara-} `grandparent'. Consider~(\ref{exe:nyongemun}) and~(\ref{exe:diantaraun}). Note that \textit{esa} and \textit{ema} are \textit{esun} and \textit{emun}, respectively, when inflected with the third-person possessive suffix \textit{-un}.

\begin{exe}
	\ex 
	{\gll kan \textbf{Nyong} \textbf{emun} tok bo=ta opa me {\ob}...{\cb} \textbf{Nyong} \textbf{esun} tok bot=nin\\
		\textsc{int.mly} Nyong mother.\textsc{3poss} first go={\glta} {\glopa} {\glme} [...] Nyong father.\textsc{3poss} yet go=\textsc{neg}\\
		\glt `Right, Nyong's mother went first, Nyong's father didn't go yet.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_7:16]}}
	}
	\label{exe:nyongemun}
	\ex 
	{\gll \textbf{Dian} \textbf{tara-un=a} Pakpak=a bot=kin\\
		Dian grandparent-\textsc{3poss=foc} Fakfak=\textsc{foc} go=\textsc{vol}\\
		\glt `Dian's grandfather wants to go to Fakfak.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD7-2}{[narr3\_1:13]}}
	}
	\label{exe:diantaraun}
\end{exe}

Second, part-whole relations may be made with \textit{-un}, illustrated in~(\ref{exe:partwholes}). The suffix \textit{-un} attaches to the base.

\begin{exe}
	\ex 
	\begin{xlist}
		\ex
		{\gll kerar nar-un\\
			turtle egg-\textsc{3poss}\\
			\glt `turtle egg' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAE-4}{[narr44\_25:08]}}
		}
		\ex
		{\gll rumrum ol-un\\
			plant\_species leaf-\textsc{3poss}\\
			\glt `rumrum leaf' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA2-F}{[conv11\_1:32]}}
		}
		\ex
		{\gll polkayak nak-un\\
			papaya fruit-\textsc{3poss}\\
			\glt `papaya fruit' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA5-0}{[conv15\_1:17]}}
		}
		\ex
		{\gll et bol-un\\
			canoe rim-\textsc{3poss}\\
			\glt `rim of canoe' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDD-5}{[narr42\_4:23]}}
		}	
		
	\end{xlist}
	\label{exe:partwholes}
\end{exe}

In part-whole relations, it is common to omit the first noun if the referent is clearly identified from the context.

Third, geographical names\is{names!place names} may make use of a construction with third-person possessive \textit{-un}. On the island where Kalamang is spoken, each coastal area around a bay has a name. Several landscape features are associated with this name. Thus, the area of Tat includes \textit{Tat Os} `Tat Beach' and \textit{Tat Karimun} `Cape Tat'. Two landscape features are inflected with \textit{-un} when they follow a geographical name: \textit{wilak} `sea', given in~(\ref{exe:wilaun}) and \textit{kelek} `mountain; inland', illustrated in~(\ref{exe:keleun}).

\begin{exe}
	\ex 
	{\gll Sek wilak-\textbf{un} metko tebolsuban\\
		Sek sea-\textsc{3poss} \textsc{dist.loc} fish\_at\_reef\_edge\\
		\glt `[She] went fishing at Sek's sea.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9C-A}{[narr9\_13:46]}}
	}
\label{exe:wilaun}
	\ex
	{\gll ma Sabaor=ka kasawari=at sarie Sabaor kelek-\textbf{un}=ka\\
		\textsc{3sg} Sabaor=\textsc{lat} cassowary=\textsc{obj} chase Sabaor inland-\textsc{3poss=lat}\\
		\glt `He chased the cassowary from Sabaor, from Sabaor inland.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B62-6}{[narr20\_0:44]}}
	}
\label{exe:keleun}
\end{exe}

The landscape feature \textit{karimun} `cape' may contain the morpheme \textit{-un}, but there is no evidence for or against this at the moment. There are no recorded occurrences of \textit{karimun-un}.

\section{Possessive constructions with a freestanding possessive pronoun}
\label{sec:posspron}
The construction with a noun followed by a freestanding possessive pronoun is relatively rare, and is mainly used for \textit{esa} `father' and \textit{ema} `mother'. These two nouns account for around 90\% of this construction type.

The corpus contains ten different nouns that are followed by freestanding possessive pronouns. Six of these nouns are also found with a possessive suffix, which means they do not require a possessive pronoun construction.\footnote{These are \textit{pendapat} `opinion' (Malay loan), \textit{toman} `bag', \textit{hak} `right' (Malay loan), \textit{kewe} `house', \textit{asal} `origin' (Malay loan) and \textit{guru} `teacher' (Malay loan). A reason that many of these are Malay loans might be that possessive constructions in Papuan Malay are made with a periphrastic construction (possessor + `have' + possessum), which might trigger the use of a possessive pronoun instead of a suffix in Kalamang. In any case, the number of occurrences of these words is so low in comparison to those of \textit{esa} `father' and \textit{ema} `mother' that we will not consider them any further.} \textit{Wowa} `aunt' and \textit{tata} `grandfather' are not found with a possessive suffix, but are uncommon in the corpus. \textit{Esa} `father' and \textit{ema} `mother' are only found with a third-person possessive suffix, and only in teknonyms, as in~(\ref{exe:nyongemun}) and~(\ref{exe:diantaraun}) above (see also §\ref{sec:names}).

Examples of \textit{esa} `father' and \textit{ema} `mother' followed by a possessive pronoun are given in~(\ref{exe:esainggon}) and~(\ref{exe:emakain}), respectively.

\begin{exe}
	\ex 
	{\gll esa \textbf{inggon} to esa \textbf{anggon} ma don-pat\\
	father \textsc{1pl.excl.poss} right father \textsc{1sg.poss} \textsc{3sg} clothing-sew\\
	\glt `Our father, right, my father, he sewed clothing.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_19:37]}}
}
\label{exe:esainggon}
\ex 
	{\gll o in ema \textbf{kain}=at konat=nin\\
	\textsc{emph} \textsc{1pl.excl} mother \textsc{2sg.poss=obj} see=\textsc{neg}\\
	\glt `O, we didn't see your mother.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA7-D}{[narr21\_4:21]}}
}
\label{exe:emakain}
\end{exe}

When these nouns are not used as teknonyms, it is ungrammatical to inflect them with a possessive suffix. 

At this point, it is unclear whether other kinship\is{kinship} terms also require a possessive pronoun.\footnote{There is no restriction against a possessive suffix on a noun ending in /a/, cf. \textit{nika-an} `my fishing line'.} Besides three occurrences of \textit{tata} `grandfather' and one of \textit{wowa} `aunt', there are no possessive constructions with other kinship terms. The only exception to this is the five inalienable\is{alienability} terms mentioned in §\ref{sec:posssuf}, which must carry a possessive suffix, as well as \textit{tumun} `child', which may carry a possessive suffix, as was illustrated in~(\ref{exe:tumunan}) above. What is clear, however, is that words from the same semantic class (kinship terms) may behave differently with respect to possessive constructions. This also shows that alienability of the possessed does not play a role in the choice of a possessive construction, as is common elsewhere in East Indonesia \parencite[see e.g.][]{klamer2008}. Note that kinship terms are also treated differently from other nouns in possessive constructions in other languages, notably Italian and Scandinavian \parencite{delsing2002}.

The nouns \textit{sanggoup} `branch', \textit{bol} `mouth', \textit{kanggirar} `face' and \textit{kor} `leg, foot' were tested for preference in possessive construction. Although both the basic possessive construction with a suffix and the possessive pronoun construction were accepted for all forms (when presented with both variants by the linguist), speakers have a clear preference for the basic construction.

%Words found in corpus 08/2019 with a freestanding poss are the following: esa, ema, pendapat, tata, wowa, toman, muap-ca, hak, ewa-un, rampa-un, kewe, harga-un, taman-un, gambar-un, adat-un, neba-un, emumur, asal, guru, et-un and kegiatan-un.

\section{Double possessive marking}
\label{sec:posscombi}
A noun marked with a possessive suffix may be followed by a possessive pronoun. I refer to this as double (possessive) marking. It is unclear what the function of this kind of double constructions is. There is one example in the corpus with the same noun \textit{taman/teman} `friend' used once with double marking, given in~(\ref{exe:tamanunmain}), and several times with just the suffix (an example is given in~\ref{exe:tamanun}).


\begin{exe}
	\ex 
	{\glll Ma wa. Wane tamanun maina.\\
		 ma wa wane taman-\textbf{un} \textbf{main}=a\\
     \textsc{3sg} \textsc{prox}	\textsc{prox} friend-\textsc{3poss} \textsc{3poss=foc}\\
    \glt `Here it is. This one is his friend.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA9-9}{[stim6\_6:38]}}
}
\label{exe:tamanunmain}
\ex 
{\glll Temanun se mia, to? Temanun mat ajakte.\\
	 teman-\textbf{un} se mia to teman-un mat ajak=te\\
    friend-\textsc{3poss} {\glse} come right friend-\textsc{3poss} \textsc{3sg.obj} invite=\glte\\
    \glt `His friends have come, right? His friends invite him.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_23:58]}}
}
\label{exe:tamanun}
\end{exe}

It is tempting to analyse the double-marked nouns as receiving extra prominence – for example, to contrast the referent with another referent – but there is no evidence for this. In~(\ref{exe:tamanunmain}), the speaker picks up a picture and identifies who is in it. It is true that the referent is placed in \is{focus}focus (with focus marker \textit{=a} on the possessive pronoun), but this could also have been achieved by placing the focus marker on the possessive suffix. There is no contrast between two people with a friend in this context. 

It is rare for a noun marked with a possessive suffix to be followed by a possessive pronoun. The corpus has around ten examples. It should be noted that, as with the majority of the nouns marked with just a possessive pronoun, many nouns that have double marking are Malay loanwords. As speculated above, the fact that Papuan Malay uses a periphrastic possessive construction may trigger the use of the possessive pronoun in Kalamang. However, this does not explain the use of a possessive suffix on top of that. (\ref{exe:inja}) shows that the double construction also occurs with indigenous Kalamang words like \textit{in} `name'.

\begin{exe}
	\ex 
	{\gll jadi kalau mu mia=ta eba in-\textbf{ca} \textbf{kain}=at=a taruo\\
		so if \textsc{3pl} come={\glta} then name-\textsc{2sg.poss} \textsc{2sg.poss=obj=foc} say\\
		\glt `So if they come, [you] say your name.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCA-4}{[conv20\_27:24]}}
	}
	\label{exe:inja}
\end{exe}

A possessive suffix and a freestanding possessive pronoun may also be adjacent when the suffix and the pronoun designate different referents. This can happen when the suffix is needed to make a compound, such as `foot-print' in~(\ref{exe:korter}). The possessive pronoun is then needed to indicate the possessor of the footprint. This is very rare: only two examples exist in the corpus.

\begin{exe}
	\ex 
	{\gll an me kor-an=a di=kor-ter-\textbf{un} \textbf{main}=ko\\
		\textsc{1sg} {\glme} foot-\textsc{1sg.poss=foc} \textsc{caus}=foot-mark-\textsc{3poss} \textsc{3poss=loc}\\
		\glt `As for me, I put my foot in his footprints.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_25:11]}}
	}
	\label{exe:korter}
\end{exe}

\section{Possessive constructions with a pronoun and possessive suffix}
\label{sec:possnew}
A small minority of the possessive constructions are made with a basic pronoun followed by a noun inflected with the possessive suffix. The pronoun and noun both refer to the subject of an intransitive verb.

\begin{exe}
	\ex \gll \textbf{ma} tumun-\textbf{un} sem\\
	\textsc{3sg} child-\textsc{3poss} afraid\\
	\glt `Her child is afraid.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDB-C}{[narr28\_3:53]}}
	\label{exe:ma-un}	
	\ex \gll \textbf{pi}=nan et-\textbf{pe} saerak=et me siamar\\
	\textsc{1pl.incl}=too canoe-\textsc{1pl.incl.poss} \textsc{neg.exist}={\glet} {\glme} very\_bad\\
	\glt `If we don't have a canoe either it's very bad.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_17:46]}}
	\label{exe:pi-pe}
\end{exe}	

Because this is a very rare construction with only six occurrences in the corpus, and the phenomenon has not been pursued with further elicitation, there is not enough data available to determine the use and meaning of this construction. It is reminiscent of the topic constructions described in §\ref{sec:leftd}.

\section{Associative relations with \textit{=kin}}
\label{sec:asspl}
The clitic \textit{=kin}, homophonous with second-person plural possessive pronoun \textit{kin}, expresses prospective or purposive association, spatial association\is{association}, general (associative) ownership and representation of type. It attaches to the right edge of the NP. The clitic does not display regular morphophonological behaviour (§\ref{sec:elis}). There is lenition\is{lenition} of the /k/ in, for example, the form \textit{neba=kin}, but not in \textit{mu=kin}. There is also no velarisation after /n/. This is perhaps a sign of recent grammaticalisation. Note that in possessive relations with \textit{=kin}, it is the possessor that is marked.

Prospective or purposive association indicates that the word it is marked on is the prospective or purposive use of the referent of another noun. For instance, the compound\is{compound} \textit{nika warkin}, from \textit{nika} `line' and \textit{war} `to fish', indicates that the line is meant for fishing and not for another activity such as tying.  \textit{Muap terkin} is used to indicate all food that is associated with drinking tea, such as cakes and cookies. Here, \textit{=kin} may mark accompaniment or purpose.

\begin{exe}
	\ex \gll nika war=\textbf{kin}\\
	line fish={\glkinp}\\
	\glt `fishing line' \jambox*{[overheard]} %object/purpose
	\ex \gll kai tolas=\textbf{kin}\\
	firewood break\_fast={\glkinp}\\
	\glt `firewood to be used when breaking the fast' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_31:27]}}%object/purpose}
	\ex \gll tumtum=\textbf{kin}\\
	children={\glkinp}\\
	\glt `children's' (used in the context of medicine) \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE4-8}{[narr31\_2:38]}}%object/purpose/beneficiar}y?
	\ex \gll muap ter=\textbf{kin}\\ 
	food tea={\glkinp}\\	
	\glt `food to consume together with tea' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD8-4}{[narr1\_3:38]}}
	\label{exe:muapter}
\end{exe}

\textit{=kin} is also found on \is{location}locations to indicate spatial association.

\begin{exe}
	\ex \gll pasar Pakpak=\textbf{kin}\\
	market Fakfak={\glkinp}\\
	\glt `Fakfak's market' \jambox*{[overheard]}
	\ex \gll kewe Arepner=\textbf{kin}\\
	house Arepner={\glkinp}\\	
	\glt `Arepner's houses' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_24:04]}}
\end{exe}

Other examples of spatial association are the names of certain species, especially fish species, as shown in~(\ref{exe:wienart}). The \textit{wienar} `parrotfish' is contrasted with other \textit{wienar} species, such as \textit{wienar saruam} `longnose parrotfish'. It can also be used in a more general sense to group for example all fish found close to the shore, as in~(\ref{exe:sorkis}). 

\begin{exe}	
	\ex \gll wienar tebol=\textbf{kin}\\
	parrotfish reef\_edge={\glkinp}\\
	\glt `roundhead parrotfish' \jambox*{\href{https://dictionaria.clld.org/units/kalamang-wienar_tebolkin}{[dict\_wienar tebolkin]}}
	\label{exe:wienart}
	\ex \gll sor kibis=\textbf{kin}\\
	fish shore={\glkinp}\\
	\glt `shore fish' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_11:41]}}
	\label{exe:sorkis}
\end{exe}	
	
\textit{=kin} can also express general (associative) ownership without specifying the possessor. This can be used, for example, with the question word \textit{naman} `who', as in~(\ref{exe:namankin}). The answer to this question also contains \textit{=kin}, marking the possession of a group of people. Thus, \textit{naman=kin} functions as a \is{placeholder}placeholder for \textit{Dian taraun mu=kin}. The answer also contains the third-person plural pronoun \textit{mu} used as an associative plural.\is{associative plural}

\begin{exe}
	\ex 
	\begin{xlist}
		\exi{A:}
	{	\gll naman=\textbf{kin}\\
		who=\textsc{poss}\\
		\glt `Whose?' }
		\exi{B:}
	{	\gll Dian tara-un mu=\textbf{kin}\\
		Dian grandfather-\textsc{3poss} \textsc{3sg}=\textsc{poss}\\
		\glt `Dian's grandfather's and associates'.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_16:17]}}}
	\end{xlist}	
\label{exe:namankin}
\end{exe}

This same marking of the possession of a group, often used with teknonyms\is{names!teknonymy} (see §\ref{sec:names}), is also illustrated in~(\ref{exe:ass}).

\begin{exe}
	\ex
	\label{exe:ass}
	{\glll Pertama nebainggo, Sainudinkin. Sainudinkin lewat, a terus menggara Kiba mukin.\\
		pertama neba=\textbf{kin}=ko sainudin=\textbf{kin} sainudin=\textbf{kin} lewat a terus mengga=ta kiba mu=\textbf{kin}\\
		first \textsc{ph-poss=loc} Sainudin-\textsc{poss} Sainudin=\textsc{poss} pass \textsc{int} further \textsc{dist.lat}={\glta} Kiba \textsc{3pl=poss}\\
		\glt `First whose place, Sainudin's. Pass Sainudin's, ah then from there [you get to] Kiba's.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C9A-E}{[narr37\_1:34]}}
	}
\end{exe}

Interestingly, because \textit{=kin} is attached to the last element of the NP, it can also be attached to a possessive pronoun. In~(\ref{exe:esakainkin}), the use of \textit{=kin} (instead of third-person possessive \textit{-un}) signals that the possessed item, some fishing equipment, belongs to the father's household and not to him personally.

\begin{exe}
	\ex
	\label{exe:esakainkin}
	{\gll mera ge {\ob}esa kain{\cb}=\textbf{kin} mera ki napaki=ta\\
		then not father \textsc{2sg.poss=poss} then \textsc{2pl} use={\glta}\\
	\glt `If not, you can use your father's.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_16:50]}}
	}
\end{exe}

In~(\ref{exe:konkin}), \textit{=kin} attaches to a numeral, which refers to the price of a certain size of fishing line (the word \textit{nika} `line' is elided). \textit{=kin} indicates that the numeral is representative of a type.

\begin{exe}
	\ex 
	\gll ripi putkon=\textbf{kin} me minggi winyal\\
	thousand ten=\textsc{poss} {\glme} \textsc{dist.ins} sail\_fish\\
	\glt `With [fishing line of the kind that costs] ten thousand [you go] fishing while sailing around.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_10:43]}}
\label{exe:konkin}
\end{exe}

Lastly, \textit{=kin} is used by one speaker in two recordings about canoe building to refer to parts of the canoe. In~(\ref{exe:tantayonkin}), the left and right side of the canoe are described, and either side is marked with \textit{=kin}. In~(\ref{exe:ass3}), the canoes are the parts, and the whole is a tree from which three canoes are cut. An alternative analysis is that these uses of \textit{=kin} do not indicate a part-whole relationship, but that there is an understood nominal referent with which \textit{=kin} expresses an associative relationship. In~(\ref{exe:tantayonkin}), this could be the drill holes (the left side's drill holes), and in~(\ref{exe:ass3}), the wood or the motor boat's base.

\begin{exe}
	\ex
	{\gll
		tantayon=\textbf{kin} di=saran=i koyet koi tanbes=\textbf{kin}=at koi bor=taet\\
		left=\textsc{poss} \textsc{caus}=ascend={\gli} finish then right=\textsc{poss}=\textsc{obj} then drill=again\\
		\glt `After putting up the left side, we drill the right side again.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB4-6}{[narr14\_6:00]}}
	}
	\label{exe:tantayonkin}
	\ex
	{\gll
		koi et mir-eir=\textbf{kin}=at kon-i potma raor=\textbf{kin}=at potma limabelas=\textbf{kin}=at potma koi et pokpok=\textbf{kin} kon\\
		then canoe \textsc{clf\_canoe}-two=\textsc{poss=obj} one-\textsc{objqnt} cut middle=\textsc{poss=obj} cut 15\_hp\_motor\_boat.\textsc{mly=poss=obj} cut then canoe small\_motor\_boat=\textsc{poss} one\\
		\glt `Then [I] cut one canoe from two, cut it in the middle, cut a motor boat, then a small motorboat is one.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDD-5}{[narr42\_0:48]}}
	}
	\label{exe:ass3}
\end{exe}
\is{possessive|)}
