\chapter{The clause}
\label{ch:clause}
This chapter describes aspects of the structure of the simple clause\is{simple clause}. Multiclausal constructions are described in Chapter~\ref{ch:compl}. This chapter is structured as follows. An overview with the characteristics of the predicate and core arguments\is{argument!core} is given in §\ref{sec:clintro}. In §\ref{sec:verbal}, declarative verbal clauses with different valencies are described, as well as comparative constructions. Non-verbal declarative clauses are described in §\ref{sec:nonverbal}. Non-declarative clauses, viz. questions and imperatives, are treated in §\ref{sec:nondecl}. Clausal negation is described in §\ref{sec:clauneg}. Variation in the syntax of the clause (elision and reordering of arguments) is treated in §\ref{sec:pragvar}.

\section{Overview}
\label{sec:clintro}
A clause is a grammatical unit that consists minimally of a predicate. A predicate is an element that takes a subject to form a clause, and it expresses something about this subject. The predicate is usually a verb, but can also be a noun, demonstrative, quantifier or location. Each verbal predicate licenses a number of arguments. An intransitive verb requires one argument: a subject NP. A transitive verb requires a subject and an object NP. Ditransitive verbs require an indirect object NP in addition to a subject and object NP. Semantic roles are sometimes used to describe constituent order. I distinguish agents (A), patients (P), themes (T) and recipients (R). Kalamang also has peripheral arguments, such as comitatives or instrumentals, which are expressed with help of postpositional enclitics (§\ref{sec:case}). Complement clauses (§\ref{sec:complclause}) are another type of peripheral argument.

All arguments precede the predicate. The constituent order\is{constituent order} in a basic and unmarked transitive clause is subject-object-verb, with nominative-accusative alignment\is{alignment}\is{accusative|see{alignment}}. 

A simplified template of the clause structure is given in~(\ref{exe:clstr}). A large number of clausal modifiers, including aspect, mood, modal markers and adverbials, occur in different slots in the clause and are described in Chapter~\ref{ch:clausmod}. Many adverbials, aspect markers and mood markers attach to the predicate. The predicate can be complex and is, except in Chapter~\ref{ch:clausmod}, discussed in Chapter~\ref{ch:verbs} and Chapter~\ref{ch:svc}. In this chapter, I focus on simple verbal and non-verbal predicates.

\begin{exe}
	\ex clause structure\\
	(Subj NP) (Obj NP=\textsc{obj}) (Indirect Obj NP) Pred
	\label{exe:clstr}
\end{exe}

Subject and object are formally differentiated by constituent order and object marking (\textit{=at}) on the last constituent of the object NP. (\ref{exe:tr}) and (\ref{exe:intr}) illustrate these characteristics for a transitive and intransitive clause, respectively.

\begin{exe}
	
	\ex
	{\gll bal se sor=at koraru\\
		dog {\glse} fish=\textsc{obj} bite\\
		\glt `The dog has bitten the fish.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBA-8}{[stim2\_3:45]}}
	}
	\label{exe:tr}
	\ex
	{\gll in se min\\
		\textsc{1pl.excl} {\glse} sleep\\	
		\glt `We slept.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_0:43]}}
	} 
	\label{exe:intr}	
	
\end{exe}

Clauses may be the complement of speech and thought verbs like \textit{toni} `say' (§\ref{sec:complclause}), and may be modified by clausal modifiers like negation (§\ref{sec:clauneg}) and mood, aspect, modal and adverbial modifiers (Chapter~\ref{ch:clausmod}). They may be combined with the conjunctions treated in~§\ref{sec:clauseconj}.

\section{Verbal clauses}
\label{sec:verbal}
Verbal clauses have a verb as predicate. Verbs may be monovalent (intransitive), bivalent (transitive), trivalent (ditransitive), or zero-intransitive (lacking an argument altogether). This section starts with an overview of clauses with different valencies in §\ref{sec:valencies}. Comparative constructions are described in §\ref{sec:compcon} and existential and possessive clauses in §\ref{sec:extpos}. This section exclusively describes affirmative declarative clauses.

\subsection{Valency}\is{valency}
\label{sec:valencies}
This section describes the marking of grammatical relations in clauses with different valencies. This is done with constituent order and with postpositions\is{postposition} on some of the arguments. Transitive and intransitive clauses are described in §\ref{sec:valency}. Three-participant events are treated in §\ref{sec:give}. Zero-intransitive clauses are presented in §\ref{sec:amb}. Valency-changing operations are treated in the chapter on verbs, in §\ref{sec:val}.

\subsubsection{Transitive and intransitive clauses}
\label{sec:valency}
Intransitive clauses have a single unmarked subject argument preceding the verb.\is{intransitive clause}\is{stative intransitive verb}

\begin{exe}
	\ex
	{\gll ma tur\\
		\textsc{3sg} fall\\
		\glt `He falls.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD5-6}{[stim34\_0:29]}}
	}
	\label{exe:SV1}
	\ex
	\label{exe:SV2}
	{\gll kaman kos=te\\
		grass grow=\glte\\
		\glt `Grass grows.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B94-F}{[stim21\_2:46]}}
	}
\end{exe}

All intransitive clauses display this pattern irrespective of whether the subject is volitional/actor or non-volitional/undergoer. Active participants such as `they' in~(\ref{exe:SV3}) are formally the same as the undergoer participants such as `he' in~(\ref{exe:SV1}) and `grass' in~(\ref{exe:SV2}), that is, they are unmarked.

\begin{exe}
	\ex
	\label{exe:SV3}
	{\gll mu kiem\\
		\textsc{3pl} run\\
		\glt `They run.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_15:26]}}
	}
\end{exe}

Transitive clauses have two arguments: subject and object. The unmarked constituent order is SOV (APV), although the object can be fronted for \is{focus}focus, which is illustrated in~(\ref{exe:APV2}) and further described in §\ref{sec:leftd} below. The object is marked with an object enclitic \textit{=at} (\ref{sec:at}), also if it is pronominal, as in~(\ref{exe:APV3}).\is{transitive clause}

\begin{exe}
	\ex
	{\gll emun tumun=at narorar\\
		mother child=\textsc{obj} take\_by\_hand\\
		\glt `The mother takes the child by the hand.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA8-B}{[stim4\_4:47]}}
	}
	\label{exe:APV}
	\ex
	{\gll Mujim=at in tok nawanggar\\
		Mujim=\textsc{obj} \textsc{1pl} still wait\\
		\glt `For Mujim we're still waiting.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD8-4}{[narr1\_1:21]}}
	}
	\label{exe:APV2}
		\ex
	{\gll bal ma=at sarie\\
		dog \textsc{3sg=obj} follow\\
		\glt `The dog follows him.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB9-6}{[stim1\_0:18]}}
	}
	\label{exe:APV3}
\end{exe}

While most verbs are either transitive or intransitive, there are also a number of ambitranstive verbs. In all of them, the transitive subject corresponds to the intransitive subject. Ambitransitive\is{ambitransitive verb} verbs include directional verbs like \textit{bara} `descend'; ingestion verbs like \textit{na} `to consume', \textit{muap} `to eat' and \textit{kosom} `to smoke'; and perception verbs like \textit{kome} `to see; to look'. The use of \textit{sara} `to ascend' in an intransitive and a transitive clause is illustrated below.

\begin{exe}
	\ex 
	\begin{xlist}
		\ex \gll an sara\\
	\textsc{1sg} ascend\\
	\glt `I went up.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCF-3}{[narr2\_2:57]}}
	\ex \gll ma afukat=at sara\\
	\textsc{3sg} avocado=\textsc{obj} ascend\\
	\glt `He climbed the avocado tree.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD2-D}{[stim30\_0:17]}}
\end{xlist}
\end{exe}	


\subsubsection{Ditransitive clauses}\is{ditransitive clause|(}\is{three-participant event|see{ditransitive clause}}
\label{sec:give}
In all ditranstive clauses, the subject/agent is unmarked and the direct object/theme is marked with object marker \textit{=at}. Ditransitive clauses with different verbs are marked in different ways. The verbs for showing and sending require that both the theme and recipient are marked with object marker \textit{=at}. The verbs for selling, buying, asking and saying require that the theme is marked with object marker \textit{=at} and the recipient with animate locative \textit{=konggo} or animate lative \textit{=kongga}. Give-constructions are also monoclausal constructions with three arguments. The recipient is optionally preceded by causative \textit{di=} and, if the recipient is nominal rather than pronominal, followed by benefactive \textit{=ki}. Table~\ref{tab:ditrans} gives an overview of the different constructions found with ditransitive verbs.


\begin{table}
	\caption{Ditransitive clause constructions}
	\label{tab:ditrans}
	
		\begin{tabular}{l l l l}
			\lsptoprule
			to show; to send & A & T=\textsc{obj} & R=\textsc{obj}\\
			to sell; to buy; to ask; to say & A & T=\textsc{obj} & R=\textsc{an.loc/an.lat}\\
			to give & A & T=\textsc{obj} & (\textsc{caus}=)R(=\textsc{ben})\\ 
			\lspbottomrule
		\end{tabular}
	
\end{table}


\subsubsubsection{Showing and sending}
The first strategy, with object marker \textit{=at} on both theme\is{theme} and recipient\is{recipient}, was elicited with the verbs \textit{naunak} `to show', \textit{kiempanaet} `to send' and \textit{kama} `to send'. The constituent order can be either agent-theme-recipient-verb (ATRV) or agent-recipient-theme-verb (ARTV), as illustrated in~(\ref{exe:naunak}).

\begin{exe}
	\ex 
	\begin{xlist}
		\ex 
	{\gll tumun\textsubscript{A} guru=at\textsubscript{T} buk=at\textsubscript{R} naunak\\
		child teacher=\textsc{obj} book=\textsc{obj} show\\
		\glt `The child shows the teacher the book.'
	}
		\ex 
{\gll tumun\textsubscript{A} buk=at\textsubscript{R} guru=at\textsubscript{T} naunak\\
	child book=\textsc{obj} teacher=\textsc{obj} show\\
	\glt `The child shows the teacher the book.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_3P\_2]}}
}
\end{xlist}
\label{exe:naunak}
\end{exe}

In elicitation it was easy to get all three participants mentioned, but in the corpus the theme is almost always elided, as in~(\ref{exe:naunakte}). The theme, the object that the speaker wants the addressee to show, is clear from the context.

\begin{exe}
	\ex 
	{\gll ka\textsubscript{A} enem=at\textsubscript{R} naunak=te\\
		\textsc{2sg} woman=\textsc{obj} show=\textsc{imp}\\
		\glt `You show [it] to the woman!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC8-8}{[stim26\_7:11]}}
	}
\label{exe:naunakte}
\end{exe}

Only one example in the corpus also has the theme mentioned. The theme, 150,000 rupiah, expressed as just a numeral, is marked with the number object marker \textit{-i} (§\ref{sec:quantinfl}). The recipient, \textit{an} `me', is marked with the general object marker \textit{=at}.

\begin{exe}
	\ex 
	{\gll ma\textsubscript{A} reitkon purap-i\textsubscript{T} an=at\textsubscript{R} kama=et\\
		\textsc{3sg} hundred fifty-\textsc{objqnt} \textsc{1sg=obj} send={\glet}\\
		\glt `He sent me one hundred and fifty [thousand rupiah].' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_3:09]}}
	}
\label{exe:kamat}
\end{exe}

\subsubsubsection{Selling, buying, asking and saying}
The second strategy, with the recipient marked with animate locative \textit{=konggo} or lative \textit{=kongga}, can be used with (at least) \textit{parin} `to sell', \textit{jie} `to buy', \textit{gerket} `to ask' and \textit{toni} `to say'. (\ref{exe:angonggoparin}) shows an object marker on the theme and a locative marker on the pronoun. (\ref{exe:esunkonggara}), on the other hand, has the number object \textit{-i} suffix on the number (as in~\ref{exe:kamat} above) and both locative and object marking on the recipient. The constituent order is agent-theme-recipient-verb, and in this case theme and recipient cannot be swapped.

\begin{exe}
	\ex 
	{\gll ma\textsubscript{A} sor=at\textsubscript{T} an=konggo\textsubscript{R} parin\\
		\textsc{3sg} fish=\textsc{obj} \textsc{1sg=an.loc} sell\\
		\glt `He sells me fish.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_3P\_3]}}
}
\label{exe:angonggoparin}
\ex 
{\gll tumun me me\textsubscript{A} don kon-i=a\textsubscript{T} esun-kongga=at=a\textsubscript{R} gerket\\
	child \textsc{dist} {\glme} thing one-{\gli}=\textsc{foc} father.\textsc{3poss}-\textsc{an.lat=obj=foc} ask\\
	\glt `The child asks its father a thing.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_3P\_5]}}
}
\label{exe:esunkonggara}
\end{exe}

As with the first strategy above, the theme is usually elided in natural speech. Consider the following example, where the theme (\textit{et} `canoe') is understood from the context.

\begin{exe}
	\ex 
	{\gll an\textsubscript{A} ki=konggo\textsubscript{R} parin=nin\\
		\textsc{1sg} \textsc{2pl=an.loc} sell=\textsc{neg}\\
		\glt `I don't want to sell [my canoe] to you guys.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDD-5}{[narr42\_33:47]}}
	}
\end{exe}

The only corpus example using the second strategy with the theme present has a theme NP with a number modifying the noun, again with the number object \textit{-i} suffix present.

\begin{exe}
	\ex 
	{\gll in\textsubscript{A} me goni kon-i\textsubscript{T} emun\_caun mu=kongga=at-a\textsubscript{R} jie=ta me\\
		\textsc{1pl.excl} {\glme} sack one-\textsc{objqnt} aunt \textsc{3pl=an.lat=obj=foc} buy={\glte} {\glme}\\
		\glt `We bought a sack from his aunt's family.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_2:08]}}
}
\end{exe}

\subsubsubsection{Giving}
The verb for `give' is a zero morpheme that requires three arguments.\footnote{Other languages with a zero morpheme for `give' are the Gum languages of Papua New Guinea, notably \ili{Amele} \citep{roberts1998give}, and \ili{Koasati}, a Muskogean language \citep{kimball1991koasati}.} The three participants in a give-construction\is{give-construction|(} are the agent (A), the theme (T) and the recipient (R). There are two elements that are combined in various constellations to make give-constructions. These are \is{causative}causative \textit{di=} (§\ref{sec:di}) and benefactive\is{benefactive} \textit{=ki} (§\ref{sec:ben}). Give-constructions differ depending on whether the recipient NP is a noun or a pronoun. Constructions with a noun as recipient take the form A T=\textsc{obj} (\textsc{caus}=)R=\textsc{ben} ∅, where the causative proclitic is optional. The benefactive \textit{=ki} is obligatory here. Constructions with a pronoun as recipient take the form A T=\textsc{obj} (\textsc{caus}=)R ∅. Again, the causative proclitic is optional. Note that in the case of pronominal recipients, however, the benefactive enclitic \textit{=ki} is ungrammatical. Table~\ref{tab:give} gives an overview of the possible constructions. These were elicited based on the corpus example in~(\ref{exe:give1}). All constructions in the table are found in the corpus.

\begin{exe}
	\ex
	{\gll ma\textsubscript{A} sandal=bon ladan=bon=at\textsubscript{T} di=teman-un=ki\textsubscript{R} ∅\\
		\textsc{3sg} sandal=\textsc{com} shirt=\textsc{com=obj} \textsc{caus}=friend-\textsc{3poss}=\textsc{ben} give\\
		\glt `He gives the sandals and shirt to his friend.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA8-B}{[stim4\_0:41]}}
	}
	\label{exe:give1}
\end{exe}


\begin{table}
	\caption[Give-constructions]{All possible give-constructions for the clauses `he gives the sandals to his friend' and `he gives the sandals to me'.}
	\label{tab:give}
\fittable{
		\begin{tabular}{l l l l l l l l}
			\lsptoprule
			&	 	& A & T=\textsc{obj} & \textsc{caus}= & R & =\textsc{ben} & give \\ \midrule
			Nominal R & Option 1	& \textit{ma} &	\textit{sandal=at} & \textit{di}= & \textit{temanun} & \textit{=ki}& \textit{∅}\\
			&& \textsc{3sg} & sandal=\textsc{obj} & \textsc{caus}= & his friend & =\textsc{ben} & give\\
			& Option 2 & \textit{ma} & \textit{sandal=at} & & \textit{temanun} &\textit{=ki} & \textit{∅}\\
			&& \textsc{3sg} & sandal=\textsc{obj} & & hisfriend & =\textsc{ben} & give\\
			\midrule 
			Pronominal R & Option 1 & \textit{ma} & \textit{sandal=at} & \textit{di=} & \textit{an} &&\textit{∅}\\
			&& \textsc{3sg} & sandal=\textsc{obj} & \textsc{caus}= & \textsc{1sg} &&give \\
			& Option 2 & \textit{ma} & \textit{sandal=at} &&\textit{an}&&\textit{∅}\\ 
			&& \textsc{3sg} & sandal=\textsc{obj} &  & \textsc{1sg} && give\\
			\lspbottomrule
		\end{tabular}
		}
\end{table}

Because Kalamang lacks an overt lexeme `give', the meaning of giving is essentially expressed by the presence and order of the participants, (optionally) accompanied by \textit{di=} and/or \textit{=ki}. The theme often remains unexpressed. The least elaborate give-construction therefore is two subsequent NPs: A and R, if the recipient NP is a pronoun. Despite this very minimal construction, a clause consisting of a noun followed by a pronoun (\textit{ema ma} `mother gives him/her') or of two pronouns (\textit{ma ma} `he/she gives him/her') is always interpreted as a give-construction by Kalamang speakers.

The position of the zero morpheme is after the recipient as in~(\ref{exe:main}) and~(\ref{exe:amat}), or after the benefactive if present as in~(\ref{exe:ditumunggi}). This is evident from the position of, for example, mood or negation marking. On the surface, everything that is attached to the verb is attached to the recipient, and is also sometimes morphophonologically integrated, as shown in~(\ref{exe:main}). Give-constructions may also be part of \is{predicate!complex}complex predicates (Chapter~\ref{ch:svc}), such as the one in~(\ref{exe:pada}), where the verbs are linked with predicate linker \textit{=i}.

\begin{exe}
	\ex 
	{\glll Ma birara main.\\
		ma bir=at=a ma ∅=\textbf{kin}\\
		\textsc{3sg} beer=\textsc{obj=foc} \textsc{3sg} give={\glkin}\\
		\glt `He wants to give him beer.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA8-B}{[stim4\_1:58]}}
	}
	\label{exe:main}
	\ex 
	{\gll ma am=at ma ∅=\textbf{nin}\\
		\textsc{3sg} breast\_milk=\textsc{obj} \textsc{3sg} give=\textsc{neg}\\
		\glt `She didn't give him milk.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA7-D}{[narr21\_1:58]}}
	}
	\label{exe:amat}
	\ex \gll sor me selet-kon-i tete=ki ∅=\textbf{te}\\
	fish {\glme} piece-one-\textsc{objqnt} grandfather=\textsc{ben} give=\textsc{imp}\\
	\glt `Give one piece of fish to grandfather!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_6:01]}}
	\label{exe:ditumunggi}
	\ex \gll naman=a padanual=at \textbf{rep=i} \textbf{ka} ∅\\
	who=\textsc{foc} pandanus=\textsc{obj} get={\gli} \textsc{2sg} give\\
	\glt `Who got pandanus [leaf] and gave it to you?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB5-B}{[conv17\_23:37]}}
	\label{exe:pada}
\end{exe}

Benefactive \textit{=ki} was also elicited in non-give constructions such as the one exemplified in~(\ref{exe:kiunggi}). Benefactive \textit{=ki} does not occur on pronouns, parallel to \textit{=ki} in give-constructions.

\begin{exe}
\ex 
{\gll ma kewe giar=ten=at paruon=i kiun=ki\\
	\textsc{3sg} house new=\textsc{at=obj} make={\gli} wife.\textsc{3poss=ben}\\
	\glt `He makes a new house for his wife.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_cla\_18]}}
}
\label{exe:kiunggi}
\end{exe}\is{ditransitive clause|)}\is{give-construction|)}

\subsubsection{Zero-intransitive clauses}\is{zero-intransitive clause}\is{ambient clause|see{zero-intransitive clause}}
\label{sec:amb}
Zero-intransitive clauses lack an argument \parencite[][267]{dryer2007}. Typically, these are clauses about environmental conditions, which is why they may also be called ambient clauses. (\ref{exe:kalizs}) illustrates how a weather phenomenon (\textit{kalis} `to rain') is expressed without argument. Other environmental conditions, such as fog, are expressed in the same way, but this is a rather marginal phenomenon in the corpus.

\begin{exe}
	\ex
{\gll mu toni kalis=kin\\
	\textsc{3pl} think rain={\glkin}\\
	\glt `They think it will rain.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC0-1}{[narr22\_4:04]}}
}
\label{exe:kalizs}
\end{exe}


\subsection{Comparative constructions}\is{comparative}
\label{sec:compcon}
In a comparative construction, two referents are compared on a gradable property. One referent, the comparee, is compared to a standard of comparison, the other referent. The construction also typically involves a parameter of comparison (the property that is compared). Kalamang employs two comparative constructions: a monoclausal `exceed' construction and a biclausal construction with antonyms\is{antonym}, neither of which involve a parameter of comparison.

The majority of elicited and natural speech comparative constructions are monoclausal, making use of \textit{lebe} `to exceed', illustrated in~(\ref{exe:matlebe}) and~(\ref{exe:hargalebe}). \textit{Lebe} `to exceed' is a loan from Malay \textit{lebih} `more (than)', which is also used in comparative constructions in Malay varieties.

\begin{exe}
	\ex \label{exe:matlebe}
\gll ma tanbes=ko ror mat \textbf{lebe}\\
\textsc{3sg} right=\textsc{loc} tree \textsc{3sg.obj} exceed\\
\glt `He's on the right, the tree exceeds him.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC8-8}{[stim26\_7:46]}}

	\ex \label{exe:hargalebe}
	\gll harga-un main me tang-un=at \textbf{lebe}\\
	price-\textsc{3poss} \textsc{3poss} {\glme} seed-\textsc{3poss=obj} exceed\\
	\glt `Its price exceeds the seed['s price].' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BF1-6}{[narr12\_6:21]}}
\end{exe}

A monoclausal exceed construction may also be made with \textit{nemies} `to exceed', but is much less frequent (one natural speech example and one corpus example, against at least ten of each for \textit{lebe}). (\ref{exe:nemies}) and (\ref{exe:nemies2}) are the two attested instances. Based on these two examples, it is possible that \textit{nemies} actually has the more specific meaning `to exceed in height'.

\begin{exe}
	\ex \label{exe:nemies}
	\gll warkin se mat \textbf{nemies}\\
	tide {\glse} \textsc{3sg.obj} exceed\\
	\glt `The tide exceeded him.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC1-0}{[narr19\_16:24]}}
	\ex \label{exe:nemies2}
	\gll ibu me ririn-un an=at nemies\\
	miss {\glme} tall-\textsc{nmlz} \textsc{1sg=obj} exceed\\
	\glt `Miss's length exceeds mine.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[comp\_5]}}
\end{exe}

In elicitation, speakers sometimes phrased a comparative construction with antonyms. That construction consists of two independent clauses, where one clause contains the standard and the other the comparee. This is possibly an artefact of the elicitation, which involved stimulus materials showing opposite types (e.g. a skinny and a fat person, or an old and a new house). An example of such a construction with antonymous predicates is seen in~(\ref{exe:temuncicaun}). It remains for further research whether this is a construction that is also used in naturalistic speech.

\begin{exe}
	\ex \label{exe:temuncicaun}
	{\gll pas wa me kaden-un temun pas wa me kaden-un cicaun\\
		woman \textsc{prox} {\glme} body-\textsc{3poss} big woman \textsc{prox} {\glme} body-\textsc{3poss} small\\
		\glt `This woman's body is big, this woman's body is small.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_comp\_2]}}
	}	
\end{exe}


\subsection{Existential and possessive clauses}
\label{sec:extpos}
Existential and possessive clauses are expressed with existential \textit{mambon}. (\ref{exe:kewmam}) is an existential clause. Possessive clauses, like~(\ref{exe:huka-ca}), also include a possessive suffix on the possessed or a possessive pronoun.

\begin{exe}
	\ex \gll kewe metko \textbf{mambon}\\
	house \textsc{dist.loc} \textsc{exist}\\
	\glt `There is a house there.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAE-4}{[narr44\_2:59]}}
	\label{exe:kewmam}
	\ex \gll ka me hukat-\textbf{ca} \textbf{mambon}\\
	\textsc{2pl} {\glme} fishing\_net-\textsc{2sg.poss} \textsc{exist}\\
	\glt `Do you have your fishing net?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCB-5}{[conv1\_6:38]}}
	\label{exe:huka-ca}
\end{exe}	

For the negation of possessive clauses, see §\ref{sec:negnonv}. Other possessive constructions are described in Chapter~\ref{ch:poss}.

\section{Non-verbal clauses}\is{non-verbal clause|(}
\label{sec:nonverbal}
Non-verbal clauses are clauses whose predicate\is{predicate} belongs to a class other than verb, and which can function as the head of a NP. This kind of clause is rather common in Kalamang, as any property or argument can act as a predicate with no overt copula\is{copula*} needed. This section is dedicated to affirmative declarative non-verbal clauses. Nominal clauses are described in §\ref{sec:nomclau}, demonstrative clauses are described in §\ref{sec:demclau}, quantifier clauses in §\ref{sec:quantclau} and locative, lative and similative clauses in §\ref{sec:locclau}. Equative constructions (§\ref{sec:eq}) and similative constructions (§\ref{sec:sim}) are also non-verbal clauses.

Kalamang has no copula verb, although \is{topic}topic marker \textit{me} (§\ref{sec:discme}) is often found with both stative verbal and non-verbal predicates. As shown in §\ref{sec:wcverb}, the construction of a noun and a stative verb is ambiguous between a copular clause and a NP reading. The use of topic marker \textit{me} makes an unambiguous copular clause.


\begin{exe}
	\ex 
	\begin{xlist}
		\ex {\gll {\ob}tumun tabusik{\cb}\textsubscript{NP}\\
			child short\\
			\glt `the/a short child'}
		\ex {\gll {\ob}tumun{\cb}\textsubscript{NP} {\ob}tabusik{\cb}\textsubscript{Pred}\\
			child short\\
			\glt `The child is short.}
		\ex {\gll {\ob}tumun me{\cb}\textsubscript{NP} {\ob}tabusik{\cb}\textsubscript{Pred}\\
			child {\glme} short\\
			\glt `The child is short.' \jambox*{[elic]}}
	\end{xlist}
	\label{exe:tabusikme}
\end{exe}

\subsection{Nominal clauses}\is{nominal clause|(}
\label{sec:nomclau}
In nominal clauses, a NP headed by a noun functions as a predicate. Nominal clauses typically consist of two juxtaposed NPs, expressing an identity relationship between them. These clauses can also be called equational clauses. I analyse the first NP as the subject and the second NP as the predicate based on the fact that arguments typically precede the predicate in Kalamang. (\ref{exe:konguru}) contains two nominal clauses expressing someone's occupation, and~(\ref{exe:paskel}) the subject NP is headed by \textit{in} `name'.

\begin{exe}
	\ex 
	{\gll {\ob}kon{\cb}\textsubscript{S} se {\ob}guru{\cb}\textsubscript{Pred}, {\ob}tumtum kon{\cb}\textsubscript{S} {\ob}guru{\cb}\textsubscript{Pred}\\
	one {\glse} teacher children one teacher\\
\glt `One is already teacher, one child is teacher.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_9:15]}}
}
\label{exe:konguru}
\ex 
{\gll {\ob}dun-un in-un{\cb}\textsubscript{S} {\ob}pas Kelengkeleng{\cb}\textsubscript{Pred}\\
	opposite\_sex\_sibling-\textsc{3poss} name-\textsc{3poss} woman Kelengkeleng\\
	\glt `His sibling's name was ``Kelengkeleng woman''.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBC-4}{[narr24\_6:08]}}
} \label{exe:paskel}
\end{exe}

The predicate NP in nominal clauses may also be a possessive pronoun. An example is given in~(\ref{exe:hakin}).

\begin{exe}
	\ex 
	{\gll karena hak inggon\\
	because task \textsc{1pl.excl.poss}\\
	\glt `Because the task is ours.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C98-7}{[narr26\_18:55]}}
} \label{exe:hakin}
\end{exe}

Although nominal clauses may occur without topic marker \textit{me} (§\ref{sec:discme}), frequently they have one. (\ref{exe:rarait}) and~(\ref{exe:kamanget}) show two nominal clauses with \textit{me} in between the subject and the predicate.

\begin{exe}
	\ex
	{\gll Nema namun me sontum Rarait\\
		Nema husband.\textsc{3poss} {\glme} person Seram \\
		\glt `Nema's husband is Seramese.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BFF-9}{[dict\_Rarait]}}
	}
	\label{exe:rarait}
	\ex
{\gll Moktar esun me kamanget\\
	Moktar father.\textsc{3poss} {\glme} medicine\_man\\
	\glt `Moktar's father is a medicine man.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BFF-9}{[dict\_kamanget]}}
}
\label{exe:kamanget}		
\end{exe}

In equational\is{equational clause} clauses (also known as identifying clauses\is{identifying clause}), a distinction can be made between clauses of proper inclusion (where the second NP describes a set of items, also known as ascriptive clauses) and completely equational clauses. The two clause types are different in that in completely equational clauses the two NPs can be swapped, whereas this is impossible for the ascriptive clauses. This is illustrated in~(\ref{exe:realeq}) for a completely equational clause, where the two NPs are entirely coreferential and thus can be switched without problems. In~(\ref{exe:propincl}), a proper inclusive clause is given, where the first NP in example a (a schoolmaster) is part of the set in the second NP (Butonese men). Switching is ungrammatical. Otherwise, clauses of proper inclusion and completely equational clauses are syntactically the same. Note that also in these examples, \textit{me} may be left out.

\begin{exe}
	\ex 
	\begin{xlist}
	\ex 
	{\gll Sebi kiun (me) Kalamang emun\\
		Sebi wife.\textsc{3poss} {\glme} Kalamang mother.\textsc{3poss}\\
		\glt `Sebi's wife is Kalamang's mother. 
	}
	\ex 
	{\gll Kalamang emun (me) Sebi kiun\\
		Kalamang mother {\glme} Sebi wife.\textsc{3poss}\\
		\glt `Kalamang's mother is Sebi's wife.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_eq\_8]}}
	}
	\end{xlist}
	\label{exe:realeq}
	\ex 
	\begin{xlist}
	\ex []
	{\gll kepala\_sekola (me) utun-ca kon\\
		schoolmaster {\glme} Buton-man one\\
		\glt `The schoolmaster is a man from Buton.'
	}
		\ex [*]
	{\gll utun-ca kon me kepala\_sekola\\
	Buton-man one {\glme} schoolmaster\\
	\glt `A man from Buton is the schoolmaster.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_eq\_14]}}
	}
	\end{xlist}
	\label{exe:propincl}
\end{exe}
\is{nominal clause|)}

\subsection{Demonstrative clauses}\is{demonstrative clause|(}
\label{sec:demclau}
In demonstrative clauses, a NP headed by a demonstrative is the predicate. These clauses function as presentational clauses. An example is given in~(\ref{exe:matam}). Note that~(\ref{exe:matam}) also contains a locative clause, which will be described in §\ref{sec:locclau}.

\begin{exe}
	\ex
	{\gll ma tamatko ah {\ob}ma{\cb}\textsubscript{S} {\ob}me{\cb}\textsubscript{Pred}\\
		\textsc{3sg} where \textsc{int} \textsc{3sg} \textsc{dist}\\
		\glt `Where is it? Ah, that's it.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE7-5}{[stim42\_6:40]}}
	}
	\label{exe:matam}
\end{exe}

The expression \textit{ma he me} `that's it; that's enough' is also a demonstrative clause. (\ref{exe:mahem}) is taken from a story where a black-furred monkey wants to turn his fur white. A cuscus puts him in a cage in the rising sea, knowing that everything looks lighter under water. When the monkey sees his legs start turning white, he thinks it's enough and wants to be released.

\begin{exe}
	\ex 
	{\glll Eih koran se iren, ma he me, anat kahetmei!\\
		eih kor-an se iren \textbf{ma} \textbf{se} \textbf{me} an=at kahetmei\\
		\textsc{surpr} leg-\textsc{1sg.poss} {\glse} white \textsc{3sg} {\glse} \textsc{dist} \textsc{1sg=obj} release.\textsc{imp}\\
		\glt `Hey, my legs are already white, that's enough, release me!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC1-0}{[narr19\_14:00]}}
	} \label{exe:mahem}
\end{exe}\is{demonstrative clause|)}

See Chapter~\ref{ch:dems} for an in-depth description of demonstratives.


\subsection{Quantifier clauses}\is{quantifier clause|(}
\label{sec:quantclau}
Quantifier clauses state the quantity of something, and have a NP headed by a quantifier as the predicate. Quantifiers were discussed in Chapter~\ref{ch:quant}. A \is{numeral}numeral quantifier predicate is shown in~(\ref{exe:keretaun}) and a non-numeral quantifier is shown in~(\ref{exe:muawanbol}).

\begin{exe}
	\ex 
	{\gll kereta-un {\ob}kansuor{\cb}\textsubscript{Pred}\\
		cart-\textsc{3poss} four\\
		\glt `It has four carts.' (lit. `Its cars are four.') \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE1-9}{[stim40\_1:44]}}
	}
	\label{exe:keretaun}
	\ex 
	{\gll an muap-an {\ob}bolodak{\cb}\textsubscript{Pred}\\
		\textsc{1sg} food-\textsc{1sg.poss} just\_little\\
		\glt `As for me, I have just a little food.' (lit. `I, my food is just a little.') \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C96-2}{[narr18\_10:37]}}
	}
	\label{exe:muawanbol}
\end{exe}

Based on elicitation, it does not seem possible to have higher numerals in predicative function. Where to draw the boundary line between possible and impossible predicative numerals, if such a boundary exists, remains for further research.\is{quantifier clause|)}


\subsection{Locative and lative clauses}\is{location clause|(}
\label{sec:locclau}
Locative and lative clauses are clauses where the predicate is a NP marked with a locative or lative postposition (§\ref{sec:case}).

Locative clauses express \is{location}location. Locative predicates are NPs that refer to a location, inflected with the locative postposition\is{postposition!locative} \textit{=ko}. A NP carrying locative \textit{=ko} may be preceded or followed by a verb. In such cases, I analyse the locative NP and the verb as a \is{predicate!complex}complex predicate. For the rules governing the ordering of verbs and locatives, see Chapter~\ref{ch:svc}. When a noun inflected with \textit{=ko} is not preceded or followed by a verb, it is the predicate of the clause on its own, as in~(\ref{exe:keweo}). 


\begin{exe}
	\ex
	{\gll mu toni {\ob}ma{\cb}\textsubscript{S} {\ob}kewe-ko{\cb}\textsubscript{Pred}\\
		\textsc{3sg} say \textsc{3sg} house-\textsc{loc}\\	
		\glt `They say he is at home.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_1:37]}}
	}
	\label{exe:keweo}
\end{exe}	
	
NPs marked with the lative postposition\is{postposition!lative} \textit{=ka}, which marks movement from or towards, are usually combined with other verbs. They are very seldom the predicate on their own. One such example, with \textit{=ka} fused with distal demonstrative \textit{me} as \textit{mengga}, is given in~(\ref{exe:Manggan}), where the speaker sums up all the houses she hasn't visited yet (an action which for married women is related to a ritual\is{ritual}). The combination of lative NPs and verbs is described in Chapter~\ref{ch:svc}. 

\begin{exe}
	\ex
	{\gll tete Manggan mu tok {\ob}mengga=nin{\cb}\textsubscript{Pred} weinun\\
		grandfather Manggan \textsc{3pl} yet \textsc{dist.lat=neg} too\\
		\glt `[I] haven't yet been to grandfather Manggan's either.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_24:27]}}
	}
	\label{exe:Manggan}
\end{exe}
\is{location clause|)}

\subsection{Equative constructions}
\label{sec:eq}
Equative constructions\is{equative} (to be distinguished from equational constructions such as `Majid is a fisherman') are constructions that ``express situations in which two referents have a gradable property to the same degree'' \parencite[][9]{haspelmath2015}.\footnote{For comparative constructions, see §\ref{sec:compcon}.} \textcite{haspelmath2015} identifies five components in equative constructions. If we exemplify this with the English phrase `She is as tall as her mother', we have a comparee (`she' [is]), a degree marker (`as'), a parameter (`tall'), a standard marker (`as') and a standard (`her mother'). 

An equative construction is made with \textit{nain} `like' as the standard marker and the similative postposition\is{postposition!similative} \textit{=kap} on the standard. As with comparative constructions, there is no parameter, but the gradable property might be expressed as part of the comparee, as in~(\ref{exe:eq1}).  In~(\ref{exe:eq2}), the verb \textit{bo} `to go' is used to indicate growth. The sentence means literally something like `[it] has become [so that it is] like [the size of] a cassava fruit'.

\begin{exe}
	\ex
	\label{exe:eq1}
	{\gll ma ririn-un \textbf{nain} emun=\textbf{kap}\\
		\textsc{3sg} tall-\textsc{nmlz} like mother.\textsc{3poss=sim}\\
		\glt `She is as tall as her mother.' (lit. `Her height is like her mother.') \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_idem\_6]}}
	}

	\ex
	\label{exe:eq2}
	{\gll se bo=te \textbf{nain} panggala naun=\textbf{kap}\\
		\textsc{asp} go=\textsc{\glte} like cassava fruit=\textsc{sim}\\
		\glt `It's already becoming as big as a cassava.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_6:51]}}
	}
\end{exe}

Kalamang is an `only equative standard marker' language according to the classification in \textcite{haspelmath2015}. We have a parameter (which can be omitted if understood), differentiated comparee, standard, and two equative standard markers \textit{nain} and \textit{=kap}, which apparently fulfil the same function. Note that the equative standard markers are not dedicated to equative constructions, but occcur in similative constructions as well (see §\ref{sec:sim}). Note also that in similative constructions, \textit{nain} and \textit{=kap} can be used independently or in combination, without a difference in meaning. It has not been tested whether equative constructions can be made with either \textit{nain} or \textit{=kap} alone or whether both have to be present.

An alternative strategy is to use the Malay loan \textit{sama} `same'. This was only elicited as shown in~(\ref{exe:sama}).

\begin{exe}
	\ex
	\label{exe:sama}
	{\gll mier ririn-un \textbf{sama}\\
		\textsc{3du} tall-\textsc{nmlz} same\\
		\glt `They have the same height.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_comp\_17]}}
	}
\end{exe}

\subsection{Similative constructions}\is{similative construction}
\label{sec:sim}
There are several ways to make a similative construction, that is, a construction to indicate that a referent has a characteristic similar to another referent. These involve the words \textit{sama} `same' (cf. Malay \textit{sama} `same'), \textit{nain} `like' and the similative postposition\is{postposition!similative} \textit{=kap} (§\ref{sec:simcase}) in different combinations, as shown in Table~\ref{tab:sim}. Kalamang can make similative constructions with just the enclitic \textit{=kap}, just the word \textit{nain}, a combination of \textit{nain} and \textit{=kap}, or (occasionally) a combination of \textit{sama, nain} and \textit{=kap}.  \is{idemative}\is{similative} 

\begin{table}[ht]
	\caption{Similative constructions}
	\label{tab:sim}
	
		\begin{tabular}{l l l l}
			\lsptoprule
			Subj &&& N-\textit{kap}\\
			Subj&&\textit{nain}&\\
			Subj&\textit{sama} & \textit{nain} & N-\textit{kap}\\
			\lspbottomrule
		\end{tabular}
	
\end{table}

The similative postposition \textit{=kap} can be used on its own, as in~(\ref{exe:kap22}) and~(\ref{exe:kap33}). In~(\ref{exe:kap22}), two things are compared, and in~(\ref{exe:kap33}), one person is compared to many. 

\begin{exe}
	\ex
	\label{exe:kap22}
	{\gll ma per=\textbf{kap}=teba\\
		\textsc{3sg} water=\textsc{sim}=\glteba\\
		\glt `It's like water.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB1-3}{[narr34\_2:51]}}
	}
	\ex
	\label{exe:kap33}
	{\gll ka kalamang emumur=\textbf{kap} to\\
		\textsc{2sg} Kalamang woman.\textsc{pl}=\textsc{sim} right\\
		\glt `You are like the Kalamang women, aren't you?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_6:51]}}
	}
\end{exe}

In~(\ref{exe:kapnain}), \textit{=kap} is used in combination with \textit{nain}, which also translates as `like'. This is not uncommon in the corpus, but it is unclear what drives the choice of a double similative marking. Constructions with \textit{nain} N\textit{=kap} are used for equative constructions (as described in §\ref{sec:eq}), but it is common with similative constructions as well.

\begin{exe}
	\ex
	\label{exe:kapnain}
	{\gll ka \textbf{nain} tenggeles=\textbf{kap} ka bo tenggeles=\textbf{kap}=te don kuek=te kuru luk=te yuwa\\
		\textsc{2sg} like eagle=\textsc{sim} \textsc{2sg} go eagle=\textsc{sim}={\glte} thing steal={\glte} bring come={\glte} \textsc{prox}\\
		\glt `You're like an eagle, you turn into an eagle stealing stuff and bringing it back.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDC-D}{[conv8\_1:59]}}
	}
\end{exe}

Like in equative constructions, a speaker can also opt to use the Malay loan word \textit{sama} `same' in similative constructions, such that we get examples with \textit{sama nain} N-\textit{kap}. In~(\ref{exe:gierunggap}), the quality of rice is described, while~(\ref{exe:nainan}) is from a story where a black-furred monkey wants to get the same white fur as a cuscus. This is a rare construction, with only four occurrences in the corpus.

\begin{exe}
	\ex
	\label{exe:gierunggap}
	{\gll ikon me \textbf{sama} \textbf{nain} sontum gier-un=\textbf{kap}\\
		some {\glme} same like person tooth-\textsc{3poss=sim}\\
		\glt `Some are like people's teeth.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_1:48]}}
	}
\end{exe}

\begin{exe}
	\ex \label{exe:nainan}
	\gll me \textbf{sama}=i \textbf{nain} an=\textbf{kap}\\
	\textsc{dist} same={\gli} like \textsc{1sg=sim}\\
	\glt `That is the same as me.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC1-0}{[narr19\_12:22]}}
\end{exe}
%comment on =i?

\textit{Nain} `like' can also be used on its own. (\ref{exe:semenn}) is said two utterances after (\ref{exe:gierunggap}), by the same speaker, still discussing the quality of rice.

\begin{exe}
	\ex
	\label{exe:semenn}
	{\gll ikon me yar ikon me \textbf{nain} semen\\
		some {\glme} stone some {\glme} like cement\\
		\glt `Some are stones, some are like cement.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_1:55]}}
	}
\end{exe}

Similative constructions can also be used to compare entities or persons, such as God and Jesus in~(\ref{exe:isa}).

\begin{exe}
	\ex
	\label{exe:isa}
	{\gll sontum kawir=ten toni Isa me nain tuhan=kap\\
		person christian=\textsc{at} say Jesus {\glme} like god=\textsc{sim}\\
		\glt `Christians say Jesus is like God.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_sim19\_3]}}
	}
\end{exe}

To single out referents (as in~\ref{exe:nainrehan}), a similative construction with only \textit{nain} `like' can be used.

\begin{exe}
	\ex
	\label{exe:nainrehan}
	{\gll tumtum opa me \textbf{nain} Kalamang=bon Rehan=bon mindi\\
		children \textsc{ana} {\glme} like Kalamang=\textsc{com} Rehan=\textsc{com} like\_that\\
		\glt `Those children were like Kalamang and Rehan, like that.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDE-7}{[narr25\_2:51]}}
	}
\end{exe}
%first ex has no =at. probably elicitation effect.

The similative constructions above do not have a demonstrative function. For that purpose, Kalamang has manner demonstratives \textit{wandi} `like this' and \textit{mindi} `like that', forms that are likely derived from the proximal and distal demonstratives, respectively. These can be used in combination with similative constructions, as in~(\ref{exe:nainrehan}) above. Manner demonstratives may be evoked to make similative constructions, as in~(\ref{exe:tanunkoi}), which expresses close similarity between two referents, in this case the position of a doll's arm on a picture. More information about manner demonstratives can be found in Chapter~\ref{ch:dems}.

\begin{exe}
	\ex \label{exe:tanunkoi}
	\gll wa=nan tan-un koi \textbf{mindi} \textbf{weinun}\\
	\textsc{prox}=too arm-\textsc{3poss} again like\_this too\\
	\glt `This one too, his arm is again like this.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B92-E}{[stim44\_1:14]}}
\end{exe}

Similative \textit{=kap} is grammaticalised on colour terms\is{colour term}, which all seem to derive from nouns. This allows us to hypothesise the nominal origins of Kalamang colour terms see Table~\ref{tab:colours}.\is{colour terms} Earlier Kalamang speakers seem to have used phrases meaning `like turmeric' or `like charcoal' to indicate colours.\footnote{This is reminiscent of languages such as Candoshi (isolate, Peru), which describes colours with terms that translate as `like the feathers of a Milvago bird' \parencite{surralles2016}. For Candoshi, it is claimed that these terms only have secondary use as colour terms, and that the language therefore does not really have colour terms. An overarching word `colour' is also lacking. Although the latter is the case for Kalamang as well, and Kalamang colour terms are clearly derived terms, the words mentioned here are used primarily as colour terms. People backtranslate \textit{baranggap} as `yellow', for example, and not as `like turmeric'.}

\begin{table}[ht]
	\caption{Possible nominal origins of Kalamang colour terms}
	\label{tab:colours}
	
		\begin{tabular}{l l}
			\lsptoprule
			colour term & possible origin\\
			\midrule
		\textit{welenggap} `blue; green' & \textit{wele} `vegetables'\\
		\textit{kerkap} `red' & unknown, perhaps \textit{karyak} `blood'\\
		\textit{baranggap} `yellow' & \textit{barang} `turmeric'\\
		\textit{iriskap} `white' & \textit{iren} `ripe; white person'\\
		\textit{kuskap} `black' & \textit{kus} `piece of charcoal'\\
		\lspbottomrule
		\end{tabular}
	
\end{table}

Other, perhaps newer colour terms are not formed according to this pattern. \textit{Kowewep(kon)} `grey; brown' (from \textit{koep} `ashes') does not take \textit{=kap}.

To express sameness or true identity, i.e. where there is a single referent, \textit{kodak} `just one' may be used, as in~(\ref{exe:kianbon}).

\begin{exe}
	\ex \label{exe:kianbon}
	{\gll mu nakal-un tai-\textbf{kodak}=ko\\
		\textsc{3pl} head-\textsc{3poss} side-just\_one=\textsc{loc}\\
		\glt `Their heads are on the same side.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD6-8}{[stim38\_3:08]}}
	}
\end{exe}

\is{non-verbal clause|)}


\section{Non-declarative clauses}\is{non-declarative clause|(}
\label{sec:nondecl}
Declarative clauses were described above. Below, I describe the syntax of the two non-declarative clause types questions (§\ref{sec:ques}) and imperatives (§\ref{sec:impcla}).

\subsection{Interrogative clauses}\is{interrogative clause|(}
\label{sec:ques}
Interrogative clauses serve to request information. In this section, I describe four types of Kalamang interrogative clauses: content questions (§\ref{sec:opq}), polar questions (§\ref{sec:polq}), confirmation-seeking questions (§\ref{sec:confquest}) and alternative questions (§\ref{sec:altquest}). All interrogative clauses adhere to the basic Kalamang word order SOV.

\subsubsection{Content questions}\is{question-word question}\is{content question|see{question-word question}}
\label{sec:opq}
\label{sec:convquest}
Content questions (also question-word questions, wh-questions or information questions) are formed with the question words listed in §\ref{sec:wcq}. Question words occur in the syntactic position\is{in situ} of the argument they are replacing and are typically \is{focus}focused with \textit{=a}. Thus, in~(\ref{exe:nebaO}), \textit{neba} `what' occurs in object position, because that is the syntactic position that the question is after. In~(\ref{exe:whereLOC}), \textit{tamatko} `where' comes before the verb \textit{kos} `to grow', just as any locative would. And in~(\ref{exe:whoA}), \textit{naman} `who' occurs in subject position. The question posed in the narrative in~(\ref{exe:whoA}) is not answered, whereas for the other two examples the answer is also given.

\begin{exe}
	\ex
		\begin{xlist}
		\exi{A:}
	{\gll esun \textbf{neba}=at=a tanggo=ta yuwa\\
		father.\textsc{3poss} what=\textsc{obj=foc} hold={\glta} \textsc{prox}\\
		\glt `What is his father holding here?'
	}

	
	\exi{B:}
		{\gll karajang=at=a tanggo\\
		basket.\textsc{obj=foc} hold\\
		\glt `[He's] holding a basket.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA8-B}{[stim4\_4:38-44]}}
	}

	\end{xlist}
	\label{exe:nebaO}
\end{exe}

\begin{exe}
	\ex
\begin{xlist}
	\exi{A:}
	{\gll ror \textbf{tamatko}=a kos\\
		tree where=\textsc{foc} grow\\
		\glt `Where does the tree grow?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C97-F}{[stim27\_13:55]}}
	}
	\exi{B:}
{\gll ror kir-un=ko tanbes=ko\\
	tree side-\textsc{3poss=loc} right\_side=\textsc{loc}\\
	\glt `The tree is on his side, on the right side.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C97-F}{[stim27\_13:57]}}
}
\end{xlist}
\label{exe:whereLOC}
\end{exe}

\begin{exe}	
	\ex
	\gll \textbf{naman}=a kat sirie\\
		who=\textsc{foc} \textsc{2sg.obj} follow\\
		\glt `Who follows you?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B91-5}{[narr39\_7:50]}}
	\label{exe:whoA}
\end{exe}

\hspace*{-4pt}A subtype of content questions are conventionalised questions\is{conventionalised question} Kalamang speakers use to greet each other. Upon meeting each other outside or passing by someone's house, one asks either `what are you doing' (example~\ref{exe:geo}) or `where are you going/where are you coming from?' (example~\ref{exe:tamanggarabot}). These questions have the same syntax as normal content questions. To the question `what are you doing?' one may either answer with a verb marked with progressive \textit{=teba}, or with the phrases \textit{ge o} or \textit{ge mera}, indicating that one does not want or feel the need to specify what one is doing.\is{conversation}

\begin{exe}	\label{exe:geo}
\ex 
\begin{xlist}
\exi{A:} \gll nebara paruo\\
	what.\textsc{obj.foc} do\\ 
	\glt 	'What are [you] doing?'
\exi{B:} \gll an muap=teba / ge o / ge\_mera\\
	\textsc{1sg} eat={\glteba} / nothing \textsc{emph} / nothing\\
	\glt `I'm eating. / Nothing. / Nothing.' \jambox*{[overheard]}
\end{xlist}
\end{exe}

\begin{exe}
	\ex 
	\label{exe:tamanggarabot}
	\begin{xlist}
		\ex \gll ka tamangga=ta bot\\
		\textsc{2sg} where.\textsc{lat}={\glta} go\\
		\glt `Where are you going?' \jambox*{[overheard]}
		
		\ex \gll ka tamangga=ta mia/yecie\\
		\textsc{2sg} where.\textsc{lat}={\glta} come/return\\
		\glt `Where are you coming/returning from?' \jambox*{[overheard]}
	\end{xlist}
\end{exe}

See also §\ref{sec:inittermconv} on the initiation and termination of conversations.

\subsubsection{Polar questions}\is{polar question}\is{yes/no question|see{polar question}}
\label{sec:polq}
Polar questions, questions which are meant to be affirmed or denied, can be divided into polar questions with and without a tag. It is common to answer a polar question with a clause. The interjections \textit{yor} `true' or \textit{ge} `no(t)' may otherwise by used (§\ref{sec:discinterj}).

Polar questions are most commonly followed by a tag \textit{ye ge} `or not'. (\ref{exe:yege}) shows a polar question answered in the negative, in this case with an inherent negative verb \textit{eranun} `cannot' (the positive form is the Malay loan \textit{bisa} `can', see §\ref{sec:lexicalneg} below).

\begin{exe}
	\ex
		\label{exe:yege}
	\begin{xlist}
		\exi{A:}
	{\gll an mat gerket ka bisa nan \textbf{ye} \textbf{ge}\\
		\textsc{1sg} \textsc{3sg.obj} ask \textsc{2sg} can consume or not\\
		\glt `I asked him: ``Can you eat or not?'''
	}
\exi{B:}
	{\gll ma toni o an nan-un eranun\\
	\textsc{3sg} say \textsc{int} \textsc{1sg} consume-\textsc{nmlz}  cannot\\
	\glt `He said: ``Oh, I can't eat.''' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAE-4}{[narr44\_4:46]}}
}
\end{xlist}
\end{exe}

A positive answer to a polar question is given in~(\ref{exe:yege2}).

\begin{exe}	
	\ex \label{exe:yege2}
	\begin{xlist}
	\exi{A:}{
	\gll yuwa me pulisi=a wa \textbf{ye} \textbf{ge}\\
	\textsc{prox} {\glme} police=\textsc{foc} \textsc{prox} or not\\
	\glt	`This one, is this the police or not?'
	}

	\exi{B:}{
	\gll pulisi=a wa\\
	police=\textsc{foc} \textsc{prox}\\
	\glt `This is the police.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA9-9}{[stim6\_11:24]}}
	}
	\end{xlist}
\end{exe}

Alternatively, polar questions can be formed without a tag. The clause structure is the same as in affirmative declarative clauses, but with a different intonation\is{intonation} (see §\ref{sec:inton}). (\ref{exe:ques}) illustrates a polar question and answer.

\begin{exe}
	\ex
	\label{exe:ques}
	\begin{xlist}
		\exi{A:}
		{\glll Ka terara lo?\\
			ka ter=at=a lo\\
			\textsc{2sg} tea=\textsc{obj=foc} want\\
			\glt `Do you want tea?'
		}
		\exi{B:}
		{\glll An terara lo.\\
			an ter=at=a lo\\
			\textsc{1sg} tea=\textsc{obj=foc} want\\
			\glt `I want tea. '} \jambox*{[elic]}
	\end{xlist}
\end{exe}

\subsubsection{Confirmation-seeking and rhetorical questions}
\label{sec:confquest}
Other tags may be used to create rhetorical questions, or questions for which an affirmative answer is expected. (\ref{exe:gebare}) illustrates \textit{ge} `no' which is used as a confirmation-seeking tag.

\begin{exe}
	\ex
	\begin{xlist}
	\exi{A:}
	{\gll wa me pulisi=a hukum \textbf{ge}\\
		\textsc{prox} {\glme} police=\textsc{foc} punish no\\
		\glt `Here the police officer is punishing, right?'
	}
	\exi{B:}
	{\gll m'm pulisi=a hukum\\
		yes police=\textsc{foc} punish\\
		\glt `Yes, the police officer is punishing.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA9-9}{[stim6\_14:29]}}
	}
\end{xlist}
	\label{exe:gebare}
\end{exe}

The tag \textit{e} is also used to evoke agreement from the listener (see §\ref{sec:wcyesno} and §\ref{sec:idphon}). \textit{E} is not only used with questions, and can also be used with declarative clauses for evoking agreement. The question in~(\ref{exe:tage}) remains unanswered.

\begin{exe}
	\ex
	\label{exe:tage}
	{\gll wa me ge \textbf{e}\\
		\textsc{prox} {\glme} no \textsc{tag}\\
		\glt `Not this one, right?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA9-9}{[stim6\_7:24]}}}
\end{exe}

\textit{To} `right' is a confirmation-seeking tag (from Malay \textit{toh}, ultimately from Dutch\il{Dutch} \textit{toch}, all with the same meaning), illustrated in~(\ref{exe:tagto2}). It is also used to check whether the listener is still following along, and is often used to mark information that was previously given by the speaker, as in~(\ref{exe:tagto1}). Usually, an answer is not expected. Like the tag \textit{e}, the use of \textit{to} is not confined to questions, and can be a confirmation-seeking tag on declaratives as well (§\ref{sec:idphon}).

\begin{exe}
	\ex
	{\gll wa me nika wa me pi watko kaniet=et \textbf{to}\\
		\textsc{prox} {\glme} fishing\_line \textsc{prox} {\glme} \textsc{1pl.incl} \textsc{prox.loc} tie={\glet} right\\
		\glt `This is fishing line, this we tie here, don't we?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C75-D}{[stim15\_1:11]}}
	}
	\label{exe:tagto2}
	\ex
	{\gll tumun opa me per nerunggo \textbf{to}\\
		child \textsc{ana} {\glme} water inside right\\
		\glt `So that child is in the water, you see?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B94-F}{[stim21\_0:28]}}
	}
	\label{exe:tagto1}
\end{exe}

\subsubsection{Alternative questions}
\label{sec:altquest}
Alternative questions are questions that pose two alternatives to the addressee, who is expected to choose one. They are made with the help of the conjunction \textit{ye} `or'. In~(\ref{exe:kademorye}), the addressee chooses the second alternative by repeating the proposition.

\begin{exe}
	\ex
	\begin{xlist}
		\exi{A:}
		{\gll ma-tain kademor ye sontum=a mat ajak\\
			\textsc{3sg}-alone mad or person=\textsc{foc} \textsc{3sg.obj} invite\\
			\glt `Did he get mad on his own or did people invite him?'
		}
		\exi{B:}
		{\gll sontum=a mat ajak\\
			person=\textsc{foc} \textsc{3sg.obj} invite\\
			\glt `People invited him.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_29:01]}}
		}
	\end{xlist}
	\label{exe:kademorye}
\end{exe}\is{interrogative clause|)}

\subsection{Imperative clauses}\is{imperative}
\label{sec:impcla}
Imperatives are directive speech acts used for orders and commands\is{command} \parencite[][303]{shopen2007b}. They are marked with enclitic \textit{=te} on the predicate. An imperative clause can but need not contain a subject (addressee). Compare~(\ref{exe:impw}), with subject, and~(\ref{exe:impwo}), without subject.

\begin{exe}
	\ex
	{\gll goras mat sirie ka menyanyi=\textbf{te}\\
		crow \textsc{3sg.obj} order \textsc{2sg} sing=\textsc{imp}\\
		\glt `The crow orders him: ``You sing!''' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB9-6}{[stim1\_0:39]}}
	}
	\label{exe:impw}
	\ex
	{\gll nokidak=\textbf{te}\\
		be\_silent=\textsc{imp}\\
		\glt `Shut up!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_1:20]}}
	}
	\label{exe:impwo}
\end{exe}

Movement verbs ending in \textit{-a} (e.g. \textit{sara} `go up', \textit{marua} `go towards sea', \textit{mia} `come') have imperative forms in \textit{-ei}, as illustrated in~(\ref{exe:erei}) for the imperative form of \textit{era} `go up diagonally'. That example, taken from a route description given to a fictional stranger, also illustrates that imperative forms are not only used for orders, but also for hortatives\is{hortative}.

\begin{exe}
	\ex 
	\gll bo masikit mul-un=ka ka se {\ob}...{\cb} uriap mengga \textbf{erei}\\
	go mosque side-\textsc{3poss=lat} \textsc{2sg} {\glse} {} street \textsc{dist.lat} ascend.\textsc{imp}\\
	\glt `Past the mosque, you go up to that street.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C9A-E}{[narr37\_0:49]}}
	\label{exe:erei}
\end{exe}

Give-constructions, which have no overt verb (§\ref{sec:give}), are put in the imperative by adding the imperative enclitic \textit{=te} to the recipient. Depending on whether the recipient is expressed as a pronoun or as a noun, it may or may not carry benefactive \textit{=ki}. If a benefactive enclitic is present, the imperative is added to that, as in~(\ref{exe:nggite}).

\begin{exe}
	\ex \gll Arifin se emun=at tiri pareir sor me seletkon-i tete=ki ∅=te\\
	Arifin {\glse} mother.\textsc{3poss=obj} run follow fish \textsc{dist} piece-\textsc{objqnt} grandfather.\textsc{mly=ben} give=\textsc{imp}\\
	\glt `Arifin ran after his mother: ``Give a piece of that fish to grandfather!''' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_6:00]}}
	\label{exe:nggite}
\end{exe}

Negated imperatives, i.e. prohibitive constructions, are described in §\ref{sec:negnondecl} below and in §\ref{sec:imp}. For the intonation of imperative clauses, see §\ref{sec:impinton}.\is{non-declarative clause|)}


\section{Negative clauses}\is{negation|(}
\label{sec:clauneg}
This section describes clausal negation. It starts with a description of standard negation with negator \textit{=nin}, the negation of declarative verbal main clauses \parencite{miestamo2005}, in §\ref{sec:standneg}. It then goes on to describe negation in non-declarative and non-verbal clauses, in §\ref{sec:negnondecl} and~§\ref{sec:negnonv}. Dedicated negative verbs with the meanings `not know', `not like' and `cannot' are described in §\ref{sec:lexicalneg}. Two negative polarity items are treated in §\ref{sec:negpol}. This account of Kalamang negative clauses is an adaptation of a part of \textcite{visserneg}.



\subsection{Negation of declarative verbal main clauses}
\label{sec:standneg}
Negation of declarative verbal main clauses (also known as standard negation, \citealt{miestamo2005}) is achieved by adding the negator \textit{=nin} to the predicate. Example~(\ref{exe:sem}) contrasts an affirmative and a negated clause with the stative verb \textit{sem} `to be afraid'. 

\ea \label{exe:sem}
\ea \label{exe:masem}{
	\gll ma sem\\
	\textsc{3sg} be\_afraid\\
	\glt `She's afraid.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA2-F}{[conv11\_2:14]}}}

\ex \label{exe:SV}{
	\gll ma sem=\textbf{nin}\\
	\textsc{3sg} be\_afraid=\textsc{neg}\\
	\glt	`She's not afraid.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_16:33]}}}
\z 
\z 

Examples~(\ref{exe:negv}) and (\ref{exe:negadj}) show a negated transitive verb and a negated intransitive stative verb. In both cases, the constituent order of the non-negated counterpart is the same.

\ea 	\label{exe:negv}	
\gll ma yuon=at konat=\textbf{nin}\\
\textsc{3sg} sun=\textsc{obj} see=\textsc{neg}\\
\glt 	`He didn't see the sun.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_22:55]}}

\z 

\ea \label{exe:negadj}	
\gll wa me mang=\textbf{nin}\\
\textsc{prox} {\glme} bitter=\textsc{neg}\\
\glt	`This one isn't bitter.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB1-3}{[narr34\_3:10]}}
\z 

Give-constructions, with a zero morpheme `give', can be negated by adding \textit{=nin} to the zero morpheme, as in~(\ref{exe:kaanin}).

\ea
\label{exe:kaanin}
\gll 	an me ka an ∅=\textbf{nin} o\\
\textsc{1sg} {\glme} \textsc{2sg} \textsc{1sg} give=\textsc{neg} \textsc{emph}\\
\glt `Me, you didn't give me!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_17:43]}}
\z

When the recipient is a noun instead of a pronoun, the recipient is followed by the benefactive enclitic \textit{=ki}. In a negated clause, the zero morpheme follows the benefactive enclitic, and \textit{=nin} is attached to the zero morpheme, as illustrated in~(\ref{exe:kinin}).

\ea	\label{exe:kinin}
{\gll guru 		tok 		dodon=at 	di=tumun=ki ∅=\textbf{nin}\\
	teacher yet thing=\textsc{obj} \textsc{caus}=child=\textsc{ben} give=\textsc{neg}\\
	\glt 	`The teacher hasn't given the children the gift yet.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_give\_15]}}}

\z 

On a surface level, it looks as if \textit{=nin} is cliticised to pronouns and postpositions in give-constructions, but underlyingly negation of give-constructions works exactly the same as other standard verbal negation: by adding \textit{=nin} to the predicate.                                             

Negative clauses differ from affirmatives in that they do not allow the same verbal marking. The enclitic \textit{=kin} (§\ref{sec:kin}), used in affirmative clauses for the expression of volition and imminent situations, is incompatible with negator \textit{=nin}. This is illustrated in example~(\ref{exe:kinneg}). The range of semantic distinctions that \textit{=kin} can express is lost, or underspecified, once a clause is negated (`neutralisation' in \cite{miestamo2016}).. Thus, (\ref{exe:turnin}) is ambiguous between a reading that the child does not want to eat or is not about to eat, and a plain negation of the verb `to eat', i.e. `the child does not eat'. Note that for explicitly expressing `not want', the Kalamang speaker can use a dedicated expression \textit{suka}-\textsc{poss} \textit{ge}, as described in §\ref{sec:lexicalneg}.

\ea 	\label{exe:kinneg}
\ea [*]	{\label{exe:turkinin}
	\gll tumun muap=\textbf{kin=nin}\\
	child eat={\glkin}\textsc{=neg}\\
	\glt 	`The child doesn't want to eat.' }

\ex 	[]	{\label{exe:turnin}
	\gll tumun muap=\textbf{nin}\\
	child eat=\textsc{neg}\\
	\glt	`The child doesn't (want to) eat.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_1]}}}
\z \z 

Irrealis mood, indicated by enclitic \textit{=et} on the predicate, is compatible with negator \textit{=nin}. It always gets the reading of a negative \is{conditional}conditional, whereas irrealis mood in affirmative clauses is not exclusively used for conditionals (§\ref{sec:et}, \ref{sec:condclause}).

\begin{exe}
	\ex \gll  ki tok newer=\textbf{nin=et} ki sabua nerungga rat-un eranun\\
	\textsc{2sg} yet pay=\textsc{neg}={\glet} \textsc{2pl} tent inside.\textsc{lat} move-\textsc{nmlz} cannot\\
	\glt `If you haven't paid yet, you cannot go inside the tent.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B85-F}{[narr5\_2:18]}}
	\label{exe:newerninet}
\end{exe}	

At least two of the five aspectual markers are compatible with negated verbs, whereby the aspect takes scope over negation. The aspectual word \textit{tok} `yet; still; first' (a.k.a. \is{nondum}nondum, \citealt{vdauwera1998}) can be combined with a negated verb to form the meaning `not yet', and \is{iamitive}iamitive \textit{se} (sometimes translatable as `already') takes the meaning `not any more' in negated clauses. For further illustration of the interaction of these aspectual markers with negation, see §\ref{sec:setok}.\is{nondum}\is{iamitive}

\subsection{Negation in non-declaratives}\is{prohibitive}
\label{sec:negnondecl}
Content questions (§\ref{sec:opq}) are negated with negator \textit{=nin}, as illustrated in~(\ref{exe:kabnin}).

\begin{exe}
\ex \gll naman=a kewe=at kabarat=\textbf{nin}\\
who=\textsc{foc} house=\textsc{obj} sweep=\textsc{neg}\\
\glt `Who doesn't want to sweep the house?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_20]}}
\label{exe:kabnin}
\end{exe}

\is{polar question}Polar questions (§\ref{sec:polq}) are formed by adding \textit{ye ge}, literally `or not', at the end of an affirmative clause, as in~(\ref{exe:yege1}). \textit{Ye ge} is neutral in the sense that there is no expectation of either a positive or negative answer. Leaving \textit{ye ge} out but keeping question intonation\is{intonation} is possible, but not often encountered. 

\ea \label{exe:yege1}
\gll mu kat gonggung ter-nan \textbf{ye} \textbf{ge}\\
\textsc{3pl} \textsc{2sg.obj} call tea-consume or not\\
\glt `Are they calling you for tea?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_20:16]}}
\z 	

To express a negative expectation to an answer (as opposed to confirmation-seeking, §\ref{sec:confquest}), one may use negator \textit{=nin}. Note that there are no occurrences of this type in the natural spoken corpus.

\ea\label{exe:natnin}{
	\gll ma pasa=at nat=\textbf{nin}\\
	\textsc{3sg} rice=\textsc{obj} consume=\textsc{neg}\\
	\glt `Doesn't he eat rice?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_15]}}
}
\z 

There are no occurrences of negated alternative questions.

Prohibitives, the negative counterpart of imperatives (§\ref{sec:impcla}), have a dedicated construction involving pronominal suffix \textit{-mun} and clitic \textit{=in} on the predicate. Note that a combination of \textit{-mun} and negator \textit{=nin} is ungrammatical. Example~(\ref{exe:proh1}) illustrates a prohibitive clause. See also §\ref{sec:proh}.

\ea \label{exe:proh1}
	{\gll ka-\textbf{mun} koyal=\textbf{in}\\
	\textsc{2sg-proh} disturb=\textsc{proh}\\
	\glt 	`Don't you disturb!' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_10:05]}}
}
\z 

\subsection{Negation in non-verbal clauses}\is{negation!non-verbal clause}
\label{sec:negnonv}
Non-verbal clauses may have a NP headed by a noun, a demonstrative\is{demonstrative clause} or a quantifier as predicate, or a noun inflected with the locative or similative postposition (§\ref{sec:nomclau} to~§\ref{sec:sim}). Locative and similative predicates can be negated with negator \textit{=nin}. This also counts for adverbial demonstratives. Predicative NPs may be negated with negator \textit{=nin} or with clause-final propositional negator \textit{ge}. Possessive and demonstrative clauses must be negated with propositional negator \textit{ge}. Quantifiers are not negated. Not-having and negative existential clauses are made with negative existential verb \textit{saerak}.

(\ref{exe:negloc}) is a negated locative, (\ref{exe:mengganin}) a negated lative demonstrative, (\ref{exe:inggonggonin}) a negated animate locative, (\ref{exe:wisenggap}) a negated similative and (\ref{exe:mindinin}) a negated manner adverbial.

\ea \label{exe:negloc}
\gll ma ewun naun-kit=ko=\textbf{nin}\\
\textsc{3sg} root.\textsc{3poss} soil-inside=\textsc{loc}=\textsc{neg}\\
\glt	`Its root is not inside the soil.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB1-3}{[narr34\_4:58]}}
\z 

\ea \label{exe:mengganin}
\gll tete Manggan mu tok mengga=\textbf{nin} weinun\\
grandfather Manggan \textsc{3pl} yet \textsc{dist.lat=neg} too\\
\glt `[I] haven't yet been to grandfather Manggan's either.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_24:27]}}
\z 

\ea \label{exe:inggonggonin}
\gll pengalaman kon yume sampi in=konggo=\textbf{nin}\\
experience one \textsc{dist} until \textsc{1pl.excl=an.loc=neg}\\
\glt `That experience hasn't reached us yet.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C98-7}{[narr26\_15:54]}}
\z

\ea \label{exe:wisenggap}
\gll wise=kap=\textbf{nin}\\
long\_time\_ago=\textsc{sim=neg}\\
\glt `It's not like it used to be.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_32:32]}}
\z 

\ea \label{exe:mindinin}
\gll mindi=\textbf{nin}\\
like\_that=\textsc{neg}\\
\glt `It's not like that.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC3-B}{[conv7\_10:11]}}
\z 

Clauses with a NP as predicate, typically clauses of proper inclusion or ascriptive clauses, may be negated with negator \textit{=nin} or with propositional negator \textit{ge}. Consider~(\ref{exe:guruneg}). What the (pragmatic) difference between the two constructions is remains for further research.

\ea \label{exe:guruneg}
\ea 	\label{exe:guruge}
\gll dun-an me me guru \textbf{ge}\\
cross.sex.sibling-\textsc{1sg.poss} \textsc{dist} {\glme} teacher not\\
\glt `That sibling of mine is not a teacher.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_26]}}

\ex 	\label{exe:gurunin}
\gll ma me guru=\textbf{nin}\\
\textsc{3sg} {\glme} teacher=\textsc{neg}\\
\glt `He is not a teacher.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_26]}}
\z \z

Possessives and demonstratives in predicate function may not be negated with negator \textit{=nin}, but must be negated with propositional negator \textit{ge}. This is illustrated for possessive clauses in~(\ref{exe:posneg}).

\ea \label{exe:posneg}
\ea [*] 		{\label{exe:anggonin}
	\gll don me anggon=\textbf{nin}\\
	thing \textsc{dist} \textsc{1sg.poss=neg}\\
	\glt `That stuff isn't mine.' 	\jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_7]}}}

\ex [] {		\label{exe:anggonge}
	\gll don me anggon \textbf{ge}\\
	thing \textsc{dist} \textsc{1sg.poss} not\\
	\glt `That stuff isn't mine.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_7]}}}
\z \z 

While the proposition `that stuff is mine' is negated in~(\ref{exe:anggonge}), non-possession (i.e. not-having) is expressed with negative existential \textit{saerak}.

\ea 	\label{exe:nohouse}
\gll tumtum-un \textbf{saerak}\\
children-\textsc{1pl.excl.poss} \textsc{neg.exist}\\
\glt `We don't have children.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE2-C}{[narr23\_6:35]}}
\z 

\is{quantifier}Quantifiers are not negated, but some have negative counterparts. The negative counterpart of \textit{tebonggan} `all' is negative existential \textit{saerak}. The negative counterpart of \textit{reidak} `much; many' is \textit{reingge} `not much; not many'. The latter contains the (prenasalised) morpheme \textit{ge} `no(t)'.

%tested but failed 2019: swap order in a this-not-that clause. only tested one way around. (was asked for in Miestamo questionnaire somehwere.)

Finally, existential clauses are negated with negative existential \textit{saerak}.	Examples~(\ref{exe:sorsaerak}) and~(\ref{exe:saerak}) show existential negation. \textit{Saerak} is the opposite of existential \textit{mambon}, as shown in example~(\ref{exe:mamsae}). It is ungrammatical to combine either \textit{mambon} or \textit{saerak} with negator \textit{=nin}. \textit{Saerak} has a slightly more differentiated use than \textit{mambon}. For non-possession, \textit{saerak} is obligatory, whereas for affirmative possessive clauses \textit{mambon} only stresses the presence of a possessed item, but is not necessary to express possession.

\ea
\label{exe:sorsaerak}
\gll sor \textbf{saerak}\\
fish \textsc{neg\_exist}\\
\glt `There is no fish.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_0:55]}}
\z 

\ea \label{exe:saerak}
\gll karena pak \textbf{saerak} karek=ki=a kanie\\
because nail \textsc{neg\_exist} string=\textsc{ins=foc} tie\\
\glt	`Because there were no nails [we] tied with string.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB3-0}{[narr7\_1:03]}}
\z 
\ea \label{exe:mamsae}
\gll utkon kadok-un \textbf{mambon} utkon muin \textbf{saerak}\\
some cloth-\textsc{3poss} \textsc{exist} some \textsc{3pl.poss} \textsc{neg\_exist}\\
\glt	`Some had their cloths, some didn't have theirs.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_19:28]}}
\z 

The aspectual particle \textit{se} `already' and the aspectual word \textit{tok} `yet; still; first' combine with \textit{saerak} as illustrated in~(\ref{exe:sesaerak}) and~(\ref{exe:toksaerak}): they get the meaning `not any more' and `not yet' respectively.

\ea 	\label{exe:sesaerak}
\gll ma nan=et me me tadon \textbf{se} \textbf{saerak}\\
\textsc{3sg} consume={\glet} \textsc{dist} {\glme} cough {\glse} \textsc{neg\_exist}\\
\glt `If they drink that, the cough disappears [is not anymore].' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE4-8}{[narr31\_2:57]}}
\z

\ea	\label{exe:toksaerak}
\gll pitis \textbf{tok} \textbf{saerak}\\
money yet \textsc{neg\_exist}\\
\glt 	`The money isn't there yet.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B98-6}{[narr45\_1:06]}}
\z 

\subsection{Dedicated negative verbs}\is{negation!lexical}
\label{sec:lexicalneg}
Apart from negative existential \textit{saerak} (the opposite of existential \textit{mambon}), described in §\ref{sec:negnonv}, there are three inherently negative verbal constructions: \textit{suka}-\textsc{poss} \textit{ge} `not want; not agree', Verb-\textsc{nmlz} \textit{eranun} `cannot' and \textit{komahal} `not know'. These meanings are all common among dedicated negative verbs cross-linguis\-tic\-ally \parencite{veselinova2013pres}. They are listed with their positive counterparts below. Negation of any of the verbs described in this section, whether positive or negative forms, with negator \textit{=nin} is ungrammatical. However, the construction \textit{suka}-\textsc{poss} \textsc{neg} `to not want; to not agree' may combine with a predicate negated with \textit{=nin}.

\begin{exe}
	\ex 	
	\begin{xlist}
		\ex \textit{lo} `to want; to agree'
		\ex \textit{=kin} volitional
		\ex \textit{suka}-\textsc{poss} V=\textsc{neg}/\textit{ge} `to not want; to not agree'
	\end{xlist}
	\ex 
	\begin{xlist}
		\ex \textit{bisa} `can'
		\ex \textit{eranun} `cannot'
	\end{xlist}
	\ex 
	\begin{xlist}
		\ex \textit{gonggin} `to know'
		\ex \textit{komahal} `to not know'
	\end{xlist}
\end{exe}

\textit{Suka}-\textsc{poss} Verb=\textsc{neg} is a dedicated negative construction that means `to not want; to not agree'. The construction is made with the borrowed verb \textit{suka} (<Indonesian \textit{suka} `to like', but not in use with that meaning in Kalamang), nominalised by inflection with a possessive suffix and followed by a negated verb. It could be paraphrased as `is not of \textsc{pron}'s liking'. The third-person possessive suffix \textit{-un} can be used for all subjects. A speaker can, however, choose to inflect \textit{suka} with another person corresponding with the subject, as in~(\ref{exe:parinin}).

\ea \label{exe:parinin}
\gll sontum haus ba an \textbf{suka-n-an} parin=\textbf{nin}\\
person thirsty but \textsc{1sg} like-{\glnn}-\textsc{1sg.poss} sell=\textsc{neg}\\
\glt `People are thirsty [keen to buy] but I don't want to sell.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDD-5}{[narr42\_34:03]}}
\z

When the predicate is lacking, the construction can be made with \textit{ge} `no', as in~(\ref{exe:leka}). 

\ea 	\label{exe:leka}
\gll canam opa me mat narorar ba lek \textbf{suka-un} \textbf{ge}\\
man {\glopa} {\glme} \textsc{3sg.obj} lead but goat want-\textsc{3poss} no\\
\glt 		`That man is leading it, but the goat doesn't want.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD1-D}{[stim31\_0:51]}}
\z

The inflections of \textit{suka} for all persons are given in Table~\ref{tab:sukaungge}. Note that there is an extra phoneme /n/ between \textit{suka} and the first-person possessive suffix \textit{-an}. It is not clear where this /n/ comes from or what it means. It might be epenthetic, inserted to prevent the first-person possessive marker from disappearing (without the /n/, the first-person form would surface as \textit{sukangge}). See also §\ref{sec:possprob} for a discussion of unexplained phonemes on possessed nouns.

\begin{table}
	\caption{Inflections of \textit{suka}-\textsc{poss} \textit{V}=\textsc{neg}/\textit{ge} `to not want; to not agree'}
	\label{tab:sukaungge}
	
		\begin{tabular}{ll} 
			\lsptoprule
			\textsc{1sg} & \textit{an suka-n-an V=nin/ge}\\
			\textsc{2sg}& \textit{ka suka-ca V=nin/ge}\\
			\textsc{3sg}& \textit{ma suka-un V=nin/ge}\\
			\textsc{1pl.incl}& \textit{pi suka-pe V=nin/ge}\\
			\textsc{1pl.excl}& \textit{in suka-un V=nin/ge}\\
			\textsc{2pl} &\textit{ki suka-ce V=nin/ge}\\
			\textsc{3pl}& \textit{mu suka-un V=nin/ge}\\
			\lspbottomrule 
		\end{tabular}
	
\end{table}

As for the lexical opposite of `not want; not agree', there is a verb \textit{lo} `to consent; to like', but it is rather uncommon. An example is given in~(\ref{exe:lo}). Wishes and desires are more commonly expressed with volitional \textit{=kin}, which is an irrealis marker that is among other functions used as a volitional marker, see example~(\ref{exe:natkin}). As we have seen in section \ref{sec:standneg}, volitional \textit{=kin} and negator \textit{=nin} are incompatible.

\ea \label{exe:lo}
\gll an bo mu et-un=at paning paning=i koyet mu se \textbf{lo}\\
\textsc{1sg} go \textsc{3sg} canoe-\textsc{3poss=obj} ask\_for ask\_for={\gli} finish \textsc{3pl} {\glse} consent\\
\glt `I went to ask for their canoe; after asking, they consented.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C99-E}{[narr8\_0:12]}}
\z

\ea \label{exe:natkin}
\gll bal napikir an sor=at nat=\textbf{kin}\\
dog think \textsc{1sg} fish=\textsc{obj} consume={\glkin}\\
\glt `The dog thinks: ``I want to eat the fish.''' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB9-6}{[stim1\_0:34]}}
\z

In the pair \textit{bisa} `can'/\textit{eranun} `cannot', we are dealing with a Malay form again, but this time for the positive form: \textit{bisa} also means `can' in Malay. But whereas \textit{bisa} in Malay is negated with standard negation marker \textit{tidak} (\textit{tidak bisa} `cannot'), Kalamang has a dedicated form, \textit{eranun}.   The use of \textit{eranun} triggers nominalisation of the negated proposition with \textit{-un} `\textsc{nmlz}'. (This is also the suffix that \textit{eranun} itself seems to carry, but if that is so, it has lost its meaning.) \textit{Eranun} is clause-final. Consider examples~(\ref{exe:bisa}) and~(\ref{exe:ningkabor}). One could also translate the construction Verb-\textsc{nmlz} \textit{eranun} as `V-ing isn't possible', such that example~(\ref{exe:ningkabor}) translates as `toilet-going isn't possible.'

\ea \label{exe:bisa}
\gll ka marua hukat yume \textbf{bisa}\\
\textsc{2sg} move\_seawards fishing\_net \textsc{dist} can\\
\glt `Can you go seawards with the fishing net?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCB-5}{[conv1\_2:56]}}
\z 

\ea 	\label{exe:ningkabor}
\gll ning-kabor pasiengga bot-un \textbf{eranun}\\
sick-stomach to\_beach go-\textsc{nmlz} cannot\\
\glt 	`[If you have] stomachache [you] cannot go to the toilet.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCA-4}{[conv20\_15:58]}}
\z 

\hspace*{-5pt} The pair \textit{gonggin} `to know'/\textit{komahal} `to not know' is illustrated in~(\ref{exe:gonggin}) and~(\ref{exe:komahal}). Note also the colloquial \textit{yeso} `don't know' in~(\ref{exe:gonggin}). This is a clause-initial \is{interjection}interjection uttered at high \is{pitch}pitch which expresses indignation, and not a verb like \textit{komahal}.

\ea \label{exe:gonggin}
\gll yeso ka-tain=a \textbf{gonggin}\\
dunno \textsc{2sg}-alone-\textsc{foc} know\\
\glt `I don't know, only you know.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_18:43]}}
\z 

\ea \label{exe:komahal}
\gll ema tamatko eh ema in=nan \textbf{komahal}\\
mother where \textsc{int} mother \textsc{1pl.excl}=too not\_know\\
\glt `Where's mother?' `Eh, mother, we don't know either.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_5:19]}}
\z 

\is{negation|)}

\subsection{Negative polarity items}
\label{sec:negpol}
Kalamang has two negative polarity\is{negative polarity} items: \textit{-barak} `any' and \textit{don kon∼kon} `any'. \textit{-barak} `any' is a negative polarity item that has to be accompanied by a negated predicate or dedicated negative verb, such as \textit{konatnin} `not see' in~(\ref{exe:lampurbarak}) or \textit{jietnin} `not get' in~(\ref{exe:konbarak}). \textit{Som konbarak} `no-one' in~(\ref{exe:nobody}) could be paraphrased as `not any person'.

\ea \label{exe:lampurbarak}
\gll lampur-\textbf{barak} ma konat=\textbf{nin}\\
lamp-any \textsc{3sg} see=\textsc{neg}\\
\glt `He didn't see any lamp.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_29:22]}}
\z 

\ea \label{exe:konbarak}
\gll sairarar reidak ba kon-\textbf{barak} an jiet=\textbf{nin}\\
lobster many but one-any \textsc{1sg} get=\textsc{neg}\\
\glt `There were many lobsters but I didn't get any.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\z 

\ea \label{exe:nobody}{
	\gll som kon-\textbf{barak} karajang-un \textbf{eranun}\\
	person one-any work-\textsc{nmlz} cannot\\
	\glt `No-one can do the work.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_ind]}}}
\z 

The same suffix is used with the meanings `too' and `even', in which cases it is not a negative polarity item.
%mention that it sometimes means too or even?

\textit{Don kon∼kon} `any' (lit `thing one∼\textsc{red}') can be used with a negative verb or negated predicate (example~\ref{exe:donkonkon}), or can be negated itself (example~\ref{exe:dkonkon2}). (For indefinite pronouns, see §\ref{sec:quantinfl}.)\is{indefinite pronoun!negation}

\ea \label{exe:donkonkon}
\gll ma \textbf{don} \textbf{kon}∼\textbf{kon} paruot=\textbf{nin}\\
\textsc{3sg} thing one∼\textsc{red} do=\textsc{neg}\\
\glt `He didn't do anything.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_25:50]}}
\z 

\ea
\gll kian ma sala-un \textbf{don} \textbf{kon∼kon=nin}\\
	wife.\textsc{1sg.poss} \textsc{3sg} mistake-\textsc{3poss} thing one∼\textsc{red=neg}\\	
	\glt `My wife's mistake doesn't matter.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_16:56]}}
\label{exe:dkonkon2}	
\z 

\section{Variation in clausal structure}\is{variation!syntax}
\label{sec:pragvar}
Two common patterns of variation apply to Kalamang clausal structure. The first is elision of arguments, and the second is reordering of constituents to mark topics and anti-topics. Arguments that are retrievable from the context may be elided, especially in dialogues. This is described in §\ref{sec:syntel}. The marking of topic and anti-topic is described in §\ref{sec:leftd}.

\subsection{Elision of arguments}\is{elision}
\label{sec:syntel}
In natural spoken Kalamang, when retrievable from the context, either the subject or the object may be elided, depending on which stays the same across clauses or utterances. 

An example of subject elision\is{elision!subject} is given in~(\ref{exe:subel}). The first clause contains both subject and object. The object and predicate are repeated in the next clause as an instance of \is{tail-head linkage}tail-head linkage (§\ref{sec:tailhead}). In the second and third clause, the subject is elided.

\begin{exe}
	\ex \glll Mu he erat dimarua. Erat dimarua, kaiat disara.\\
	mu se et=at di=marua et=at di=marua kai=at di=sara\\
	\textsc{3pl} {\glse} canoe=\textsc{obj} \textsc{caus}=move\_seawards canoe=\textsc{obj} \textsc{caus}=move\_seawards firewood=\textsc{obj} \textsc{caus}=ascend\\
	\glt `They moved the canoe towards the water. Moved the canoe towards the water, put the firewood on.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC1-0}{[narr19\_1:32]}}
\label{exe:subel}
\end{exe}

Subject elision is also possible when the subject has been mentioned by another speaker. Consider the following dialogue, where speaker B responds to speaker A eliding the transitive subject (the first response of speaker B) and the intransitive subject (the second response of speaker B).

\begin{exe}
	\ex 
	\begin{xlist}
		\exi{A:}
		{\gll taman-un mat ajak=te kasi minuman\\
			friend-\textsc{3poss} \textsc{3sg.obj} invite={\glte} give drink\\
			\glt `His friends invite him, give him drinks.'
		}
		\label{exe:A1}	
		\exi{B:}
		{\gll mat nacoba\\
			\textsc{3sg.obj} try\\
			\glt `[\textbf{They}] try him'
		}
		\label{exe:B1}	
		\exi{A:}
		{\gll nacoba to ma toni ma se tobat\\
			try \textsc{tag} \textsc{3sg} say \textsc{3sg} {\glse} repent\\	
			\glt `Try, right, he says he repents.'
		}
		\label{exe:A2}	
		\exi{B:}
		{\gll tobat yor se koyet\\
			repent true {\glse} finish\\
			\glt `[\textbf{He}] repents, true, finished.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA9-9}{[stim6\_24:05]}}
		}
		\label{exe:B2}
	\end{xlist}
\end{exe}

In procedural descriptions, the subject may be elided throughout the whole text, being generic or non-referential \parencite[][156]{schapper2014}. Consider the beginning of the monologue of a woman explaining how to weave baskets in~(\ref{exe:zeropro}). There is no subject, and throughout the whole explanation she does not introduce one.

\begin{exe}
	\ex
	\label{exe:zeropro}
	{\gll kiem=at paruot=kin {\ob}...{\cb} gous=at potman\\
		basket=\textsc{obj} make={\glkin} {\ob}...{\cb} bamboo=\textsc{obj} cut\\
		\glt `Making a basket. [\textbf{One}] cuts the bamboo.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB8-C}{[narr11\_0:04]}}
	}
\end{exe}

The first-person inclusive\is{clusivity} pronoun also occurs in a generic sense. In some stretches of discourse, it occurs once every now and then, being left out for a few clauses in a row, as in~(\ref{exe:genericwe}), a description of how traditional houses were built. Here, \textit{pi} `we' does not refer to a specific group of people building a house on a specific occasion, but is a generic pronoun referring to whoever used to build houses the traditional way.

\begin{exe}
	\ex
	\label{exe:genericwe}
	\begin{xlist}
		\exi{}
		{\glll Kewe opa me \textbf{pi} he usar. \\
			kewe opa me pi se usar\\
			house \textsc{ana} {\glme} \textsc{1pl.incl} {\glse} erect\\
			\glt `We erected that house.'
		}
		\exi{}
		{\glll Usari koyet \textbf{pi} he mulai padenat usar,\\
			usar=i koyet pi se mulai paden=at usar \\
			erect={\gli} finish \textsc{1pl.incl} {\glse} begin pole=\textsc{obj} erect\\
			\glt `After erecting we started erecting the poles,'
		}
		\exi{}
		{\glll terus larat napasan,\\
			terus lat=at napasan\\
			then plank.\textsc{mly=obj} put.up\\
			\glt `then [\textbf{we}] put up the planks,'
		}
		\exi{}
		{\glll karena pak saerak karekia kanie.\\
			karena pak saerak karek-ki=a kanie\\
			because nail \textsc{neg\_exist} string-\textsc{ins=foc} tie\\
			\glt `because there were no nails [\textbf{we}] tied with string.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB2-1}{[narr6\_0:29]}}
			
		}
	\end{xlist}
\end{exe}

An example of object elision\is{elision!object} is given in~(\ref{exe:giergit}). The object, \textit{karek} `string', is mentioned in the first clause and elided in the second.

\begin{exe}
	\ex
	{\glll Ma kareat tolma to, ma giergi tolma.\\
		ma karek=at tolma to ma gier=ki tolma\\
		\textsc{3sg} string=\textsc{obj} cut tight \textsc{3sg} teeth=\textsc{ins} cut\\
		\glt `She cut \textbf{string}, right, she cut [\textbf{it}] with her teeth.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_4:27]}}
	} 
	\label{exe:giergit}
\end{exe}

Another example comes from a narrative of which the object is tobacco, but only has tobacco in object position three times in the beginning, given in~(\ref{exe:tabai3}).
\is{object omission}

\begin{exe}
	\ex \glll Su tabaiat kasi kaluar, nene mu. Tabaiat kasi kaluarkan an me kan tabaiat kosomnin.\\
	su \textbf{tabai=at} kasi kaluar nene mu \textbf{tabai=at} kasi kaluarkan an me kan \textbf{tabai=at} kosom=nin\\
	already tobacco=\textsc{obj} make go\_out grandmother \textsc{3pl} tobacco=\textsc{obj} make go\_out \textsc{1sg} {\glme} \textsc{int.mly} tobacco=\textsc{obj} smoke=\textsc{neg}\\
	\glt `Already got out the tobacco, grandmother and those. Got out the tobacco, you know, as for me, I don't smoke, you know.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC5-7}{[narr16\_0:15]}}
	\label{exe:tabai3}
\end{exe}

Throughout the rest of the story there is reference to tobacco, or specific packages of tobacco that are introduced later, but the object is always elided. Object elision is not limited to narratives only; it can occur anywhere in Kalamang \is{discourse}discourse.

\begin{exe}
	\ex \gll jadi mu se kasi keluar=ta me\\
	so \textsc{3pl} {\glse} make go\_out={\glta} {\glme}\\
	\glt `So they already got [\textbf{the tobacco}] out.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC5-7}{[narr16\_0:22]}}
	\ex \gll kodaet koi sarat=et mayilma komet=et\\
	one\_more then ascend={\glet} flip look={\glet}\\
	\glt `Then one more came up, flipped [\textbf{the tobacco}] and looked [at \textbf{it}].' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC5-7}{[narr16\_0:38]}}
	\ex \gll an se kuru bara\\
	\textsc{1sg} {\glse} bring descend\\
	\glt `I brought [\textbf{the tobacco}] down.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC5-7}{[narr16\_2:05]}}	
	\ex \gll mu kasetma	kasetman=i koyet mu se kies\\
	\textsc{3pl} open open={\gli} finish \textsc{3pl} {\glse} wrap\\
	\glt `They opened [\textbf{the tobacco pouch}], after opening [\textbf{it}], they rolled [\textbf{a} \textbf{cigarette}].' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC5-7}{[narr16\_2:05]}}
	\ex \gll mu se potman=i koyet mier yap\\
	\textsc{3pl} {\glse} cut={\gli} finish \textsc{3du} divide\\
	\glt `After cutting [\textbf{the tobacco pouch}] they two divided [\textbf{it}].' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC5-7}{[narr16\_2:56]}}
\end{exe}	


\subsection{Reordering arguments}
\label{sec:leftd}
\is{topic}
The reordering of arguments is used to mark topics and anti-topics. Topics can be marked by fronting or by doubling the subject NP, and anti-topics are at the right edge of the clause. %doubling not really reordering. separate if I have time/energy.

A \is{topic}topic is ``an entity that the speaker identifies, about which information {\ob}...{\cb} is then given'' \parencite[][27]{krifka2012}. A topic may be marked by fronting the topicalised constituent. In addition, it may be marked with topic marker \textit{me} (§\ref{sec:discme}). If the constituent is a subject, or a recipient in a give-construction, it is repeated in the main clause. This kind of fronting is only found with first-person pronoun \textit{an}. A fronted subject is exemplified in~(\ref{exe:anmean}) and a fronted recipient in~(\ref{exe:leftd2}). Fronted objects, in contrast, are not repeated in the main clause. (\ref{exe:kiunatme}) has a fronted object with topic marker \textit{me}, and~(\ref{exe:leftd3}) has a fronted object without topic marker. Topic marker \textit{me} is described in §\ref{sec:discme}. 
\is{focus}Focused constituents can be marked with focus marker \textit{=a} or \textit{=ba}. The constituents remain in their usual position in the clause. Focus is described in §\ref{sec:a}.

\begin{exe}
		\ex
	{\gll an me an kona watko komana=et kanggir-un=ko to\\
		\textsc{1sg} {\glme} \textsc{1sg} think \textsc{prox.loc} skewer={\glet} eye-\textsc{3poss=loc} right\\
		\glt `As for me, I think you skewer it here, in the eye, right?' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C75-D}{[stim15\_4:32]}}
	}
	\label{exe:anmean}
	\ex
	{\gll an me ka an ∅=nin o\\
		\textsc{1sg} {\glme} \textsc{2sg} \textsc{1sg} give=\textsc{neg} \textsc{emph}\\
		\glt `Me, you didn't give me.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_17:44]}}
	}
	\label{exe:leftd2}
	\ex \gll kiun=at me ma gonggung=te ma kirarun=ko\\
	wife.\textsc{3poss=obj} {\glme} \textsc{3sg} call={\glte} \textsc{3sg} side=\textsc{loc}\\
	\glt `His wife he calls, she sits beside him.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_28:27]}}
	\label{exe:kiunatme}
		\ex
	{\gll Mujim=at in tok nawanggar\\
		Mujim=\textsc{obj} \textsc{1pl.excl} still wait\\
		\glt `For Mujim we're still waiting.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD8-4}{[narr1\_1:21]}}
	}
	\label{exe:leftd3}
\end{exe}

Another way to topicalise a subject is by apposition of a more specific NP to a more general NP. This is done by apposing a pronoun and a noun, where the noun specifies the referent of the pronoun. There is no pause between the two NPs. (\ref{exe:doubleA}) shows the apposition of \textit{esa main} `his father' to the subject \textit{ma} `he'. These constructions have been double-checked with two speakers to confirm that these are actual constructions and not repairs. In this example, the subject NPs are pronounced under one intonation contour.

\begin{exe}
	\ex
	\label{exe:doubleA}
	{\gll {\ob}ma{\cb}\textsubscript{NP} {\ob}esa main{\cb}\textsubscript{NP} afukat-sara\\
		\textsc{3sg} father \textsc{3poss} avocado-ascend\\
		\glt `He, his father climbed the avocado tree.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD5-6}{[stim34\_0:16]}}
	}
\end{exe}

In~(\ref{exe:doubleft}) there are even three juxtaposed NPs.

\begin{exe}
	\ex
	\label{exe:doubleft}
	{\gll {\ob}ma{\cb}\textsubscript{NP} {\ob}tatanina opa{\cb}\textsubscript{NP} {\ob}ma{\cb}\textsubscript{NP} buok kuru mia\\
		\textsc{3sg} woman \textsc{ana} \textsc{3sg} betel\_nut bring come\\
		\glt `She, that woman, she came and brought betel nut just now.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_12:23]}}
	}
\end{exe}

Anti-topics (AT) are constituents at the right edge of the clause, which serve to specify or reactivate a topic \parencite{lambrecht1981}. In~(\ref{exe:rd1}), the subject of the clause, \textit{tumun opa me} `that child', is specified in the postposed NP. It is also a reactivation of the topic, which had been mentioned a few turns earlier. Note that the anti-topic is also marked with anaphoric demonstrative \textit{opa} and topic marker \textit{me}. In~(\ref{exe:rd2}), the object \textit{pesawat nunat} `plane sounds', which was left out in the main clause, is mentioned in the postposed NP. Planes but not their sounds had been mentioned before in the narrative, so the specification at the end of the clause serves to indicate what people went to listen to. 

\begin{exe}
	\ex
	{\gll ma minggalot-un=ko {\ob}tumun opa me{\cb}\textsubscript{AT}\\
		\textsc{3sg} bedroom-\textsc{3poss=loc} child \textsc{ana} {\glme}\\
		\glt `He's in his bed, that child.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAF-5}{[stim20\_0:34]}}
	}
	\label{exe:rd1}
		\ex
	{\gll mu utkon se bo keluan to {\ob}pesawat nun=at{\cb}\textsubscript{AT}\\
		\textsc{3pl} some {\glse} go listen right plane sound=\textsc{obj}\\
		\glt `Some went to listen, right, to the plane sounds.' \jambox*{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_12:14]}}
	}
	\label{exe:rd2}
\end{exe}
